### 2015-08-03 jing.Rmd
Add markdown for writeup on what island model is giving us

---
output: pdf_document
---
Hi Jing + Martin,

(Attached is the markdown-processed version of the email - wouldn't it be nice if an email client could do that automagically...)

I figured out what was going on. Short answer: The stuff output in islandR is what we thought it was (i.e. it assumes that $P(j) = 1/j$ for all sources $j$) but isn't what the island model is using. i.e. my screwup.

The genomic bit of the island model gives us
$$
P(i | j, X_{ij}, \theta)
$$
where $X_{ij}$ is the observed STs $i$ on each source $j$ and $\theta$ are the genomic parameters (mutation/recombination/migration).

This is then used to estimate the likelihood of observing a given **human** isolate via the law of total probability to sum over the unknown source $j$
$$
P(h) = \sum_j P(i(h)|j) P(j)
$$
where $P(j)$ is the (unknown) probability that a random human isolate comes from source $j$, and $i(h)$ is the type of human isolate $h$.

Thus, we can estimate the unknown $P(j)$ that we're interested in by putting some prior on it (e.g. a $\mathsf{Dirichlet}(1,1,...1)$) and then using the likelihood over all human cases
$$
\begin{aligned}
P(H) &= \prod_{h \in H} P(i(h))\\
     &= \prod_{h \in H} \sum_j P(i(h)|j) P(j)
\end{aligned}
$$
to estimate $P(j)$ via Metropolis-Hastings.

Thus, the dumb Dirichlet prior which would have a huge influence on $P(j|i)$ is really on the attribution to human cases, which is informed by the distribution of STs we observe on the humans.

In the islandR code, I misread the code and ended up effectively setting $P(j) = 1/j$ and outputting $P(j|i)$ rather than just saving $P(i|j)$.

Jing: Given this, can you spot in the islandR processing code where this happens and fix it? Your graphs should then be showing $P(i|j)$ rather than $P(j|i)$.

I think continuing to look at the Dirchlet/Multinomial we talked about on Wednesday is still valuable to see how little information about $P(j)$, and thus $P(j|i)$ we have before we go and add in the human data. In particular, once you have some simulation results we should be able to add in the human data and run some M-H to estimate the $P(j)$'s.

Cheers,

Jonathan

### 2017-09-26 paper/article_0626.tex
Initial commit of paper (some changes from Martin already incorporated)

\documentclass[times, doublespace]{WileyNJD-v2}%For paper submission
%\documentclass[times]{simauth}
% Packages
%\usepackage{fcyheader}
\usepackage{mdframed}
%\usepackage{subcaption}
\usepackage{subfig}
\usepackage{url}
\usepackage{color}

%\usepackage[T1]{fontenc}
%\usepackage{times}
%\usepackage{bm}
%\usepackage{mathtime}

% Customised Commands
%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{corollary}{Corollary}[theorem]
%\newtheorem{lemma}[theorem]{Lemma}
%\newenvironment{proof}[1][Proof]{\begin{trivlist}
%\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
%\newcommand{\qed}{\nobreak \ifvmode \relax \else
%      \ifdim\lastskip<1.5em \hskip-\lastskip
%      \hskip1.5em plus0em minus0.5em \fi \nobreak
%      \vrule height0.75em width0.5em depth0.25em\fi}

% Bibliographystyle
\bibliographystyle{wileyj}

%\title{Statistical models for the source attribution of zoonotic diseases: A study of campylobacteriosis}

\begin{document}
%\runninghead{Statistical models for the source attribution of zoonotic diseases}
%\author{Sih-Jing Liao\affil{a},
%Jonathan Marshall\affil{a}\corrauth\ \footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
%\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
%\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
%\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}

\begin{abstract}

Preventing and controlling zoonoses with a public health policy depends on the knowledge scientists have about the transmitted pathogens. Modeling jointly the epidemiological data and genetic information on the cases provides a methodology for tracing back the source of infection. The genetic information on cases may be modelled at a variety of levels, as can the epidemiological data. In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the effect of the rurality of each case, is estimated utilising a model that incorporates genetic processes alongside a newly developed genetic-free model. We show that the inference from each model is comparable, and that the effect of rurality may be modelled linearly, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.

\end{abstract}
%\keywords{source attribution; \textit{Campylobacter}; multinomial model; Dirichlet prior; HPD interval; DIC}

%\maketitle

\section{Introduction}
Modeling for disease surveillance data to explore patterns of infectious diseases has had a long history in public health. Infectious diseases can cause high economic and medical costs due to morbidity and mortality. In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}. In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the course of disease transmission. Therefore, quantifying the dynamics of infection is the key to help us better understand zoonoses, which also provides valuable information to professionals who are making decisions on public health policy for disease control.

Over the last few decades there have been many models developed for the analysis of zoonotic diseases. Epidemiological models make use of proportions of cases and geographic data to capture the main characteristics of the spread system. For example, the Susceptible-Infectious-Recovered model \cite{Kerm} classifies populations into a susceptible, infective and recovered status enabling an estimation of the number of individuals who are likely to be infected. Also, models with a combination of spatial and temporal data enable the prediction of the time frame and areas a disease or an outbreak is likely to occur \cite{HeldL, Hoehl, Simo}. However, a framework of a zoonotic transmission passed from animals to humans can have a complexity that involves various exposures contaminated via different pathways. For instance, infected wild birds may contaminate environmental water and cause the spread to water users, either humans or non-humans \cite{Wagen}. This implies that tracing back the source of infection is crucial in enriching the resolution of the network, thereby increasing the ability to implement risk management and intervention \cite{Wilso, Morel}. In terms of statistical analysis, it also means that modeling for zoonoses requires an advanced approach with the focus changed from epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}. Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing the epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrrrr}
      \toprule
      ST & \textit{asp} & \textit{gln} & \textit{glt} & \textit{gly} & \textit{pgm}& \textit{tkt}& \textit{unc}\\ \midrule
      45  & 4 & 7 & 10 & 4 & 1 & 7 & 1\\
      474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The allelic profile of genotypes ST-45 and ST-474 is composed of seven allele numbers at each of the seven loci.}
  \label{tab0}
\end{table}

The genetic information used in the integrated models is typically the genotypes that capture the dynamic changes on DNA sequences occurring in the transmitted pathogens \cite{Cotta}. In order to incorporate the genetic information in the epidemiological data, some molecular tools may need to be deployed in the analysis. For example, a multilocus sequence typing (MLST) \cite{Dingl, Coll} is developed and used particularly to identify the species of the pathogen \textit{Campylobacter}, which causes campylobacteriosis, a worldwide gastrointestinal disease in humans. This method uses the location of genes (called loci) and the variants of genes (called alleles, caused by mutation and recombination) observed from fragmented DNA sequences typed from isolates, which are subjects taken from groups of bacterial colonies produced by a small piece of sample in culture plates. In the process of type identification, MLST determines seven specific housekeeping genes (or loci), representing the main functions of the whole genome sequence, to characterize the species of \textit{Campylobacter} \cite{Urwin}. These housekeeping genes are \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit). Once the seven loci are targeted on sequences, MLST produces an allelic profile consisting of seven positive integers addressing the unique genetic sequence at the respective locus, and it also assigns a unique positive integer as a sequence type (ST) to the associated profile for designation. An illustrative example of a part of MLST data is presented in Table~(\ref{tab0}). It shows that the genotypes ST-45 and ST-474 have different allelic combinations across the seven loci. The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made.

The epidemiological and molecular-based models provide greater insight into source attribution. It improves knowledge about infection attributed to sources from the time scale and the location point of view if spatial-temporal covariates are considered in the model. However, few authors in the literature put an emphasis on validations to these integrated models, in which the genetic effects are difficult to assess analytically due to increased model complexity. We are therefore motivated to develop a simpler model so that we can evaluate the integrated source attribution model and examine its rationality for depicting the actual attribution.

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused by \textit{Campylobacter}, of which \textit{C. jejuni} and \textit{C. coli} are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome might be developed, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In the study, the applied model \cite{Wilso} particularly considers genetic evolution in the pathogen that gives an estimation of the proportion of cases that are more likely to get infected from sources when some genotypes are observed from infected individuals. To uncover the genetic effects contributed behind the model, a simple model is particularly developed to simulate comparative information as the integrated model provides. Based on the information produced by the two models we estimate and compare the attribution probability for each source by fitting a general model, from which more covariates can be incorporated, to explore the variation of the attribution probability from a fresh perspective. Last, we apply the approach to the case study to explore the differences in source attribution between the integrated model and the simple model in relation to the geographical areas where the human cases resided.

\section{Models for source attribution with/without information of evolutionary change}
\subsection{Notation and assumptions}
The ultimate goal of the model development is to estimate the source attribution probability, which can quantify the attribution to each source with probability given a genotype is identified in an individual. The main idea behind the establishment has two parts: the sampling distribution of genotypes associated with sources, and the proportion of human cases for each source. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.

Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection. Let $I$ genotypes be the total number of unique STs detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$. Typically, the number of detected STs $I$ is smaller than the sample size of isolates because some isolates may have the same genetic information.

The likelihood of observing a human case h can be found via the law of total probability, which is expressed as,

\begin{align}
  L\Big(\text{source }j; \text{ST} \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}i(h) \vert \text{source } j\Big) p\Big(\text{source }j\Big),
  \label{Olikh}
\end{align}

where $p(\text{ST }i(h)\vert \text{source }j)$ is the estimated sampling distribution of genotypes for the h$^{th}$ case, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$. The likelihood over all human cases is multiplying each individual likelihood, which sums up the probability of detecting a genotype from the h$^{th}$ case over sources with the associated attribution probability. It is then used in the Metropolis-Hastings algorithm with a priori knowledge of $p(\text{source }j)$ to simulate samples of attribution probability. We can compare the attribution results produced from the integrated and the genetic-free models to explore the genetic efforts the integrated model contributes.

The model with genetic information used in this paper is the integrated model \cite{Wilso} adopted in the source attribution study for human campylobacteriosis \cite{Marsh}. The model has two components: an epidemiological model and an evolutionary model. The epidemiological component is a multinomial-based model capturing the number of genotypes typed from human and non-human isolates. Meanwhile, the evolutionary component estimates the rates of mutation and recombination occurring within sources, and also the migration rate at which the pathogen moves between sources by associating each source with an island. The joint model therefore makes use of the two estimates to find out how likely it is the observed genotypes detected from individuals can be attributed to sources. For simplicity, we will refer to the joint model as ``the island model'' and denote $p(\text{ST }i\vert \text{source }j, \boldsymbol{X}, \boldsymbol{G})$ as the probability the model estimates, where $\boldsymbol{X}$ represents the frequency information provided from the multinomial model and $\boldsymbol{G}$ is the parameter set of evolutionary rates evaluated in the genetic part.

To discover the genetic efforts behind the island model, a simple model is developed to produce similar inforamtion to what the island model provides. With an assumption that the sampling distribution of genotypes on each source is similar to the distribution of observed genotypes, the model is also on a multinomial basis. Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$. To make an inference about $\pi_{ij}$, the likelihood is of a multinomial form,

\begin{align}
        L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) & = \frac{n_j!}{\prod_{i=1}^{I} x_{ij}!} \prod_{i=1}^I \pi_{ij}^{x_{ij}}, \notag
\end{align}

where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$. As the family of Dirichlet distribution is a conjugate pair for the multinomial distribution, assume the prior of $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,

\begin{align}
  p(\boldsymbol{\pi}_{j}) & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij}-1}. \notag
\end{align}

Then, the posterior of $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,

\begin{align}
        p(\boldsymbol{\pi}_{j} \vert \boldsymbol{x}_{j} ) & \propto L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) p(\boldsymbol{\pi}_{j}) \notag \\ 
       & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij} + x_{ij}-1}, \quad \alpha_{ij} >0. \notag
 \end{align}

To express the belief that every human case has the same chance of getting infected from each of $J$ sources, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$ in the application. Therefore, p(ST $i\vert$source $j$, $\boldsymbol{X}$) can be obtained by simulating pseudo counts for each genotype with the posterior probability $\boldsymbol{\pi}_{j}$.

Last, the overall likelihood can be computed by substituting the respective sampling distribution from the island model and the simple model for $p(\text{ST }i(h)\vert\text{source }j)$ in equation~(\ref{Olikh}).

\subsection{Campylobacteriosis data}
The source attribution study used here has MLST data collected in the Manawatu region of New Zealand from 2005 to 2014. The sampling has been done in the same time and from the same geographical location that gives 3,480 isolates taken from human and non-human cases, 1,400 of which are from humans. The non-human isolates are sampled from cattle, sheep, wild birds, environmental water and chicken carcasses, and we therefore put them into four groups of major source of infection: poultry, ruminant, water and others. The total number of unique ST typed from all isolates is 355, and 35\% of genotypes are found among human cases. Table~(\ref{tab1}) lists five common genotypes typed in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-61 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or other sources.

The data also contain location variables such as a binary variable specifying whether a typed human case is from urban or rural areas, and a categorical variable further splitting up the areas into seven levels: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Figure~(\ref{fig1}) presents the number of typed human cases living either in urban or rural areas between 2005 and 2014. It shows that the number of typed cases in urban areas was more than that in rural areas, particularly from 2005 to 2007, then it dropped in 2008 and rose up steadily in the following years. The results are from an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. Table~(\ref{tab2}) further lists the number of typed human cases in each classification of rurality. The degree of rurality is a set of integers with a range of [-3,3] corresponding to each defined geographical level. Overall, the table tells us that more than 70\% of typed cases lived in urban areas, and approximately 7.5\% of individuals have no information about the location. We will neglect the missing data in the following analysis by assuming that these cases are randomly missing. Also, as 2006 Census Meshblock Dataset pointed out \cite{Stat}, the total number of dwellers living in urban and rural areas in the Manawatu region was 130,962 and 20,337, respectively. Suppose the population size in the region during 2005 and 2014 is constant, the case rates of the two areas in 2006 are similar with about 0.12\%, this then implies that more cases found in urban areas is simply because more residents are living there. After the intervention, the rate in rural areas still stays at the same level, but the rate in urban areas remarkably decreases.

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrr}
      \toprule
      ST & Human & Poultry & Ruminants & Water & Others\\ \midrule
      42 & 54 & 7 & 52 & 10 & 2\\
      45  & 141 & 154 & 10 & 21 & 54\\
      61 & 64 & 7 & 54 & 2 & 1\\
      474 & 239 & 60 & 15 & 5 & 7\\
      2381 & 0 & 0 & 0 & 21 & 3\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The frequency of five genotypes found from human and four source isolates.}
  \label{tab1}
\end{table}

\begin{figure}
\centering
%\includegraphics[width=.5\linewidth]{Figures/Databar}
\caption{The number of typed human cases coming from urban/rural areas in the Manawatu region of New Zealand during 2005 and 2014. A poultry intervention conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following two years in urban areas.}
\label{fig1}
\end{figure}

\begin{table}
  \begin{center}
    \begin{tabular}{rlrr}
      \toprule
      Rurality scale & Description & Typed human cases & Population in 2006\\ \midrule
      -3 & Highly rural/remote area & 19 & 1,572\\
      -2 & Rural area with low urban influence & 92 & 8,373\\
      -1 & Rural area with moderate urban influence & 127 & 10,392\\
      0 & Rural area with high urban influence & 70 & 6,579\\
      1 & Independent urban area & 230 & 28,611\\
      2 & Satellite urban area & 178 & 19,725\\
      3 & Main urban area & 579 & 76,047 \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The number of human cases typed during 2005 and 2014 in the Manawatu region of New Zealand whose location fell in each grade of rurality corresponding to the defined area against the respective total number of residents in 2006. In general, the scales that ranged between 0 and 3 are regarded as the urban areas and the remaining scales belong to the rural areas.}
  \label{tab2}
\end{table}

\section{Model fitting for source attribution on rurality scale}
The island model and the simple model in the preceding section describe the distribution of STs by integrating with or without genetic variations in DNA sequences respectively. To further estimate the attribution probability a source contributes the infection to human beings, let $F_{jh}$ denote the attribution probability of source $j$ for the h$^{th}$ human case, with constraints $\sum_{j=1}^J F_j =1$ and $0 \leq F_j \leq 1$, where $j=1, \ldots, J$. We model the probabilities $F_{jh}$ using a linear model on the logit scale, that is,

\begin{align}
  F_{jh}  = \frac{\exp (f_{jh})}{\sum_{j=1}^4 \exp(f_{jh})},
  \label{capF}
\end{align}

where $f_{4h}=0$ is treated as the baseline of $f_{jh}$. Consider the case where the data comprise of $H$ human cases, and with the associated assigned STs and $p$ variables, we develop a general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$. The model for the h$^{th}$ individual is then of the form,

\begin{align}
f_{jh}  = \alpha_{j} + \beta_{j1} c_{1h} + \beta_{j2} c_{2h} + \ldots + \beta_{jp} c_{ph},  \notag   
\end{align}

so that more covariates can be considered in the model to further investigate how the attribution is associated with the factors of interest.

To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the categorical variable representing the classified location of each human case. Then two approaches of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical. To differentiate the performance between the two fitted models, we link ``the linear model'' and ``the categorical model'' to the first and the latter fitted model, respectively. Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as,

\begin{align}
  f_{jh} = \alpha_{j} + \beta_{j} z_{h},
  \label{modL}
\end{align}

where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality. Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~(\ref{modL}) can be rewritten as,

\begin{align}
  f_{jh} & = \beta_{1j} z_{1h}+ \beta_{2j} z_{2h} + \ldots + \beta_{dj} z_{dh} + \ldots +  \beta_{7j} z_{7h}, \\
  \text{where} & \notag\\
z_{dh} & =
\left\{\begin{array}{rcl}
1 && \mbox{if case $h$ is in the category $d$,} \notag \\ 0 && \mbox{otherwise}. 
\end{array}\right.
  \label{modC}
\end{align}

As a consequence, the estimated attribution probabilities are obtained via equation~(\ref{capF}) after fitting the data to model~(\ref{modL}) or to model (4).

\subsection{MCMC algorithm}
In the interest of quantifying the uncertainty of the posterior attribution probability, the Markov chain Monte Carlo (MCMC) method is applied. Assume the priors on parameters of interest in model~(\ref{modL}) and model (4) follow a standard normal distribution. Denote $\boldsymbol{\theta}$ as a parameter set of $\theta_{(t)}$, where t indicates the t$^{th}$ order of parameters in the model, $t=1, \ldots, T$, namely, either $(\alpha_1, \alpha_2, \alpha_3, \beta_1, \beta_2, \beta_3)$ in model~(\ref{modL}) or $(\beta_{11}, \beta_{12}, \ldots, \beta_{dj}, \ldots, \beta_{73})$ in model (4). We use the Metropolis-Hastings algorithm to update $\theta_{(t)}$ and hence $f_{jh}$, and $ F_{jh}$ in a Markov chain with a length of 30,000 iterations. With the aim of the chain convergence, the first 5,000 samples are removed and the sequence is thinned every 10$^{th}$ sample. Here are the steps in detail:

\begin{enumerate}

\item Sample $T$ random values from $N(0,1)$ as starting points of the parameter set $\boldsymbol{\theta}$ for model~(\ref{modL}) or model (4)

\item Propose a candidate $\boldsymbol{\theta}^*$ with one $\theta_{(t)}$ updated by a normal proposal distribution, $N(\theta_{(t)}, 1)$

\item Use $\boldsymbol{\theta}^*$ to calculate a new set of $f^*$ for source $j$, $j=1, 2, 3$, via model~(\ref{modL}) or model (4), and find the associated $F^*$ for each case by putting the vector $(f^*, f_4=0)$ in equation~(\ref{capF})

\item Compute the acceptance rate $\alpha$, a product of the likelihood ratio, the proposal ratio and the prior ratio, given by,

\begin{align*}
\frac{L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)}{L\Big(F; \text{ST}, \boldsymbol{Z}\Big)}\frac{Q\Big(\theta_{(t)} \vert \theta_{(t)}^*\Big)}{Q\Big(\theta_{(t)}^* \vert \theta_{(t)}\Big)}\frac{p\Big(\theta_{(t)}^*\Big)}{p\Big(\theta_{(t)}\Big)}, 
\end{align*}

where the likelihood $L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)$ involves the rurality covariate $\boldsymbol{Z}$ resulting from the attribution probability, $p(\text{source }j\vert \boldsymbol{Z}$) with an assumption that the genotype distribution on sources is independent of the covariate. Then, a random value of $u$ is generated from $U(0,1)$. If $\alpha$ meets $u < \min \left\{1, \alpha \right\}$, in the next state the current proposals of $f^{(*)}$, $F^{(*)}$ and $L\Big(F^{(*)}; \text{ST}, \boldsymbol{Z}\Big)$ are updated; otherwise, the previous proposals are retained. 
\end{enumerate}

%KEEP IT IN .TEX AND ADD IT AS COMMENTS IN YOUR CODE. The above function can be equivalently written as a product of the likelihood ratio and the prior ratio due to an assumption of the symmetric proposal function $Q$ resulting in $\frac{Q(\theta_{(t)} \vert \theta_{(t)}^*)}{Q(\theta_{(t)}^* \vert \theta_{(t)})}=1$. For the prior ratio in the later part of above function, we can obtain by calculating,

%\begin{align*}
%\frac{p(\theta_{(t)}^*)}{p(\theta_{(t)})} & =\frac{\frac{1}{\sqrt{2\pi\sigma^2}}\exp (-\frac{1}{2}\frac{\theta_{(t)}^*^2}{\sigma^2})}{\frac{1}{\sqrt{2\pi\sigma^2}}\exp (-\frac{1}{2}\frac{\theta_{(t)}^2}{\sigma^2})}\\
%& = \exp\Big(\frac{\theta_{(t)}^2-\theta_{(t)}^*^2}{2\sigma^2}\Big).
%\end{align*}

\section{Results}

\begin{figure}
\centering
%\includegraphics[width=.9\linewidth]{Figures/FCILandC(5)}
\caption{The posterior mean of samples $F$ with 80\% HPD interval for source: poultry (red), ruminants (green), water (blue) and others (pinks) over the rurality degrees. The samples of attribution probability are generated from the two respective fitted models: the linear model (upper panel) and the categorical model (lower panel), given the sampling distribution of genotypes with evolutionary information (the island model) or without any genetic information (the simple model).}
\label{fig2}
\end{figure}

\begin{figure}
\centering
%\includegraphics[width=.9\linewidth]{Figures/FCILandC(8)}
\caption{The posterior mean of samples $F$ with 80\% HPD interval for four sources over seven rurality categories after applying our methods to data with the post-intervention time period from 2008 to 2014. The samples of attribution probability are generated from the two respective fitted models -- the linear model (upper panel) and the categorical model (lower panel) -- given the genotype distribution based upon the island model or the simple model.}
\label{fig3}
\end{figure}

To investigate the association between source attribution and human infection, we apply the methods demonstrated in the preceding sections to the campylobacteriosis data and also rearrange the data frame from the ST-based structure to the case-based one for model fitting. The posterior mean of samples $F$ with the 80\% highest posterior density (HPD) interval for each source is illustrated on each rurality grade in Figure~(\ref{fig2}). Given different genotype distributions, the upper panel of the figure is based on the linear model, whereas the lower panel is generated from the categorical model.

%80% credible interval can be 10th through 90th percentile (central credible interval) or highest posterior intervals where 80% is picked that minimises length
%
For the linear model, the simple model suggests a slightly higher attribution probability for poultry and a little decreased probability for ruminants in main urban areas, compared to what the island model estimates. Also, the 80\% HPD interval of $F$ for poultry and for ruminants based on the simple model seems marginally wider than that for the island model. So, there is no significant difference in the proportion of human cases attributed to each source between the island and the simple models. For the categorical model, the comparison between the island and the simple models is quite similar to that for the linear model except for the distinguishable attribution for poultry and ruminants in highly rural areas and the much wider credible intervals of $F$ for four sources. For the two fitted models, the narrower credible intervals noticed in very rural areas from the island model might indicate that the island model gives a little more accurate sampling distribution for poultry and ruminants genotypes especially typed in humans living in rural areas. Further, the relatively larger credible intervals observed in the categorical model might be the effect of allocating the number of cases to each qualitative scale (as a dummy variable), particularly in rural areas due to small observations.

As is known, poultry is the major cause of human campylobacteriosis in urban areas \cite{MullA, Marsh, MullM}. Figure~(\ref{fig2}) shows further evidence that poultry starts becoming the dominant source contributable to the infection from places between minor rural and minor urban areas. In contrast, ruminants are becoming the most attributable source when locations are getting farther away from rural areas with a moderate urban influence. As to water and other sources, they seem to be less influenced by the rurality levels because the probabilities remain less than 0.2 over the seven categories.

Moreover, with a view to examining the robustness of our results, the approach is also applied to data with the time period of post-intervention from 2008 to 2014. The source attribution at different rurality levels is then illustrated in Figure~(\ref{fig3}), displaying that the overall attribution over four sources is moderately consistent with the patterns we previously discovered. However, Figure~(\ref{fig3}) shows a marked decline of attribution probability for poultry and a small rise in probability for ruminants in main urban areas. Despite the fact that the intervention did not eliminate the infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived. This finding also accords with what Sears A. and other researchers investigated for the effect on poultry intervention \cite{AnnS}.

Last, we use deviance information criterion (DIC) to evaluate if the linear model or the categorical model is more suitable to describe the data with different time periods of study. Given the sequence type information from the island model and the simple model, Table~(\ref{tab3}) compares the DIC values of the linear model to that of the categorical model, applied to data with the whole period of study and to data with the post-intervention time period. It tells that the linear model always performs better than the categorical model as this model is penalized more in DIC due to the larger number of parameters. This shows that the assumption of a linear effect on the rurality scales against the source attribution is reasonable.

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrr}
      \toprule
      & \multicolumn{2}{c}{Island model}  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{simple model} \\
      %&  \cmidrule(lr){2-3}\cmidrule(lr){5-6} 
      Data & Linear & Categorical & & Linear & Categorical\\ \midrule
      whole period  & 7913.6 & 7943.3  & & 10328.4  & 10389.8 \\
      post-intervention period  & 5413.4  & 5439.5  & & 6615.1   & 6668.9 \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 (whole period) and to the data from 2008 to 2014 (post-intervention period) given the genotypic information is from the island model or the simple model.}
  \label{tab3}
\end{table}

\section{Discussions}
In recent epidemiological literature, modeling for zoonotic diseases with genomic data has commonly emerged that provides an added value to unravel the transmission of pathogens, but increases difficulties in analytically assessing models. In this paper, with an application of a campylobacteriosis study, we not only compare the source attribution based on the genotype distribution with or without pathogenic evolutionary information, but also unfold the infection attributable to sources at different geographical levels.

In light of the different estimates of sampling distribution of genotypes from the island and the simple models, two approaches of model fitting (linear and non-linear) are utilized to estimate the attribution probability via the Matropolis-Hastings algorithm. The estimate of attribution probability $F$ is not straightforward as it is on a logit scale. To examine how the values of $f$ affect the attribution probabilities, we change the standard deviation of the prior distribution from 1 to 2 in the algorithm. It turns out that the associated posterior mean of $F$ for each source is invariant despite the fact that the proposals of $f$ jump even wider in the random walk. This is because the denominator of the transformation equation~(\ref{capF}) balances the impact made by $f$. Further, a linear model is applied to fit the data, assuming the relationship between the predictor and the response variable is linear. This assumption is also in some way supported by the results produced by the categorical model as it displays a linear-like pattern, illustrating how the level of rurality affects the attribution probability.

This study can contribute considerably to the assessment of the final inference made between the models considered with and without the genetic information. The source attribution estimated by the linear and the non-linear models agrees fairly given two different genotype distributions, indicating that the simple model can at least do similar work to the island model. Also, the extension of the developed model fitting for this application is attainable. There are two possible directions: the genetic data, which will become more comprehensive due to rapidly evolving molecular tools, and the role of water source.

Concerns about the water source have been recently raised \cite{Wagen}. Water plays an important role in the transmission network that can disperse pathogens to uninfected users such as humans, wild birds and livestock over wide areas \cite{Cart, Skel, Leve, Case}. But, water is unlike hosts providing a physical environment to pathogens evolving and amplifying; it is simply a medium storing pathogens that come from infected users, and passes them to the next users who are yet to be infected. Hence, excluding water as a source of infection and discovering a better way to consider it in the extended model will be one of the next steps. Furthermore, a novel MLST approach, whole-genome MLST (wgMLST), has recently appeared in the literature \cite{Bigg, Kova, Llar}. The wgMLST method considers more than 1,000 loci enabling the investigation of the association between isolates with identical allelic profiles and the common source of infection. The new genotyping approach overcomes the limit of the traditional MLST method that only captures genetic diversity on seven loci \cite{Bigg}. This means that how to select invaluable information from such large data, and how to develop a sophisticated model to fit the well-filtered data, will be two potential challenges in future research.

\section*{Acknowledgements}
% We would like to thank two anonymous reviewers for their helpful comments and suggestions, which have led to an improved version of the paper.

%\appendix
%\section{Appendix A}

\bibliography{BibArt}
\end{document}

### 2017-09-26 paper/article_0626.tex
adds a bunch of writing changes

@@
 In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the course of disease transmission.
-Therefore, quantifying the dynamics of infection is the key to help us better understand zoonoses, which also provides valuable information to professionals who are making decisions on public health policy for disease control.
+For zoonotic diseases, where transmission to humans may be via food, water, direct contact with animals, or through environmental contamination, knowledge of the potential sources and pathways of infection is key in improving understanding of the burden of disease.
+Quantifying the burden of illness to each pathway and source, provides valuable information to health professionals making decisions on public health policy for disease control.
 
 Over the last few decades there have been many models developed for the analysis of zoonotic diseases.
-Epidemiological models make use of proportions of cases and geographic data to capture the main characteristics of the spread system.
+Epidemiological models make use of the incidence of cases and geographic data to capture the main characteristics of disease spread.
 For example, the Susceptible-Infectious-Recovered model \cite{Kerm} classifies populations into a susceptible, infective and recovered status enabling an estimation of the number of individuals who are likely to be infected.
-Also, models with a combination of spatial and temporal data enable the prediction of the time frame and areas a disease or an outbreak is likely to occur \cite{HeldL, Hoehl, Simo}.
-However, a framework of a zoonotic transmission passed from animals to humans can have a complexity that involves various exposures contaminated via different pathways.
+Also, models with a combination of spatial and temporal data enable the prediction of the time frame and areas where a disease outbreak is likely to occur \cite{HeldL, Hoehl, Simo}.
+However zoonotic transmission from animals to humans may be complex, involving many sources and exposures linked by differing pathways.
 For instance, infected wild birds may contaminate environmental water and cause the spread to water users, either humans or non-humans \cite{Wagen}.
-This implies that tracing back the source of infection is crucial in enriching the resolution of the network, thereby increasing the ability to implement risk management and intervention \cite{Wilso, Morel}.
-In terms of statistical analysis, it also means that modeling for zoonoses requires an advanced approach with the focus changed from epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}.
-Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing the epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.
+Tracing the source of infection is thus crucial to increasing the ability to implement risk management and intervention \cite{Wilso, Morel}.
+In terms of statistical analysis, it means that modelling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}.
+Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.
 
@@
 
-The genetic information used in the integrated models is typically the genotypes that capture the dynamic changes on DNA sequences occurring in the transmitted pathogens \cite{Cotta}.
-In order to incorporate the genetic information in the epidemiological data, some molecular tools may need to be deployed in the analysis.
-For example, a multilocus sequence typing (MLST) \cite{Dingl, Coll} is developed and used particularly to identify the species of the pathogen \textit{Campylobacter}, which causes campylobacteriosis, a worldwide gastrointestinal disease in humans.
-This method uses the location of genes (called loci) and the variants of genes (called alleles, caused by mutation and recombination) observed from fragmented DNA sequences typed from isolates, which are subjects taken from groups of bacterial colonies produced by a small piece of sample in culture plates.
-In the process of type identification, MLST determines seven specific housekeeping genes (or loci), representing the main functions of the whole genome sequence, to characterize the species of \textit{Campylobacter} \cite{Urwin}.
-These housekeeping genes are \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
-Once the seven loci are targeted on sequences, MLST produces an allelic profile consisting of seven positive integers addressing the unique genetic sequence at the respective locus, and it also assigns a unique positive integer as a sequence type (ST) to the associated profile for designation.
-An illustrative example of a part of MLST data is presented in Table~(\ref{tab0}).
+The genetic information used in such integrated models is typically at the level of genotyping that groups closely related organisms together to capture broader changes in DNA occurring in the transmitted pathogens \cite{Cotta}.
+Typically molecular typing methods, such as multilocus sequence typing (MLST) \cite{Dingl, Coll} might be used, which utilises nucleotide sequences of internal fragments of a small set of housekeeping genes.
+Such sequences have sufficient variation to distinguish differing pathogen lineages, while being relatively stable within lineages.
+Each unique nucleotide sequence (allele) at each housekeeping gene (loci) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, with is usually also assigned a number.
+
+For the pathogen \textit{Campylobacter}, which causes campylobacteriosis, a worldwide gastrointestinal disease in humans.
+For \textit{Campylobacter}, the MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
+An illustrative example of a part of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab0}).
 It shows that the genotypes ST-45 and ST-474 have different allelic combinations across the seven loci.
@@
 
-The epidemiological and molecular-based models provide greater insight into source attribution.
-It improves knowledge about infection attributed to sources from the time scale and the location point of view if spatial-temporal covariates are considered in the model.
-However, few authors in the literature put an emphasis on validations to these integrated models, in which the genetic effects are difficult to assess analytically due to increased model complexity.
-We are therefore motivated to develop a simpler model so that we can evaluate the integrated source attribution model and examine its rationality for depicting the actual attribution.
+The first step for attribution models that utilise genetic information is building the sampling distribution of genotypes among each putative source.
+This may range from using the proportion of each observed genotype (CITATION: TINA HALD) on each source through to utilising allelic profile information to derive mutation and recombination rates within each source, and migration rates between each source (CITATION: ISLAND MODEL).
+A key question is whether more complex genetic models yield superior attribution results or whether a significantly simpler model may suffice, however few authors in the literature have addressed this point.
+This becomes more important as model complexity extends to include epidemiological covariates.
+We are therefore motivated to develop a simple genetic model in order to assess the additional information that the more complex models provide.
 
@@
 The common symptoms of infection are diarrhea, abdominal pain and fever.
-However, a severe complication named Guillain Barre syndrome might be developed, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}.
+However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}.
 The pathogen can be spread between animals, or from animals and wild birds to humans.
 The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces.
-In the study, the applied model \cite{Wilso} particularly considers genetic evolution in the pathogen that gives an estimation of the proportion of cases that are more likely to get infected from sources when some genotypes are observed from infected individuals.
-To uncover the genetic effects contributed behind the model, a simple model is particularly developed to simulate comparative information as the integrated model provides.
-Based on the information produced by the two models we estimate and compare the attribution probability for each source by fitting a general model, from which more covariates can be incorporated, to explore the variation of the attribution probability from a fresh perspective.
-Last, we apply the approach to the case study to explore the differences in source attribution between the integrated model and the simple model in relation to the geographical areas where the human cases resided.
+In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilises only the prevalence of each type to derive the sampling distribution.
+We then extend both models to incorporate covariates, exploring the effect of human case rurality on attribution results, modelling attribution by rurality using a linear trend or using separate categories and performing model comparison.
 
-\section{Models for source attribution with/without information of evolutionary change}
+\section{Models for source attribution with and without information of evolutionary change}
 \subsection{Notation and assumptions}

### 2017-09-26 paper/article_0626.tex
redo the modelling section a bit

@@
 In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilises only the prevalence of each type to derive the sampling distribution.
-We then extend both models to incorporate covariates, exploring the effect of human case rurality on attribution results, modelling attribution by rurality using a linear trend or using separate categories and performing model comparison.
+We then extend both models to incorporate covariates, exploring the effect of human case rurality on attribution results, modelling attribution by rurality using a linear trend or using separate categories in a Bayesian context, and performing model comparison.
 
@@
 \subsection{Notation and assumptions}
-The ultimate goal of the model development is to estimate the source attribution probability, which can quantify the attribution to each source with probability given a genotype is identified in an individual.
-The main idea behind the establishment has two parts: the sampling distribution of genotypes associated with sources, and the proportion of human cases for each source.
+The ultimate goal of attribution models is to estimate the probability that the observed human cases arise from each putative source.
+Given genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to that observed among human cases.
 Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.
@@
 Let $I$ genotypes be the total number of unique STs detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$.
-Typically, the number of detected STs $I$ is smaller than the sample size of isolates because some isolates may have the same genetic information.
+Typically, the number of detected STs $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.
 
-The likelihood of observing a human case h can be found via the law of total probability, which is expressed as,
-
+The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as
 \begin{align}
-  L\Big(\text{source }j; \text{ST} \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}i(h) \vert \text{source } j\Big) p\Big(\text{source }j\Big),
+  L\Big(\text{ST}_1, \text{ST}_2, \ldots, \text{ST}_H\Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p(\text{source }j),
   \label{Olikh}
 \end{align}
+where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$.
+Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \ref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.
 
-where $p(\text{ST }i(h)\vert \text{source }j)$ is the estimated sampling distribution of genotypes for the h$^{th}$ case, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$.
-The likelihood over all human cases is multiplying each individual likelihood, which sums up the probability of detecting a genotype from the h$^{th}$ case over sources with the associated attribution probability.
-It is then used in the Metropolis-Hastings algorithm with a priori knowledge of $p(\text{source }j)$ to simulate samples of attribution probability.
-We can compare the attribution results produced from the integrated and the genetic-free models to explore the genetic efforts the integrated model contributes.
+The asymmetric Island model \cite{Wilso} adopted in the source attribution study for human campylobacteriosis \cite{Marsh} utilises the allelic profile information for each genotype in an evolutionary model, estimating mutation and recombination probabilities within, and migration probabilities between, each source `island'.
+It thus estimates $p(\text{ST}_h \vert \text{source }j)$ indirectly, by first estimating the evolutionary parameters, and then deriving the sampling distributions.
+This allows the asymmetric Island model to estimate the likelihood of observing a genotype on a source when it has not been previously observed.
 
-The model with genetic information used in this paper is the integrated model \cite{Wilso} adopted in the source attribution study for human campylobacteriosis \cite{Marsh}.
-The model has two components: an epidemiological model and an evolutionary model.
-The epidemiological component is a multinomial-based model capturing the number of genotypes typed from human and non-human isolates.
-Meanwhile, the evolutionary component estimates the rates of mutation and recombination occurring within sources, and also the migration rate at which the pathogen moves between sources by associating each source with an island.
-The joint model therefore makes use of the two estimates to find out how likely it is the observed genotypes detected from individuals can be attributed to sources.
-For simplicity, we will refer to the joint model as ``the island model'' and denote $p(\text{ST }i\vert \text{source }j, \boldsymbol{X}, \boldsymbol{G})$ as the probability the model estimates, where $\boldsymbol{X}$ represents the frequency information provided from the multinomial model and $\boldsymbol{G}$ is the parameter set of evolutionary rates evaluated in the genetic part.
-
-To discover the genetic efforts behind the island model, a simple model is developed to produce similar inforamtion to what the island model provides.
-With an assumption that the sampling distribution of genotypes on each source is similar to the distribution of observed genotypes, the model is also on a multinomial basis.
+To discover the effect of incorporating genetic information at the allelic profile level as used in the island model, a simple model is developed for the genotype sampling distribution.
+With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution.
 Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$.
-To make an inference about $\pi_{ij}$, the likelihood is of a multinomial form,
-
+To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
 \begin{align}
@@
 \end{align}
-
 where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$.
-As the family of Dirichlet distribution is a conjugate pair for the multinomial distribution, assume the prior of $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
-
+As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
 \begin{align}
@@
 
-Then, the posterior of $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
-
+Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
 \begin{align}
@@
 
-To express the belief that every human case has the same chance of getting infected from each of $J$ sources, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$ in the application.
-Therefore, p(ST $i\vert$source $j$, $\boldsymbol{X}$) can be obtained by simulating pseudo counts for each genotype with the posterior probability $\boldsymbol{\pi}_{j}$.
+To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$.
+Therefore, p(ST $i\vert$source $j$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.
 
-Last, the overall likelihood can be computed by substituting the respective sampling distribution from the island model and the simple model for $p(\text{ST }i(h)\vert\text{source }j)$ in equation~(\ref{Olikh}).
+Lastly, the overall likelihood can be computed by substituting the respective sampling distribution from the island model and the simple model for $p(\text{ST }i(h)\vert\text{source }j)$ in equation~(\ref{Olikh}), and integrating over the posteriors.
+
+TODO: JM: I think including an algorithm here might be useful. It could then be extended in section 3.1
 
 \subsection{Campylobacteriosis data}
+
 The source attribution study used here has MLST data collected in the Manawatu region of New Zealand from 2005 to 2014.

### 2017-09-26 paper/article_0626.tex
adds some more improvements to the introduction

@@
 \section{Introduction}
-Modeling for disease surveillance data to explore patterns of infectious diseases has had a long history in public health.
+Modeling of disease surveillance data to explore patterns of infectious diseases has had a long history in public health.
 Infectious diseases can cause high economic and medical costs due to morbidity and mortality.
@@
 In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the course of disease transmission.
-For zoonotic diseases, where transmission to humans may be via food, water, direct contact with animals, or through environmental contamination, knowledge of the potential sources and pathways of infection is key in improving understanding of the burden of disease.
+For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key in improving understanding of the burden of disease.
 Quantifying the burden of illness to each pathway and source, provides valuable information to health professionals making decisions on public health policy for disease control.
@@
 However zoonotic transmission from animals to humans may be complex, involving many sources and exposures linked by differing pathways.
-For instance, infected wild birds may contaminate environmental water and cause the spread to water users, either humans or non-humans \cite{Wagen}.
+For instance, infected wild birds may contaminate environmental water and cause disease spread to water users, either humans or other animals \cite{Wagen}.
 Tracing the source of infection is thus crucial to increasing the ability to implement risk management and intervention \cite{Wilso, Morel}.
-In terms of statistical analysis, it means that modelling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}.
+In terms of statistical analysis, modelling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}.
 Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.
@@
 
-The genetic information used in such integrated models is typically at the level of genotyping that groups closely related organisms together to capture broader changes in DNA occurring in the transmitted pathogens \cite{Cotta}.
+The genetic information used in such integrated models is typically through genotyping that groups closely related organisms together to capture broader changes in DNA occurring in the transmitted pathogens \cite{Cotta}.
 Typically molecular typing methods, such as multilocus sequence typing (MLST) \cite{Dingl, Coll} might be used, which utilises nucleotide sequences of internal fragments of a small set of housekeeping genes.
@@
 
-For the pathogen \textit{Campylobacter}, which causes campylobacteriosis, a worldwide gastrointestinal disease in humans.
-For \textit{Campylobacter}, the MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
-An illustrative example of a part of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab0}).
+For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
+An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab0}).
 It shows that the genotypes ST-45 and ST-474 have different allelic combinations across the seven loci.
-The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made.
+The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made, by comparing the distribution of genotypes from human cases with those from potential reservoirs.
 

### 2018-01-24 paper/finalish/article_1023.tex
adds final paper draft

--- 2017-09-26 paper/article_0626.tex
@@
-\documentclass[times, doublespace]{WileyNJD-v2}%For paper submission
+\documentclass[times, doublespace]{simauth}%For paper submission
 %\documentclass[times]{simauth}
 % Packages
-%\usepackage{fcyheader}
 \usepackage{mdframed}
-%\usepackage{subcaption}
+\usepackage{subcaption}
 \usepackage{subfig}
@@
 \usepackage{color}
-
 %\usepackage[T1]{fontenc}
@@
 % Customised Commands
-%\newtheorem{theorem}{Theorem}[section]
-%\newtheorem{corollary}{Corollary}[theorem]
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newenvironment{proof}[1][Proof]{\begin{trivlist}
-%\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-%\newcommand{\qed}{\nobreak \ifvmode \relax \else
-%      \ifdim\lastskip<1.5em \hskip-\lastskip
-%      \hskip1.5em plus0em minus0.5em \fi \nobreak
-%      \vrule height0.75em width0.5em depth0.25em\fi}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
+\newenvironment{proof}[1][Proof]{\begin{trivlist}
+\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+\newcommand{\qed}{\nobreak \ifvmode \relax \else
+      \ifdim\lastskip<1.5em \hskip-\lastskip
+      \hskip1.5em plus0em minus0.5em \fi \nobreak
+      \vrule height0.75em width0.5em depth0.25em\fi}
 
@@
 
-%\title{Statistical models for the source attribution of zoonotic diseases: A study of campylobacteriosis}
-
 \begin{document}
-%\runninghead{Statistical models for the source attribution of zoonotic diseases}
-%\author{Sih-Jing Liao\affil{a},
-%Jonathan Marshall\affil{a}\corrauth\ \footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
-%\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
-%\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
-%\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}
+\runninghead{Extending statistical models for the source attribution of zoonotic diseases}
+\title{Extending statistical models for the source attribution of zoonotic diseases: A study of campylobacteriosis} 
+\author{Sih-Jing Liao\affil{a},
+Jonathan Marshall\affil{a}\corrauth\ \footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
+\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
+\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
+\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}
 
@@
 
-Preventing and controlling zoonoses with a public health policy depends on the knowledge scientists have about the transmitted pathogens.
-Modeling jointly the epidemiological data and genetic information on the cases provides a methodology for tracing back the source of infection.
-The genetic information on cases may be modelled at a variety of levels, as can the epidemiological data.
-In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the effect of the rurality of each case, is estimated utilising a model that incorporates genetic processes alongside a newly developed genetic-free model.
-We show that the inference from each model is comparable, and that the effect of rurality may be modelled linearly, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.
+Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways.
+Modeling jointly the epidemiological data and genetic information on the microbial isolates derived from cases provides a methodology for tracing back the source of infection.
+The isolate genetic information and the epidemiological data may be modeled at a variety of levels.
+In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the extent to which each case resides in a rural compared to urban environment, is estimated.
+A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic data-free model.
+We show that the inference from each model is comparable, and that the effect of ``rurality'' may be modeled linearly, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.
 
 \end{abstract}
-%\keywords{source attribution; \textit{Campylobacter}; multinomial model; Dirichlet prior; HPD interval; DIC}
 
-%\maketitle
+\keywords{source attribution; \textit{Campylobacter}; genetic model; Dirichlet prior; HPD interval; DIC}
+\maketitle
 
@@
 In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}.
-In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the course of disease transmission.
-For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key in improving understanding of the burden of disease.
-Quantifying the burden of illness to each pathway and source, provides valuable information to health professionals making decisions on public health policy for disease control.
+In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the nature of disease transmission.
+For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease.
+Quantifying the burden of illness attributable to each pathway and source, provides valuable information to health professionals making decisions on public health policy for disease control.
 
@@
 Tracing the source of infection is thus crucial to increasing the ability to implement risk management and intervention \cite{Wilso, Morel}.
-In terms of statistical analysis, modelling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}.
+In terms of statistical analysis, modeling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}.
 Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.
@@
 
-The genetic information used in such integrated models is typically through genotyping that groups closely related organisms together to capture broader changes in DNA occurring in the transmitted pathogens \cite{Cotta}.
-Typically molecular typing methods, such as multilocus sequence typing (MLST) \cite{Dingl, Coll} might be used, which utilises nucleotide sequences of internal fragments of a small set of housekeeping genes.
+The genetic information used in such integrated models is typically derived from molecular genotyping that groups closely related organisms together \cite{Cotta}.
+Typically molecular typing methods, such as multilocus sequence typing (MLST) \cite{Dingl, Coll, Urwin} might be used, which utilizes nucleotide sequences of internal fragments of a small set of housekeeping genes.
 Such sequences have sufficient variation to distinguish differing pathogen lineages, while being relatively stable within lineages.
-Each unique nucleotide sequence (allele) at each housekeeping gene (loci) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, with is usually also assigned a number.
+Each unique nucleotide sequence (allele) at each housekeeping gene (locus) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, with is usually also assigned a number.
 
-For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
+For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the commonly used 7-gene MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
 An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab0}).
@@
 
-The first step for attribution models that utilise genetic information is building the sampling distribution of genotypes among each putative source.
-This may range from using the proportion of each observed genotype (CITATION: TINA HALD) on each source through to utilising allelic profile information to derive mutation and recombination rates within each source, and migration rates between each source (CITATION: ISLAND MODEL).
+The first step for attribution models that utilize genetic information is building the sampling distribution of genotypes among each putative source.
+This may range from using the proportion of each observed genotype \cite{Hald} on each source through to utilizing allelic profile information to derive mutation and recombination rates within each source, and migration rates between each source \cite{Wilso}.
 A key question is whether more complex genetic models yield superior attribution results or whether a significantly simpler model may suffice, however few authors in the literature have addressed this point.
 This becomes more important as model complexity extends to include epidemiological covariates.
-We are therefore motivated to develop a simple genetic model in order to assess the additional information that the more complex models provide.
+We are therefore motivated to develop a simple model in order to assess the additional information that the more complex models provide.
 
 In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration.
-Human campylobacteriosis is caused by \textit{Campylobacter}, of which \textit{C.
+Human campylobacteriosis is caused mainly by \textit{C.
 jejuni} and \textit{C.
-coli} are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}.
+coli} which are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}.
 The common symptoms of infection are diarrhea, abdominal pain and fever.
@@
 The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces.
-In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilises only the prevalence of each type to derive the sampling distribution.
-We then extend both models to incorporate covariates, exploring the effect of human case rurality on attribution results, modelling attribution by rurality using a linear trend or using separate categories in a Bayesian context, and performing model comparison.
+In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilizes only the prevalence of each type to derive the sampling distribution.
+We then extend both models to incorporate covariates such as rurality indicating the area where a human case resides, exploring the effect of human case rurality on attribution results, modeling attribution by rurality using a linear trend or using separate categories in a Bayesian context, and performing model comparison.
 
-\section{Models for source attribution with and without information of evolutionary change}
+\section{Models for source attribution with and without microbial genetic information}
 \subsection{Notation and assumptions}
@@
 The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as
+
 \begin{align}
-  L\Big(\text{ST}_1, \text{ST}_2, \ldots, \text{ST}_H\Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p(\text{source }j),
+  L\Big(\text{ST}_1, \text{ST}_2, \cdots, \text{ST}_H \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p\Big(\text{source }j\Big),
   \label{Olikh}
 \end{align}
+
 where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$.
@@
 
-The asymmetric Island model \cite{Wilso} adopted in the source attribution study for human campylobacteriosis \cite{Marsh} utilises the allelic profile information for each genotype in an evolutionary model, estimating mutation and recombination probabilities within, and migration probabilities between, each source `island'.
+The asymmetric Island model \cite{Wilso} adopted in the source attribution study for human campylobacteriosis \cite{Marsh} utilizes the allelic profile information for each genotype in an evolutionary model, estimating mutation and recombination probabilities within, and migration probabilities between, each source `island'.
 It thus estimates $p(\text{ST}_h \vert \text{source }j)$ indirectly, by first estimating the evolutionary parameters, and then deriving the sampling distributions.
@@
 
-To discover the effect of incorporating genetic information at the allelic profile level as used in the island model, a simple model is developed for the genotype sampling distribution.
+To discover the effect of incorporating genetic information at the allelic profile level as used in the asymmetric Island model, a simple model is developed for the genotype sampling distribution.
 With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution.
@@
 To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
+
 \begin{align}
@@
 \end{align}
+
 where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$.
 As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
+
 \begin{align}
@@
 Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
+
 \begin{align}
@@
 To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$.
-Therefore, p(ST $i\vert$source $j$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.
+Therefore, $p(\text{ST}_h \vert \text{source }j)$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.
 
-Lastly, the overall likelihood can be computed by substituting the respective sampling distribution from the island model and the simple model for $p(\text{ST }i(h)\vert\text{source }j)$ in equation~(\ref{Olikh}), and integrating over the posteriors.
-
-TODO: JM: I think including an algorithm here might be useful. It could then be extended in section 3.1
+Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models.
+For each posterior sample, we estimate $p(\text{source }j)$ and its uncertainty from the attribution model~(\ref{Olikh}).
+This has the effect of integrating over the uncertainty in $p(\text{ST}_h \vert \text{source }j)$ when estimating $p(\text{source }j)$.
 
@@
 
-The source attribution study used here has MLST data collected in the Manawatu region of New Zealand from 2005 to 2014.
-The sampling has been done in the same time and from the same geographical location that gives 3,480 isolates taken from human and non-human cases, 1,400 of which are from humans.
-The non-human isolates are sampled from cattle, sheep, wild birds, environmental water and chicken carcasses, and we therefore put them into four groups of major source of infection: poultry, ruminant, water and others.
-The total number of unique ST typed from all isolates is 355, and 35\% of genotypes are found among human cases.
-Table~(\ref{tab1}) lists five common genotypes typed in human and source isolates.
+The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014.
+The sampling was carried out over the same time period and from the same geographical location, providing 3,480 isolates taken from human and non-human cases, 1,400 of which are from humans.
+The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others.
+The total number of unique STs typed from all isolates is 355, and 35\% of genotypes are found among human cases.
+Table~(\ref{tab1}) lists five common genotypes found in human and source isolates.
 The first four genotypes are frequently observed in human cases.
@@
 Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}.
-Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or other sources.
+Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.
 
-The data also contain location variables such as a binary variable specifying whether a typed human case is from urban or rural areas, and a categorical variable further splitting up the areas into seven levels: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area.
-Figure~(\ref{fig1}) presents the number of typed human cases living either in urban or rural areas between 2005 and 2014.
-It shows that the number of typed cases in urban areas was more than that in rural areas, particularly from 2005 to 2007, then it dropped in 2008 and rose up steadily in the following years.
-The results are from an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
-Table~(\ref{tab2}) further lists the number of typed human cases in each classification of rurality.
+The data also contain two location variables: one is binary, the other is categorical.
+The binary variable specifies whether a typed human case is from urban or rural areas, whereas the categorical variable further splits up the areas into seven levels.
+The levels are highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area.
+Table~(\ref{tab2}) lists the number of typed human cases in each classification of rurality.
 The degree of rurality is a set of integers with a range of [-3,3] corresponding to each defined geographical level.
-Overall, the table tells us that more than 70\% of typed cases lived in urban areas, and approximately 7.5\% of individuals have no information about the location.
+Overall, more than 70\% of typed cases lived in urban areas, and approximately 7.5\% of individuals have no information about the location.
 We will neglect the missing data in the following analysis by assuming that these cases are randomly missing.
-Also, as 2006 Census Meshblock Dataset pointed out \cite{Stat}, the total number of dwellers living in urban and rural areas in the Manawatu region was 130,962 and 20,337, respectively.
-Suppose the population size in the region during 2005 and 2014 is constant, the case rates of the two areas in 2006 are similar with about 0.12\%, this then implies that more cases found in urban areas is simply because more residents are living there.
-After the intervention, the rate in rural areas still stays at the same level, but the rate in urban areas remarkably decreases.
+Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is pretty similar, indicating that the population size in the region over the sampling period may be constant.
+We then calculate the infection rate per 100,000 population and illustrate the prevalence in urban and rural areas from 2006 to 2014 in Figure~(\ref{fig1}).
+The figure excludes the year of 2005 because the estimates population of both areas is unavailable and the data of typed human cases are incomplete for the entire year.
+As illustrated in the figure, the infection rates in 2006 are about 60 and 85 in urban and rural areas, respectively.
+We notice that the burden of infection in both areas drop remarkably in 2008, resulting from an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
+It shows the intervention improves the infection in urban areas, however, it only has a temporary effect in rural areas.
 
@@
   \begin{center}
-    \begin{tabular}{crrrrr}
+    \begin{tabular}{rrrrrr}
       \toprule
@@
 
+\begin{table}
+%  \begin{threeparttable}
+  \centering
+      \begin{tabular}{rlrrrr}
+        \toprule
+        Rurality scale & Description & Human cases & 2006 & 2013 \\ \midrule
+        -3 & Highly rural/remote area & 19 & 1,572 & 1,527\\
+        -2 & Rural area with low urban influence & 92 & 8,373 & 8,304\\
+        -1 & Rural area with moderate urban influence & 127 & 10,392 & 10,734\\
+        0 & Rural area with high urban influence & 70 & 6,579 & 7,155\\
+        1 & Independent urban area & 230 & 28,611 & 28,188\\
+        2 & Satellite urban area & 178 & 19,725 & 20,525\\
+        3 & Main urban area & 579 & 76,047 & 78,108\\
+        \bottomrule
+      \end{tabular}
+      \caption{Each category of rurality is defined along with the respective number of human cases typed during 2005 and 2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
+      \label{tab2}     
+\end{table}
+
 \begin{figure}
   \centering
-%\includegraphics[width=.5\linewidth]{Figures/Databar}
-\caption{The number of typed human cases coming from urban/rural areas in the Manawatu region of New Zealand during 2005 and 2014.
-A poultry intervention conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following two years in urban areas.}
+  \includegraphics[width=.5\linewidth]{Figures/Databar}
+  \caption{The rate per 100,000 population given the typed human cases and the population coming from urban or rural areas in the Manawatu region of New Zealand during 2006 and 2014.
+A poultry intervention conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years, particularly in urban areas.}
   \label{fig1}
@@
 
-\begin{table}
-  \begin{center}
-    \begin{tabular}{rlrr}
-      \toprule
-      Rurality scale & Description & Typed human cases & Population in 2006\\ \midrule
-      -3 & Highly rural/remote area & 19 & 1,572\\
-      -2 & Rural area with low urban influence & 92 & 8,373\\
-      -1 & Rural area with moderate urban influence & 127 & 10,392\\
-      0 & Rural area with high urban influence & 70 & 6,579\\
-      1 & Independent urban area & 230 & 28,611\\
-      2 & Satellite urban area & 178 & 19,725\\
-      3 & Main urban area & 579 & 76,047 \\
-      \bottomrule
-    \end{tabular}
-  \end{center}
-  \caption{The number of human cases typed during 2005 and 2014 in the Manawatu region of New Zealand whose location fell in each grade of rurality corresponding to the defined area against the respective total number of residents in 2006.
-In general, the scales that ranged between 0 and 3 are regarded as the urban areas and the remaining scales belong to the rural areas.}
-  \label{tab2}
-\end{table}
-
 \section{Model fitting for source attribution on rurality scale}
-The island model and the simple model in the preceding section describe the distribution of STs by integrating with or without genetic variations in DNA sequences respectively.
+The asymmetric Island model and the simple model in the preceding section describe the distribution of STs by integrating with or without genetic variations in DNA sequences respectively.
 To further estimate the attribution probability a source contributes the infection to human beings, let $F_{jh}$ denote the attribution probability of source $j$ for the h$^{th}$ human case, with constraints $\sum_{j=1}^J F_j =1$ and $0 \leq F_j \leq 1$, where $j=1, \ldots, J$.
@@
 
-%KEEP IT IN .TEX AND ADD IT AS COMMENTS IN YOUR CODE.
-The above function can be equivalently written as a product of the likelihood ratio and the prior ratio due to an assumption of the symmetric proposal function $Q$ resulting in $\frac{Q(\theta_{(t)} \vert \theta_{(t)}^*)}{Q(\theta_{(t)}^* \vert \theta_{(t)})}=1$.
-For the prior ratio in the later part of above function, we can obtain by calculating,
-
-%\begin{align*}
-%\frac{p(\theta_{(t)}^*)}{p(\theta_{(t)})} & =\frac{\frac{1}{\sqrt{2\pi\sigma^2}}\exp (-\frac{1}{2}\frac{\theta_{(t)}^*^2}{\sigma^2})}{\frac{1}{\sqrt{2\pi\sigma^2}}\exp (-\frac{1}{2}\frac{\theta_{(t)}^2}{\sigma^2})}\\
-%& = \exp\Big(\frac{\theta_{(t)}^2-\theta_{(t)}^*^2}{2\sigma^2}\Big).
-%\end{align*}
-
 \section{Results}
@@
 \centering
-%\includegraphics[width=.9\linewidth]{Figures/FCILandC(5)}
-\caption{The posterior mean of samples $F$ with 80\% HPD interval for source: poultry (red), ruminants (green), water (blue) and others (pinks) over the rurality degrees.
-The samples of attribution probability are generated from the two respective fitted models: the linear model (upper panel) and the categorical model (lower panel), given the sampling distribution of genotypes with evolutionary information (the island model) or without any genetic information (the simple model).}
+\includegraphics[width=.9\linewidth]{Figures/FCILandC(5)}
+\caption{The posterior mean of samples $F$ with 80\% HPD interval for source: poultry (red), ruminants (green), water (blue) and others (pink) over the rurality scales from highly rural area to main urban area.
+The samples of attribution probability are generated from the two respective fitted models: the linear model (upper panel) and the categorical model (lower panel), given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the simple model).}
 \label{fig2}
@@
 \centering
-%\includegraphics[width=.9\linewidth]{Figures/FCILandC(8)}
+\includegraphics[width=.9\linewidth]{Figures/FCILandC(8)}
 \caption{The posterior mean of samples $F$ with 80\% HPD interval for four sources over seven rurality categories after applying our methods to data with the post-intervention time period from 2008 to 2014.
-The samples of attribution probability are generated from the two respective fitted models -- the linear model (upper panel) and the categorical model (lower panel) -- given the genotype distribution based upon the island model or the simple model.}
+The samples of attribution probability are generated from the two respective fitted models -- the linear model (upper panel) and the categorical model (lower panel) -- given the genotype distribution based upon the asymmetric Island model or the simple model.}
 \label{fig3}
@@
 
-%80% credible interval can be 10th through 90th percentile (central credible interval) or highest posterior intervals where 80% is picked that minimises length
-%
-For the linear model, the simple model suggests a slightly higher attribution probability for poultry and a little decreased probability for ruminants in main urban areas, compared to what the island model estimates.
-Also, the 80\% HPD interval of $F$ for poultry and for ruminants based on the simple model seems marginally wider than that for the island model.
-So, there is no significant difference in the proportion of human cases attributed to each source between the island and the simple models.
-For the categorical model, the comparison between the island and the simple models is quite similar to that for the linear model except for the distinguishable attribution for poultry and ruminants in highly rural areas and the much wider credible intervals of $F$ for four sources.
-For the two fitted models, the narrower credible intervals noticed in very rural areas from the island model might indicate that the island model gives a little more accurate sampling distribution for poultry and ruminants genotypes especially typed in humans living in rural areas.
+For the linear model, the simple model estimates 75\% and 32\% of cases dwelling in main urban areas have acquired the infection from poultry and ruminants, respectively, whereas the asymmetric Island model provides less than 70\% and approximately 30\% of cases attributable to the respective source.
+Further, the 80\% HPD interval of $F$ for poultry and for ruminants based on the simple model seems marginally wider than that for the asymmetric Island model.
+So, there is no significant difference in the proportion of human cases attributed to each source between the asymmetric Island and the simple models.
+For the categorical model, the comparison between the asymmetric Island and the simple models is quite similar to that for the linear model except for the distinguishable attribution for poultry and ruminants in highly rural areas and the much wider credible intervals of $F$ for four sources.
+For the two fitted models, the narrower credible intervals noticed in very rural areas from the asymmetric Island model might indicate that the asymmetric Island model gives a little more accurate sampling distribution for poultry and ruminants genotypes especially typed in humans living in rural areas.
 Further, the relatively larger credible intervals observed in the categorical model might be the effect of allocating the number of cases to each qualitative scale (as a dummy variable), particularly in rural areas due to small observations.
 
-As is known, poultry is the major cause of human campylobacteriosis in urban areas \cite{MullA, Marsh, MullM}.
-Figure~(\ref{fig2}) shows further evidence that poultry starts becoming the dominant source contributable to the infection from places between minor rural and minor urban areas.
+Figure~(\ref{fig2}) shows that poultry starts becoming the dominant source contributable to the infection between rural areas with moderate and with high urban influence.
 In contrast, ruminants are becoming the most attributable source when locations are getting farther away from rural areas with a moderate urban influence.
@@
 Despite the fact that the intervention did not eliminate the infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.
-This finding also accords with what Sears A.
-and other researchers investigated for the effect on poultry intervention \cite{AnnS}.
 
-Last, we use deviance information criterion (DIC) to evaluate if the linear model or the categorical model is more suitable to describe the data with different time periods of study.
-Given the sequence type information from the island model and the simple model, Table~(\ref{tab3}) compares the DIC values of the linear model to that of the categorical model, applied to data with the whole period of study and to data with the post-intervention time period.
+Lastly, we use deviance information criterion (DIC) to evaluate if the linear model or the categorical model is more suitable to describe the data with different time periods of study.
+Given the sequence type information from the asymmetric Island model and the simple model, Table~(\ref{tab3}) compares the DIC values of the linear model to that of the categorical model, applied to data with the whole period of study and to data with the post-intervention time period.
 It tells that the linear model always performs better than the categorical model as this model is penalized more in DIC due to the larger number of parameters.
@@
       \toprule
-      & \multicolumn{2}{c}{Island model}  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{simple model} \\
-      %&  \cmidrule(lr){2-3}\cmidrule(lr){5-6} 
+      & \multicolumn{2}{c}{Island model}  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Simple model} \\
       Data & Linear & Categorical & & Linear & Categorical\\ \midrule
@@
   \end{center}
-  \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 (whole period) and to the data from 2008 to 2014 (post-intervention period) given the genotypic information is from the island model or the simple model.}
+  \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 (whole period) and to the data from 2008 to 2014 (post-intervention period) given the genotypic information is from the asymmetric Island model or the simple model.}
   \label{tab3}
@@
 \section{Discussions}
-In recent epidemiological literature, modeling for zoonotic diseases with genomic data has commonly emerged that provides an added value to unravel the transmission of pathogens, but increases difficulties in analytically assessing models.
-In this paper, with an application of a campylobacteriosis study, we not only compare the source attribution based on the genotype distribution with or without pathogenic evolutionary information, but also unfold the infection attributable to sources at different geographical levels.
+Models that determine the source of human infection, particularly with zoonotic pathogens that originate in animal populations, are of considerable value to public health policy-makers.
+However, the complexity of model building may increase because of the complex transmission pathways of pathogens, in particular when the model combines the epidemiological and molecular data.
+But, is it necessary to use more complex genetic models for the source attribution of zoonoses?
+There is limited research on whether using a simpler model could work as effective as complex genetic models.
+Here we developed a relatively simple model to estimate the attribution probability for each source of \textit{Campylobacter} infection.
+This model differs to the asymmetric Island model, it does not model the pathogenic evolution but uses a simple genetic classification such as a sequence type of each isolate to illustrate the sampling distribution of genotypes.
+With the aim of filling the research gap, our results showed that the simple model can do at least similar work, compared to the asymmetric Island model.
 
-In light of the different estimates of sampling distribution of genotypes from the island and the simple models, two approaches of model fitting (linear and non-linear) are utilized to estimate the attribution probability via the Matropolis-Hastings algorithm.
-The estimate of attribution probability $F$ is not straightforward as it is on a logit scale.
-To examine how the values of $f$ affect the attribution probabilities, we change the standard deviation of the prior distribution from 1 to 2 in the algorithm.
-It turns out that the associated posterior mean of $F$ for each source is invariant despite the fact that the proposals of $f$ jump even wider in the random walk.
-This is because the denominator of the transformation equation~(\ref{capF}) balances the impact made by $f$.
-Further, a linear model is applied to fit the data, assuming the relationship between the predictor and the response variable is linear.
-This assumption is also in some way supported by the results produced by the categorical model as it displays a linear-like pattern, illustrating how the level of rurality affects the attribution probability.
+In light of the different estimates of the sampling distribution of genotypes from the asymmetric Island and the simple models, two approaches of model fitting (linear and non-linear) were utilized to estimate the attribution probability via the Metropolis-Hastings algorithm.
+Under the assumption of a logit scale on the attribution probability, we examined how the values of $f$ affected the resulting attribution probability.
+We changed the standard deviation of the prior distribution from 1 to 2 in the algorithm.
+The associated posterior mean of attribution probability for each source was invariant despite the fact that the proposals of $f$ jumped even wider in the random walk.
+This is because the denominator of the equation~(\ref{capF}) balances the impact made by the values of $f$.
+Further, we included a rurality covariate in the model fitting.
+Our findings not only coincided with the hypothesis that poultry is the major cause of human campylobacteriosis in urban areas \cite{MullA, Marsh, MullM, Leve} but also revealed the linear geographic effect on the infection.
+It means that poultry is less likely responsible for the infection when cases dwell away from urban areas.
+Furthermore, the result of decreasing poultry contribution in main urban areas from fitting the data with post-intervention period accorded with what Sears A.
+and other researchers investigated the effect of poultry intervention \cite{AnnS}.
+This could provide additional information that would help policy-makers to target interventions to populations residing in urban compared to rural areas.
 
-This study can contribute considerably to the assessment of the final inference made between the models considered with and without the genetic information.
-The source attribution estimated by the linear and the non-linear models agrees fairly given two different genotype distributions, indicating that the simple model can at least do similar work to the island model.
-Also, the extension of the developed model fitting for this application is attainable.
-There are two possible directions: the genetic data, which will become more comprehensive due to rapidly evolving molecular tools, and the role of water source.
-
-Concerns about the water source have been recently raised \cite{Wagen}.
-Water plays an important role in the transmission network that can disperse pathogens to uninfected users such as humans, wild birds and livestock over wide areas \cite{Cart, Skel, Leve, Case}.
-But, water is unlike hosts providing a physical environment to pathogens evolving and amplifying; it is simply a medium storing pathogens that come from infected users, and passes them to the next users who are yet to be infected.
-Hence, excluding water as a source of infection and discovering a better way to consider it in the extended model will be one of the next steps.
-Furthermore, a novel MLST approach, whole-genome MLST (wgMLST), has recently appeared in the literature \cite{Bigg, Kova, Llar}.
-The wgMLST method considers more than 1,000 loci enabling the investigation of the association between isolates with identical allelic profiles and the common source of infection.
-The new genotyping approach overcomes the limit of the traditional MLST method that only captures genetic diversity on seven loci \cite{Bigg}.
-This means that how to select invaluable information from such large data, and how to develop a sophisticated model to fit the well-filtered data, will be two potential challenges in future research.
+In conclusion, the simple model is comparable to the asymmetric Island model.
+The potential of the inclusion of covariates in the model could be applied to help make a decision on the control of the infection.
+In addition, the simple model (although based on the genetic data) does not capture the evolutionary processes that could be also applied to any type of classification such as phenotype in the research of interest.
+Nonetheless, there is some concern to apply our approach.
+One is the role of water that has been recently addressed in the literature \cite{Wagen} by treating it as a medium instead in the transmission network.
+The other is the larger pathogen datasets resulted from rapidly evolving molecular tools such as the extending MLST schemes typing from seven loci to whole-genome MLST.
+Hence, reconsidering the role of water as a source of infection and selecting the invaluable information from such large pathogenic data to fit the extended model will be the potential challenges in future research.
 
 \section*{Acknowledgements}
-% We would like to thank two anonymous reviewers for their helpful comments and suggestions, which have led to an improved version of the paper.
-
-%\appendix
-%\section{Appendix A}
 
 \bibliography{BibArt}
+
 \end{document}

### 2018-02-05 paper/finalish/article_1023.tex
add simple changes to first 3 sections

@@
 \runninghead{Extending statistical models for the source attribution of zoonotic diseases}
-\title{Extending statistical models for the source attribution of zoonotic diseases: A study of campylobacteriosis} 
+\title{Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis} 
 \author{Sih-Jing Liao\affil{a},
@@
 A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic data-free model.
-We show that the inference from each model is comparable, and that the effect of ``rurality'' may be modeled linearly, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.
+We show that the inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.
 
@@
 For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease.
-Quantifying the burden of illness attributable to each pathway and source, provides valuable information to health professionals making decisions on public health policy for disease control.
+Quantifying the burden of illness attributable to each pathway and source provides valuable information to health professionals making decisions on public health policy for disease control.
 
@@
 This may range from using the proportion of each observed genotype \cite{Hald} on each source through to utilizing allelic profile information to derive mutation and recombination rates within each source, and migration rates between each source \cite{Wilso}.
-A key question is whether more complex genetic models yield superior attribution results or whether a significantly simpler model may suffice, however few authors in the literature have addressed this point.
+A key question is whether more complex genetic models yield superior attribution results or whether a significantly simpler model may suffice, but few authors in the literature have addressed this point.
 This becomes more important as model complexity extends to include epidemiological covariates.
@@
 In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilizes only the prevalence of each type to derive the sampling distribution.
-We then extend both models to incorporate covariates such as rurality indicating the area where a human case resides, exploring the effect of human case rurality on attribution results, modeling attribution by rurality using a linear trend or using separate categories in a Bayesian context, and performing model comparison.
+We then extend both models to incorporate covariates such as rurality indicating the area where a human case resides, exploring the effect of human case rurality on attribution results, modeling attribution by rurality using a linear trend on the logit scale or using separate categories in a Bayesian context, and performing model comparison.
 
@@
 The ultimate goal of attribution models is to estimate the probability that the observed human cases arise from each putative source.
-Given genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to that observed among human cases.
+Our data comprise microbial genotype information from each observed human case obtained from analysis of stool samples, and also from a pool of non-human cases corresponding to potential zoonotic sources of disease
+
+Given this genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to the set of genotypes observed among human cases.
 Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.
@@
 Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection.
-Let $I$ genotypes be the total number of unique STs detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$.
+Let $I$ genotypes be the total number of unique sequence types (STs) detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$.
 Typically, the number of detected STs $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.
@@
 The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as
-
 \begin{align}
@@
 \end{align}
-
-where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$.
-Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \ref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.
+where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, and $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$.
+Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \eqref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.
 
@@
 To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
-
 \begin{align}
@@
 \end{align}
-
 where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$.
 As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
-
 \begin{align}
@@
 \end{align}
-
 Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
-
 \begin{align}
@@
  \end{align}
-
 To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$.
@@
 Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models.
-For each posterior sample, we estimate $p(\text{source }j)$ and its uncertainty from the attribution model~(\ref{Olikh}).
+For each posterior sample, we make inferences for $p(\text{source }j)$ using the likelihood~\eqref{Olikh}.
 This has the effect of integrating over the uncertainty in $p(\text{ST}_h \vert \text{source }j)$ when estimating $p(\text{source }j)$.
@@
 The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014.
-The sampling was carried out over the same time period and from the same geographical location, providing 3,480 isolates taken from human and non-human cases, 1,400 of which are from humans.
+The sampling was carried out over the same time period and from the same geographical location, providing $H+N=3,480$ isolates taken from human and non-human cases, $H=1,400$ of which are from humans.
 The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others.
-The total number of unique STs typed from all isolates is 355, and 35\% of genotypes are found among human cases.
+The total number of unique STs typed from all isolates is $I=355$, and 35\% of genotypes are found among human cases.
 Table~(\ref{tab1}) lists five common genotypes found in human and source isolates.
@@
 We will neglect the missing data in the following analysis by assuming that these cases are randomly missing.
-Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is pretty similar, indicating that the population size in the region over the sampling period may be constant.
+Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is quite stable.
 We then calculate the infection rate per 100,000 population and illustrate the prevalence in urban and rural areas from 2006 to 2014 in Figure~(\ref{fig1}).
-The figure excludes the year of 2005 because the estimates population of both areas is unavailable and the data of typed human cases are incomplete for the entire year.
-As illustrated in the figure, the infection rates in 2006 are about 60 and 85 in urban and rural areas, respectively.
+The Figure excludes the year of 2005 because the estimates population of both areas is unavailable and the data of typed human cases are incomplete for the entire year.
+As illustrated in the Figure, the infection rates in 2006 are about 60 and 85 in urban and rural areas, respectively.
 We notice that the burden of infection in both areas drop remarkably in 2008, resulting from an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
-It shows the intervention improves the infection in urban areas, however, it only has a temporary effect in rural areas.
+It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.
 

### 2018-02-05 paper/finalish/article_1023.tex
converts line endings
adds some more change to section 3

@@
 \section{Model fitting for source attribution on rurality scale}
-The asymmetric Island model and the simple model in the preceding section describe the distribution of STs by integrating with or without genetic variations in DNA sequences respectively.
-To further estimate the attribution probability a source contributes the infection to human beings, let $F_{jh}$ denote the attribution probability of source $j$ for the h$^{th}$ human case, with constraints $\sum_{j=1}^J F_j =1$ and $0 \leq F_j \leq 1$, where $j=1, \ldots, J$.
+
+\subsection{Incorporating covariates}
+
+Previously we described how to estimate the marginal probabilities that a randomly selected human case is due to a given source.
+To extend this analysis so as to include individual level covariates, we need to calculate subject-specific attribution (conditional) probabilities.
+To that end, let $F_{jh}$ denote the attribution probability of source $j$ for the $h^\textrm{th}$ human case, with constraints $\sum_{j=1}^J F_j =1$ and $0 \leq F_j \leq 1$, where $j=1, \ldots, J$.
 We model the probabilities $F_{jh}$ using a linear model on the logit scale, that is,
-
 \begin{align}
@@
 \end{align}
-
 where $f_{4h}=0$ is treated as the baseline of $f_{jh}$.
-Consider the case where the data comprise of $H$ human cases, and with the associated assigned STs and $p$ variables, we develop a general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$.
-The model for the h$^{th}$ individual is then of the form,
-
+Consider the case where the ST data for each human case is supplemented by $p$ additional variables.
+We develop a general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$.
+The model for the $h^\textrm{th}$ individual is then of the form,
 \begin{align}
@@
 \end{align}
-
 so that more covariates can be considered in the model to further investigate how the attribution is associated with the factors of interest.
+Note that if there is a single categorical variable with $L$ levels, then $F_{jh}$ and $f_{jh}$ will take no more than $L$ distinct values.
+In a slight abuse of notation, we will at times refer to $F_{jh}$ in which the $h$ index refers to the factor level, rather than to a particular subject at that level.
 
 To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the categorical variable representing the classified location of each human case.
-Then two approaches of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical.
+Then two ways of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical.
 To differentiate the performance between the two fitted models, we link ``the linear model'' and ``the categorical model'' to the first and the latter fitted model, respectively.
 Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as,
-
 \begin{align}
@@
 \end{align}
-
 where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality.
 Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~(\ref{modL}) can be rewritten as,
-
-\begin{align}
-  f_{jh} & = \beta_{1j} z_{1h}+ \beta_{2j} z_{2h} + \ldots + \beta_{dj} z_{dh} + \ldots +  \beta_{7j} z_{7h}, \\
-  \text{where} & \notag\\
+\begin{equation}
+  f_{jh} = \beta_{1j} z_{1h}+ \beta_{2j} z_{2h} + \ldots + \beta_{dj} z_{dh} + \ldots +  \beta_{7j} z_{7h},
+  \label{modC}
+\end{equation}
+where
+\begin{equation*}
 z_{dh} & =
@@
 \end{array}\right.
-  \label{modC}
-\end{align}
-
-As a consequence, the estimated attribution probabilities are obtained via equation~(\ref{capF}) after fitting the data to model~(\ref{modL}) or to model (4).
+\end{equation*}
+As a consequence, the estimated attribution probabilities are obtained via equation~(\ref{capF}) after fitting the data to model~\eqref{modL} or to model~\eqref{modC}.
 
 \subsection{MCMC algorithm}
-In the interest of quantifying the uncertainty of the posterior attribution probability, the Markov chain Monte Carlo (MCMC) method is applied.
+In the interest of quantifying the uncertainty of the posterior attribution probability, we perform Bayesian inference for source attribution probabilities using Markov chain Monte Carlo (MCMC) methods.
 Assume the priors on parameters of interest in model~(\ref{modL}) and model (4) follow a standard normal distribution.
-Denote $\boldsymbol{\theta}$ as a parameter set of $\theta_{(t)}$, where t indicates the t$^{th}$ order of parameters in the model, $t=1, \ldots, T$, namely, either $(\alpha_1, \alpha_2, \alpha_3, \beta_1, \beta_2, \beta_3)$ in model~(\ref{modL}) or $(\beta_{11}, \beta_{12}, \ldots, \beta_{dj}, \ldots, \beta_{73})$ in model (4).
-We use the Metropolis-Hastings algorithm to update $\theta_{(t)}$ and hence $f_{jh}$, and $ F_{jh}$ in a Markov chain with a length of 30,000 iterations.
-With the aim of the chain convergence, the first 5,000 samples are removed and the sequence is thinned every 10$^{th}$ sample.
+Let $\boldsymbol{\theta}$ denote the vector of parameters, with elements $\theta_{t}$ for $t=1, \ldots, T$.
+For example, $\boldsymbol{\theta}=\{\alpha_1, \alpha_2, \alpha_3, \beta_1, \beta_2, \beta_3\}$ in model~\eqref{modL} or $\boldsymbol{\theta}=\{\beta_{11}, \beta_{12}, \ldots, \beta_{dj}, \ldots, \beta_{73}\}$ in model~\eqref{modC}.
+We use the Metropolis-Hastings algorithm to update $\theta_{(t)}$ and hence $f_{jh}$ and $ F_{jh}$, in a Markov chain with a length of 30,000 iterations.
+The first 5,000 samples are removed as the burn-in period (during which time the chain converges) and the sequence is thinned every 10$^{th}$ sample to reduce computer storage.
 Here are the steps in detail:
@@
 \begin{enumerate}
-
 \item Sample $T$ random values from $N(0,1)$ as starting points of the parameter set $\boldsymbol{\theta}$ for model~(\ref{modL}) or model (4)

### 2018-02-05 paper/finalish/article_1023.tex
updates the rest of the things for section 3

@@
 We develop a general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$.
-The model for the $h^\textrm{th}$ individual is then of the form,
+The model for the $h^\textrm{th}$ individual is then of the form
 \begin{align}
@@
 To differentiate the performance between the two fitted models, we link ``the linear model'' and ``the categorical model'' to the first and the latter fitted model, respectively.
-Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as,
+Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as
 \begin{align}
@@
 where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality.
-Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~(\ref{modL}) can be rewritten as,
+Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~(\ref{modL}) can be rewritten as
 \begin{equation}
@@
 \begin{equation*}
-z_{dh} & =
+z_{dh} =
 \left\{\begin{array}{rcl}
@@
 \begin{enumerate}
-\item Sample $T$ random values from $N(0,1)$ as starting points of the parameter set $\boldsymbol{\theta}$ for model~(\ref{modL}) or model (4)
+\setcounter{enumi}{-1}
+\item Sample $T$ random values from $N(0,1)$ as the initial values points of the parameter set $\boldsymbol{\theta}$ for model~\eqref{modL} or model~\eqref{modC}.
+\item Sample a permutation $P_T$ of $\{1, \ldots, T\}$.\label{algorithm_loop}
+\item For each $t \in P_T$:
+\begin{enumerate}
+\item Propose a candidate $\boldsymbol{\theta}^*$ with $\theta_{(t)}$ updated by a normal proposal distribution, $N(\theta_{(t)}, 1)$
 
-\item Propose a candidate $\boldsymbol{\theta}^*$ with one $\theta_{(t)}$ updated by a normal proposal distribution, $N(\theta_{(t)}, 1)$
+\item Use $\boldsymbol{\theta}^*$ to calculate a new set of $f^*$ for source $j$, $j=1, 2, 3$, via model~\eqref{modL} or model~\eqref{modC}, and find the associated $F^*$ for each case by putting the vector $(f^*, f_4=0)$ in equation~\eqref{capF}
 
-\item Use $\boldsymbol{\theta}^*$ to calculate a new set of $f^*$ for source $j$, $j=1, 2, 3$, via model~(\ref{modL}) or model (4), and find the associated $F^*$ for each case by putting the vector $(f^*, f_4=0)$ in equation~(\ref{capF})
+\item Compute the acceptance probability $\alpha$, a product of the likelihood ratio, the proposal ratio and the prior ratio, given by
+\begin{align*}
+\alpha=\frac{L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)}{L\Big(F; \text{ST}, \boldsymbol{Z}\Big)}\frac{Q\Big(\theta_{(t)} \vert \theta_{(t)}^*\Big)}{Q\Big(\theta_{(t)}^* \vert \theta_{(t)}\Big)}\frac{p\Big(\theta_{(t)}^*\Big)}{p\Big(\theta_{(t)}\Big)},
+\end{align*}
+where the likelihood $L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)$ involves the rurality covariate $\boldsymbol{Z}$ resulting from the attribution probability, $p(\text{source }j\vert \boldsymbol{Z}$) with an assumption that the genotype distribution on sources is independent of the covariate.
+\item Sample $u$ from $U(0,1)$.
+If $\alpha$ meets $u < \min \left\{1, \alpha \right\}$, in the next state the current proposals of $f^{(*)}$, $F^{(*)}$ and $L\Big(F^{(*)}; \text{ST}, \boldsymbol{Z}\Big)$ are updated; otherwise, the previous proposals are retained.
+\end{enumerate}
+\item Repeat from step \ref{algorithm_loop} for the given number of iterations.
 
-\item Compute the acceptance rate $\alpha$, a product of the likelihood ratio, the proposal ratio and the prior ratio, given by,
-
-\begin{align*}
-\frac{L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)}{L\Big(F; \text{ST}, \boldsymbol{Z}\Big)}\frac{Q\Big(\theta_{(t)} \vert \theta_{(t)}^*\Big)}{Q\Big(\theta_{(t)}^* \vert \theta_{(t)}\Big)}\frac{p\Big(\theta_{(t)}^*\Big)}{p\Big(\theta_{(t)}\Big)},
-\end{align*}
-
-where the likelihood $L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)$ involves the rurality covariate $\boldsymbol{Z}$ resulting from the attribution probability, $p(\text{source }j\vert \boldsymbol{Z}$) with an assumption that the genotype distribution on sources is independent of the covariate.
-Then, a random value of $u$ is generated from $U(0,1)$.
-If $\alpha$ meets $u < \min \left\{1, \alpha \right\}$, in the next state the current proposals of $f^{(*)}$, $F^{(*)}$ and $L\Big(F^{(*)}; \text{ST}, \boldsymbol{Z}\Big)$ are updated; otherwise, the previous proposals are retained.
 \end{enumerate}

### 2018-02-05 paper/finalish/article_1023.tex
update equation references to use eqref

@@
 where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality.
-Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~(\ref{modL}) can be rewritten as
+Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~\eqref{modL} can be rewritten as
 \begin{equation}
@@
 \end{equation*}
-As a consequence, the estimated attribution probabilities are obtained via equation~(\ref{capF}) after fitting the data to model~\eqref{modL} or to model~\eqref{modC}.
+As a consequence, the estimated attribution probabilities are obtained via equation~\eqref{capF} after fitting the data to model~\eqref{modL} or to model~\eqref{modC}.
 
@@
 In the interest of quantifying the uncertainty of the posterior attribution probability, we perform Bayesian inference for source attribution probabilities using Markov chain Monte Carlo (MCMC) methods.
-Assume the priors on parameters of interest in model~(\ref{modL}) and model (4) follow a standard normal distribution.
+Assume the priors on parameters of interest in model~\eqref{modL} and model~\eqref{modC} follow a standard normal distribution.
 Let $\boldsymbol{\theta}$ denote the vector of parameters, with elements $\theta_{t}$ for $t=1, \ldots, T$.
@@
 The associated posterior mean of attribution probability for each source was invariant despite the fact that the proposals of $f$ jumped even wider in the random walk.
-This is because the denominator of the equation~(\ref{capF}) balances the impact made by the values of $f$.
+This is because the denominator of the equation~\eqref{capF} balances the impact made by the values of $f$.
 Further, we included a rurality covariate in the model fitting.

### 2018-02-05 paper/finalish/article_1023.tex
update the paragraph (and table) on DIC

@@
 
-Lastly, we use deviance information criterion (DIC) to evaluate if the linear model or the categorical model is more suitable to describe the data with different time periods of study.
-Given the sequence type information from the asymmetric Island model and the simple model, Table~(\ref{tab3}) compares the DIC values of the linear model to that of the categorical model, applied to data with the whole period of study and to data with the post-intervention time period.
-It tells that the linear model always performs better than the categorical model as this model is penalized more in DIC due to the larger number of parameters.
-This shows that the assumption of a linear effect on the rurality scales against the source attribution is reasonable.
+Lastly, we use deviance information criterion (DIC) for model comparison.
+DIC values obtained from our MCMC runs are displayed in Table~\ref{tab3}.
+Overall there is a clear signal that a linear representation of rurality (on the logit scale) is adequate.
+Note that the asymmetric Island and Dirichlet models are not directly comparable by DIC as the likelihoods are on different scales: the Dirichlet model assumes all potential sequence types have been observed so that $\sum_j p(\boldsymbol{\pi}_j) = 1$, whereas the asymmetric Island model allows for unobserved sequence types so that $\sum_j p(\boldsymbol{\pi}_j) < 1$.
 
@@
       \toprule
-      & \multicolumn{2}{c}{Island model}  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Simple model} \\
+      & \multicolumn{2}{c}{Island model}  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Dirichlet model} \\
       Data & Linear & Categorical & & Linear & Categorical\\ \midrule
@@
   \end{center}
-  \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 (whole period) and to the data from 2008 to 2014 (post-intervention period) given the genotypic information is from the asymmetric Island model or the simple model.}
+  \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 (whole period) and to the data from 2008 to 2014 (post-intervention period) given the genotypic information is from the asymmetric Island model or the Dirichlet model.}
   \label{tab3}

### 2018-02-05 paper/finalish/article_1023.tex
update figure legends

@@
 \includegraphics[width=.9\linewidth]{Figures/FCILandC(5)}
-\caption{The posterior mean of samples $F$ with 80\% HPD interval for source: poultry (red), ruminants (green), water (blue) and others (pink) over the rurality scales from highly rural area to main urban area.
-The samples of attribution probability are generated from the two respective fitted models: the linear model (upper panel) and the categorical model (lower panel), given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the simple model).}
+\caption{Posterior mean attribution probabilities ($F$) with 80\% credible intervals for source: poultry (red), ruminants (green), water (blue) and others (pink) over the rurality scales from highly rural area to main urban area.
+The samples of attribution probability are generated from the two respective fitted models: the linear model (upper panel) and the categorical model (lower panel), given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
 \label{fig2}
@@
 \includegraphics[width=.9\linewidth]{Figures/FCILandC(8)}
-\caption{The posterior mean of samples $F$ with 80\% HPD interval for four sources over seven rurality categories after applying our methods to data with the post-intervention time period from 2008 to 2014.
-The samples of attribution probability are generated from the two respective fitted models -- the linear model (upper panel) and the categorical model (lower panel) -- given the genotype distribution based upon the asymmetric Island model or the simple model.}
+\caption{Posterior mean attribution probabilities ($F$) with 80\% credible interval for four sources over seven rurality categories after applying our methods to data with the post-intervention time period from 2008 to 2014.
+The samples of attribution probability are generated from the two respective fitted models -- the linear model (upper panel) and the categorical model (lower panel) -- given the genotype distribution based upon the asymmetric Island model or the Dirichlet model.}
 \label{fig3}

### 2018-02-06 paper/figures.R
adds code for figure

@@ -1,0 +1,112 @@
+library(islandR)
+library(dplyr)
+library(tidyr)
+
+# Fit the sequence type distribution using the island model
+st_i = replicate(10, {
+  s=st_fit(formula = Source ~ ST,
+            non_primary = "Human",
+            data = manawatu,
+            method="island",
+            sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
+            iters=10000)
+  lapply(1:100, function(i) { data.frame(ST=s$types, s$sampling_distribution[,,i], Iteration=i) })
+  }
+  )
+
+# Reformat to a data.frame
+st_il <- apply(st_i, 2, function(x) { do.call(rbind, x) })
+st_il2 <- lapply(seq_along(st_il), function(x) { cbind(st_il[[x]], Chain=x)})
+st_id <- do.call(rbind, st_il2) %>% mutate(Iteration = Iteration + (Chain-1)*100) %>%
+  select(-Chain)
+
+# and using the Dirichlet
+st_d = st_fit(formula = Source ~ ST,
+            non_primary = "Human",
+            data = manawatu,
+            method="dirichlet",
+            iters=1000)
+
+st_dl <- lapply(1:1000, function(i) { data.frame(ST=as.numeric(st_d$types), st_d$sampling_distribution[,,i], Iteration=i) })
+st_dd <- do.call(rbind, st_dl)
+
+# Join them up
+st <- bind_rows(cbind(st_id, Model='Island'), cbind(st_dd, Model='Dirichlet')) %>%
+  gather(Source, P, -ST, -Iteration, -Model)
+
+#final <- st %>% group_by(Model, Iteration, Source) %>%
+#  mutate(P = P/sum(P)) %>% ungroup %>%
+#  spread(Model, P)
+
+final <- st %>% group_by(Iteration, Model, ST) %>%
+  mutate(P = P/sum(P)) %>% spread(Model, P)
+
+#save(list='final', file='model_compare.Rdata')
+#load('model_compare.Rdata')
+
+# OK, now the Kolmogorov-Smirnov statistic
+kolm_smirn <- function(x, y) {
+  x <- sort(x)
+  y <- sort(y)
+  ex <- seq_along(x)/length(x)
+  ey <- seq_along(y)/length(y)
[…]

### 2018-02-06 R/sample_dist_dirichlet.R
dirichlet ignored iters in favour of iters=100

@@ -59,7 +59,7 @@ st_fit_dirichlet <- function(formula, prior = 1, non_primary = "Human", iters = 
     a
   }
 
-  out <- replicate(100, dirichlet_matrix(alpha))
+  out <- replicate(iters, dirichlet_matrix(alpha))
   # righto, now construct a useful object...
   x = list(types = rownames(counts), sources = source_names,
            sampling_distribution = out, model = "dirichlet")

### 2018-02-06 R/RcppExports.R
adds burnin/samples/thin to island model for consistency

@@ -5,7 +5,7 @@ log_lik <- function(humans, phi, p) {
     .Call('_islandR_log_lik', PACKAGE = 'islandR', humans, phi, p)
 }
 
-island <- function(isolates, beta_migration, gamma_recombination, niter = 10000L) {
-    .Call('_islandR_island', PACKAGE = 'islandR', isolates, beta_migration, gamma_recombination, niter)
+island <- function(isolates, beta_migration, gamma_recombination, samples = 100L, burnin = 10L, thin = 100L) {
+    .Call('_islandR_island', PACKAGE = 'islandR', isolates, beta_migration, gamma_recombination, samples, burnin, thin)
 }
 

### 2018-02-06 R/sample_dist_island.R

@@ -14,13 +14,14 @@ NULL
 #' @param sequences A formula of the form ~ASP + GLN + GLT specifying columns for the allelic profile of each type
 #' @param non_primary one or more sources that should be considered 'output' sources. No genotype distribution is computed for these,
 #'        but P(ST | source) will be computed for these STs in addition to those observed on other sources.
-#' @param iters the number of iterations to sample. An additional burnin period of 10% of this value will be used, and
-#'        100 samples will be taken (or every sample, whichever is smaller).
+#' @param samples the number of samples required from the posterior after burnin. Defaults to 100.
+#' @param burnin  the number of samples to discard for burnin. Defaults to 10.
+#' @param thin    the number of iterations per sample taken. Defaults to 100.
 #' @param priors - a list of priors for the fit. Defaults to 1,1.
 #' @param data optional data frame from which to take variables in \code{formula} and \code{sequence}.
 #' @return an object of class island which derives from sampling_dist.
 #' @seealso \code{\link{st_fit}}, \code{\link{print.sample_dist}}, \code{\link{plot.sample_dist}}, \code{\link{summary.sample_dist}}
-st_fit_island <- function(formula, sequences, non_primary = "Human", iters = 10000, priors = list(migration=1,recombination=1), data) {
+st_fit_island <- function(formula, sequences, non_primary = "Human", samples = 100, burnin = 10, thin = 100, priors = list(migration=1,recombination=1), data) {
   mod.terms = terms(formula, data=data)
   mod.frame = model.frame(formula, data=data)
   allele.frame = model.frame(sequences, data=data)
@@ -65,7 +66,7 @@ st_fit_island <- function(formula, sequences, non_primary = "Human", iters = 100
   island.mat = as.matrix(rbind(source.frame, type.frame))
 
   # run the island model
-  out = island(island.mat, beta_migration = priors$migration, gamma_recombination = priors$recombination, niter = iters)
+  out = island(island.mat, beta_migration = priors$migration, gamma_recombination = priors$recombination, samples = samples, burnin = burnin, thin = thin)
 
   # now regenerate useful summary information... (in future assign names in C++ land?) Also
   # need to change output in C++ land so it gives all types rather than duplicates and only

### 2018-02-06 src/RcppExports.cpp

@@ -19,23 +19,25 @@ BEGIN_RCPP
 END_RCPP
 }
 // island
-List island(IntegerMatrix isolates, NumericVector beta_migration, NumericVector gamma_recombination, int niter);
-RcppExport SEXP _islandR_island(SEXP isolatesSEXP, SEXP beta_migrationSEXP, SEXP gamma_recombinationSEXP, SEXP niterSEXP) {
+List island(IntegerMatrix isolates, NumericVector beta_migration, NumericVector gamma_recombination, int samples, int burnin, int thin);
+RcppExport SEXP _islandR_island(SEXP isolatesSEXP, SEXP beta_migrationSEXP, SEXP gamma_recombinationSEXP, SEXP samplesSEXP, SEXP burninSEXP, SEXP thinSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
     Rcpp::traits::input_parameter< IntegerMatrix >::type isolates(isolatesSEXP);
     Rcpp::traits::input_parameter< NumericVector >::type beta_migration(beta_migrationSEXP);
     Rcpp::traits::input_parameter< NumericVector >::type gamma_recombination(gamma_recombinationSEXP);
-    Rcpp::traits::input_parameter< int >::type niter(niterSEXP);
-    rcpp_result_gen = Rcpp::wrap(island(isolates, beta_migration, gamma_recombination, niter));
+    Rcpp::traits::input_parameter< int >::type samples(samplesSEXP);
+    Rcpp::traits::input_parameter< int >::type burnin(burninSEXP);
+    Rcpp::traits::input_parameter< int >::type thin(thinSEXP);
+    rcpp_result_gen = Rcpp::wrap(island(isolates, beta_migration, gamma_recombination, samples, burnin, thin));
     return rcpp_result_gen;
 END_RCPP
 }
 
 static const R_CallMethodDef CallEntries[] = {
     {"_islandR_log_lik", (DL_FUNC) &_islandR_log_lik, 3},
-    {"_islandR_island", (DL_FUNC) &_islandR_island, 4},
+    {"_islandR_island", (DL_FUNC) &_islandR_island, 6},
     {NULL, NULL, 0}
 };
 

### 2018-02-06 src/asymmetric_island.cpp

@@ -174,7 +174,7 @@ void append_traces(int iter, NumericMatrix &A, NumericMatrix &R, double lik, Num
 }
 
 /* This version uses the clonal frame version of the likelihood */
-void Island::mcmc6f(const double beta, const double gamma_, const int niter, const int thin) {
+void Island::mcmc6f(const double beta, const double gamma_, const int samples, const int burnin, const int thin) {
 	precalc();
 
   /* Initialize the random number generator */
@@ -211,7 +211,7 @@ void Island::mcmc6f(const double beta, const double gamma_, const int niter, con
 	double sigma_r = 0.5;							//	factor for normal proposal in MH change of r (case 5)
 
 	/* Trace output matrix */
-	evolution_traces = NumericMatrix(niter/thin+1, 1+ng*(ng+1)+ng+1); // iter, A, r, loglik
+	evolution_traces = NumericMatrix(samples+1, 1+ng*(ng+1)+ng+1); // iter, A, r, loglik
 	int trace_row = 0;
 	append_traces(0, A, R, loglikelihood, evolution_traces, trace_row++);
 
@@ -219,12 +219,11 @@ void Island::mcmc6f(const double beta, const double gamma_, const int niter, con
 //	clock_t next = start + (clock_t)CLOCKS_PER_SEC;
 //	Rcout << "Done 0 of " << niter << " iterations";
 
-	const int burnin = (int)floor((double)niter*.1);
-	const int inc = std::max((int)floor((double)niter/100.),1);
-	for (int iter = 0; iter < niter+burnin; iter++) {
-		if (iter>=burnin && (iter-burnin)%inc==0) {
+	for (int iter = 0; iter < (samples+burnin)*thin; iter++) {
+		if ((iter+1) % thin == 0 && iter/thin >= burnin) {
 
-			/* Now dump the likelihoods for this iteration.
+		  /* JM Changed: Thinning is now the same - we just want given number of samples */
+			/* Dump the likelihoods for this iteration.
 			   Weird that this has nothing to do with the thinning, which applies only
 		     to the evolutionary parameters. */
 
@@ -358,7 +357,7 @@ void Island::mcmc6f(const double beta, const double gamma_, const int niter, con
 		}
 
 		/* output thinned traces of island model fit */
-		if(iter >= burnin && (iter+1)%thin==0)
+		if((iter+1)%thin==0 && iter/thin >= burnin)
 		  append_traces(iter+1, A, R, loglikelihood, evolution_traces, trace_row++);
 
 //		if((current=clock())>next) {

### 2018-02-06 src/asymmetric_island.h

@@ -39,8 +39,8 @@ public:
 	}
   void initialise(Rcpp::IntegerMatrix isolates);
 
-	// mcmc6f infers M and R from seqs of known origin, and runs 100 side-chains to infer F given M and R
-	void mcmc6f(const double beta, const double gamma_, const int niter, const int thin);
+	// mcmc6f infers M and R from seqs of known origin, sampling nsamples after nburnin samples, with given thinning
+	void mcmc6f(const double beta, const double gamma_, const int samples, const int burnin, const int thin);
 
 	~Island() {
 		/* free memory */

### 2018-02-06 src/island_model.cpp

@@ -7,7 +7,7 @@ using namespace Rcpp;
 // run island model
 
 // [[Rcpp::export]]
-List island(IntegerMatrix isolates, NumericVector beta_migration, NumericVector gamma_recombination, int niter = 10000) {
+List island(IntegerMatrix isolates, NumericVector beta_migration, NumericVector gamma_recombination, int samples = 100, int burnin = 10, int thin = 100) {
 
   // create model class
   Island island;
@@ -16,11 +16,7 @@ List island(IntegerMatrix isolates, NumericVector beta_migration, NumericVector 
   double beta  = beta_migration[0];
   double gamma = gamma_recombination[0];
 
-  int thin  = 50;
-
-  // output traces. The thin here is how often we sample the evolution parameters.
-  // it is NOT how often we sample the probabilities, which is the only thing we care about ATM.
-  island.mcmc6f(beta, gamma, niter, thin);
+  island.mcmc6f(beta, gamma, samples, burnin, thin);
 
   List out;
   out["hum_lik"] = island.human_likelihoods;

### 2018-02-06 tests/testthat/helper_setup.R

@@ -4,4 +4,4 @@ library(islandR)
 set.seed(123)
 
 # some helpful fits with which to test stuff
-st = st_fit(Source ~ ST, sequences=~ASP+GLN+GLT+GLY+PGM+TKT+UNC, non_primary = "Human", method="island", data=manawatu, iters=100)
+st = st_fit(Source ~ ST, sequences=~ASP+GLN+GLT+GLY+PGM+TKT+UNC, non_primary = "Human", method="island", data=manawatu, samples=100)

### 2018-03-16 R/attribution.R
add a note, adn some commentary on the naming of columns returned from the attribution summary functions

@@ -227,7 +227,8 @@ print.attribution <- function(x, ...) {
   p_pred = list()
   for (j in seq_along(post_theta)) {
     p_pred[[j]] = t(exp(X %*% t(post_theta[[j]])))
-    colnames(p_pred[[j]]) = apply(X, 1, function(x, y) { paste(y[x == 1], collapse=":")}, colnames(X))
+    # TODO: This is a bit icky. We can't rely on this within the functions that use this
+    colnames(p_pred[[j]]) = apply(X, 1, function(x, y) { paste(paste0(y, '=', x), collapse=":")}, colnames(X))
   }
   p_pred[[length(p_pred)+1]] = matrix(1, nrow(p_pred[[1]]), ncol(p_pred[[1]]))
 

### 2018-03-16 paper/attr_figures.R
adds code to generate figures (excluding the barplot one)

@@ -1,0 +1,82 @@
+library(islandR)
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(forcats)
+
+manawatu <- read.csv("~/data/sa_report/data/extract_attribution.csv") %>%
+  mutate(SampledDate = as.Date(SampledDate), Year = year(SampledDate)) %>%
+  filter(Year >= 2005, Year <= 2014) %>%
+  mutate(Source4 = fct_collapse(Source, Poultry = c("Supplier A", "Supplier B", "Supplier Other", "Otherpoultry"),
+                                Ruminants = c("Cattle", "Sheep"),
+                                Water = "Environmental water",
+                                Other = c("Cat_dog_pet", "Duck_poultry", "Spent_hen", "Pig",
+                                          "Turkey", "Water_bird_wild", "Wild_bird_other"))) %>%
+  filter(Source4 != "Human" | !is.na(UR_num)) %>%
+  mutate(UR_fact = factor(UR_num))
+
+# Fit the sequence type distribution using the island model
+st_i = st_fit(formula = Source4 ~ ST,
+         non_primary = "Human",
+         data = manawatu,
+         method="island",
+         sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
+         samples = 1000)
+
+# and using the Dirichlet
+st_d = st_fit(formula = Source4 ~ ST,
+              non_primary = "Human",
+              data = manawatu,
+              method="dirichlet",
+              iters=1000)
+
+# do the attribution
+at_if <- attribution(ST ~ UR_fact, st_i, manawatu %>% filter(Source4 == "Human"))
+at_il <- attribution(ST ~ UR_num, st_i, manawatu %>% filter(Source4 == "Human"))
+at_df <- attribution(ST ~ UR_fact, st_d, manawatu %>% filter(Source4 == "Human"))
+at_dl <- attribution(ST ~ UR_num, st_d, manawatu %>% filter(Source4 == "Human"))
+
+df_if <- predict(at_if, FUN=identity) %>% extract(X, into="UR_num", regex="UR_fact([-0-9]+)=1", convert=TRUE) %>%
+  replace_na(replace=list(UR_num=-3)) %>% mutate(GenotypeModel = 'Island', AttributionModel='Categorical')
[…]

### 2018-03-16 paper/attr_figures_2009.R

@@ -1,0 +1,83 @@
+library(islandR)
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(forcats)
+
+manawatu <- read.csv("~/data/sa_report/data/extract_attribution.csv") %>%
+  mutate(SampledDate = as.Date(SampledDate), Year = year(SampledDate)) %>%
+  filter(Year >= 2005, Year <= 2014) %>%
+  mutate(Source4 = fct_collapse(Source, Poultry = c("Supplier A", "Supplier B", "Supplier Other", "Otherpoultry"),
+                                Ruminants = c("Cattle", "Sheep"),
+                                Water = "Environmental water",
+                                Other = c("Cat_dog_pet", "Duck_poultry", "Spent_hen", "Pig",
+                                          "Turkey", "Water_bird_wild", "Wild_bird_other"))) %>%
+  filter(Source4 != "Human" | (!is.na(UR_num) & Year >= 2008) ) %>%
+  mutate(UR_fact = factor(UR_num))
+
+# Fit the sequence type distribution using the island model
+st_i = st_fit(formula = Source4 ~ ST,
+         non_primary = "Human",
+         data = manawatu,
+         method="island",
+         sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
+         samples = 1000)
+
+# and using the Dirichlet
+st_d = st_fit(formula = Source4 ~ ST,
+              non_primary = "Human",
+              data = manawatu,
+              method="dirichlet",
+              iters=1000)
+
+# do the attribution
+at_if <- attribution(ST ~ UR_fact, st_i, manawatu %>% filter(Source4 == "Human"))
+at_il <- attribution(ST ~ UR_num, st_i, manawatu %>% filter(Source4 == "Human"))
+at_df <- attribution(ST ~ UR_fact, st_d, manawatu %>% filter(Source4 == "Human"))
+at_dl <- attribution(ST ~ UR_num, st_d, manawatu %>% filter(Source4 == "Human"))
+
+df_if <- predict(at_if, FUN=identity) %>% extract(X, into="UR_num", regex="UR_fact([-0-9]+)=1", convert=TRUE) %>%
+  replace_na(replace=list(UR_num=-3)) %>% mutate(GenotypeModel = 'Island', AttributionModel='Categorical')
[…]

### 2018-03-16 paper/figures.R

@@ -42,7 +42,7 @@ final <- st %>% group_by(Iteration, Model, ST) %>%
   mutate(P = P/sum(P)) %>% spread(Model, P)
 
 save(list='final', file='model_compare.Rdata')
-#load('model_compare.Rdata')
+load('model_compare.Rdata')
 
 # OK, now the Kolmogorov-Smirnov statistic
 kolm_smirn <- function(x, y) {
@@ -106,26 +106,29 @@ diff %>% left_join(dist %>%
   left_join(manawatu %>% select(ST2=ST, Source) %>% group_by(ST2) %>% summarize(Human=sum(Source=="Human"), Source=sum(Source != "Human")), by="ST2") %>%
   filter(ST %in% c(403, 2343, 2026, 474, 45))
 
-sts <- c(45, 2343, 2026, 403, 474)
+sts <- c(403, 2343, 2026, 474)
 plot_dat <- final %>% filter(ST %in% sts) %>%
   gather(Model, P, Dirichlet, Island) %>% ungroup %>% mutate(ST = factor(ST, levels=sts, labels = paste0("ST-", sts)),
                                                              Scale = ifelse(Model == "Dirichlet" & ST %in% c("ST-403","ST-2343"), 2, 1.2))
 
 library(ggjoy)
-library(tikzDevice)
-tikz("genotype_figure.tex", standAlone = TRUE, width=5, height=6)
-#pdf("genotype_figure.pdf", width=5, height=6)
+#library(tikzDevice)
+#tikz("genotype_figure.tex", standAlone = TRUE, width=5, height=6)
+cairo_pdf("genotype_figure.pdf", width=7, height=4)
 ggplot(plot_dat) + geom_joy(aes(x=P, y=Source, fill=Model, scale=Scale), alpha=0.7, size=0.1, bandwidth=0.01) +
-  facet_wrap(~ST, ncol=1) +
+  facet_wrap(~ST, ncol=2) +
   ylab("") +
-  scale_x_continuous(name = "$P(\\mathsf{Source} \\mid \\mathsf{ST})$", limits=c(0,1), expand = c(0,0)) +
+  scale_x_continuous(name = "P(Source | ST)", limits=c(0,1), expand = c(0,0)) +
   scale_fill_manual(values = c("steelblue", "brown")) +
   theme_bw() +
-  theme(legend.position = c(0.87,0.935),
+  theme(text = element_text(family="Times"),
+        legend.position = c(0.92,0.895),
         legend.box.background = element_rect(),
[…]

### 2018-03-16 R/water.R
some more stuff for water attribution

@@ -1,3 +1,80 @@
+log_likelihood <- function(node, F) {
+  # run over data in this and any downstream nodes, multiplying shit up
+  sinks <- nodes[[node$sinks]]
+  p <- numeric(length(node$y))
+  for (i in 1:length(node$sinks))
+    p <- p + nodes[[i]]$p * F[i]
+  node$y * log(p)
+}
+
+delogit <- function(p) {
+  P <- exp(cbind(p, 0))
+  P/rowSums(P)
+}
+
+# The model object is a list of sources/sinks. Each node
+# has data, current probability dist, links to sources and
+# links to sinks. The links to sources are associated with
+# parameter values and a design matrix.
+update_nodes <- function(nodes) {
+  # run through the model nodes, updating each one
+  for (i in 1:length(nodes)) {
+    # if we're a sink, run over the parameters
+    node <- nodes[[i]]
+    rows <- sample(nrow(node$theta))
+    for (r in rows) {
+      # for each source, sampled at random
+      cols = sample(ncol(node$theta))
+      for (c in cols) {
+        # update theta[r,c] with RWMH
+        theta = node$theta
+        theta[r,c] = rnorm(1, theta[r,c], theta_proposal_sigma)
+        # compute new f
+        f = node$X %*% theta
+
+        # proposal ratio is symmetric, so need only the prior ratio
+        t0_prop = theta[,j]      - theta_0[,j]
+        t0_curr = node$theta[,j] - theta_0[,j]
+        # exp(-0.5*(t(t0_prop) %*% theta_prec %*% t0_prop)) / exp(-0.5*(t(t0_curr) %*% theta_prec %*% t0_curr))
+
+        log_hastings_ratio = -0.5*(t(t0_prop) %*% theta_prec %*% t0_prop -
+                                     t(t0_curr) %*% theta_prec %*% t0_curr)
+
+        # transform f from logit scale
+        F = delogit(f)
+
+        # use source mixing to compute ST distribution for this sink.
+
+        # Hmm, how do covariates at a sink/source level work? The next sink
+        # down doesn't care about those covariates, right?? So if we get an
+        # improved attribution for subclasses of water how does that flow
+        # through to the end result? What we normally end up with is the
+        # ST dist given the covariate pattern (i.e. covariate pattern alters
+        # the weights of the source ST dists). What we want is the ST dist
+        # marginal. P(ST) = sum_i P(ST | covariate_i) P(covariate_i)
+
+        # p will be a matrix of size #ST times #covariate patterns (X)
[…]

### 2018-03-16 paper/finalish/article_1023.tex
updates text and figures to the new ones

--- 2018-02-05 paper/finalish/article_1023.tex
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/FCILandC(5)}
-\caption{The posterior mean of samples $F$ with 80\% HPD interval for source: poultry (red), ruminants (green), water (blue) and others (pink) over the rurality scales from highly rural area to main urban area.
-The samples of attribution probability are generated from the two respective fitted models: the linear model (upper panel) and the categorical model (lower panel), given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the simple model).}
+\includegraphics[width=.9\linewidth]{Figures/attr_all.pdf}
+\caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas.
+The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
 \label{fig2}
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/FCILandC(8)}
-\caption{The posterior mean of samples $F$ with 80\% HPD interval for four sources over seven rurality categories after applying our methods to data with the post-intervention time period from 2008 to 2014.
-The samples of attribution probability are generated from the two respective fitted models -- the linear model (upper panel) and the categorical model (lower panel) -- given the genotype distribution based upon the asymmetric Island model or the simple model.}
+\includegraphics[width=.9\linewidth]{Figures/attr_2009.pdf}
+\caption{Posterior mean attribution ($F$) of human cases from the post-intervention time period (2008-2014) with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area.
+The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
 \label{fig3}
@@
 
-To investigate the association between source attribution and human infection, we apply the methods demonstrated in the preceding sections to the campylobacteriosis data and also rearrange the data frame from the ST-based structure to the case-based one for model fitting.
+To investigate the association between source attribution and human infection, we apply the methods demonstrated in the preceding sections to the campylobacteriosis data from the Manawatu in New Zealand.
 The posterior mean of samples $F$ with the 80\% highest posterior density (HPD) interval for each source is illustrated on each rurality grade in Figure~(\ref{fig2}).

### 2018-03-18 paper/finalish/article_1023.tex
update tables to use consistent STs and commentary in results

@@
       \toprule
-      ST & \textit{asp} & \textit{gln} & \textit{glt} & \textit{gly} & \textit{pgm}& \textit{tkt}& \textit{unc}\\ \midrule
-      45  & 4 & 7 & 10 & 4 & 1 & 7 & 1\\
+      ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm}& \textit{tkt}& \textit{unc}A\\ \midrule
+      403 & 10 & 27 & 16 & 19 & 10 & 5 & 7\\
+      2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
       474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
+      2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
       \bottomrule
@@
   \caption{The allelic profile of genotypes ST-45 and ST-474 is composed of seven allele numbers at each of the seven loci.}
-  \label{tab0}
+  \label{tab:mlst}
 \end{table}
@@
 For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the commonly used 7-gene MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
-An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab0}).
-It shows that the genotypes ST-45 and ST-474 have different allelic combinations across the seven loci.
+An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab:mlst}).
+It shows that the genotypes ST-2026 and ST-474 have different allelic combinations across all seven loci, while ST-403 differs from ST-2026 only at \textit{gln}A, and ST-2343 differs from ST-474 at \textit{glt}A and \textit{pgm}.
 The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made, by comparing the distribution of genotypes from human cases with those from potential reservoirs.
@@
 The total number of unique STs typed from all isolates is $I=355$, and 35\% of genotypes are found among human cases.
-Table~(\ref{tab1}) lists five common genotypes found in human and source isolates.
+Table~\ref{tab:counts} lists five common genotypes found in human and source isolates.
 The first four genotypes are frequently observed in human cases.
-For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-61 are typically found in ruminants.
-Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}.
+For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-2026 are typically found in ruminants.
+Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-2026 are more likely attributable to ruminants \cite{Muell, Coll, Cart}.
 Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.
@@
       45  & 141 & 154 & 10 & 21 & 54\\
-      61 & 64 & 7 & 54 & 2 & 1\\
       474 & 239 & 60 & 15 & 5 & 7\\
+      2026 & 37 & 0 & 46 & 2 & 5\\
       2381 & 0 & 0 & 0 & 21 & 3\\
@@
   \caption{The frequency of five genotypes found from human and four source isolates.}
-  \label{tab1}
+  \label{tab:counts}
 \end{table}
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/attr_all.pdf}
+\includegraphics[width=.8\linewidth]{Figures/attr_all.pdf}
 \caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas.
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/attr_2009.pdf}
+\includegraphics[width=.8\linewidth]{Figures/genotype_figure.pdf}
+\caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
+\label{fig:genotype_dist}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{Figures/attr_2009.pdf}
 \caption{Posterior mean attribution ($F$) of human cases from the post-intervention time period (2008-2014) with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area.
@@
 
-To investigate the association between source attribution and human infection, we apply the methods demonstrated in the preceding sections to the campylobacteriosis data from the Manawatu in New Zealand.
-The posterior mean of samples $F$ with the 80\% highest posterior density (HPD) interval for each source is illustrated on each rurality grade in Figure~(\ref{fig2}).
-Given different genotype distributions, the upper panel of the figure is based on the linear model, whereas the lower panel is generated from the categorical model.
+To investigate the association between source attribution and human infection, we apply the methods described in the preceding sections to the campylobacteriosis data from the Manawatu in New Zealand.
+The posterior mean of $F$ with 80\% credible intervals for each source are illustrated for each rurality grade in Figure~\ref{fig2}.
+The graphs are categorized by the type of genetic model (asymmetric Island or Dirichlet) and the manner in which the rurality variable is modelled (categorical or linear on the logit scale).
 
-For the linear model, the simple model estimates 75\% and 32\% of cases dwelling in main urban areas have acquired the infection from poultry and ruminants, respectively, whereas the asymmetric Island model provides less than 70\% and approximately 30\% of cases attributable to the respective source.
-Further, the 80\% HPD interval of $F$ for poultry and for ruminants based on the simple model seems marginally wider than that for the asymmetric Island model.
-So, there is no significant difference in the proportion of human cases attributed to each source between the asymmetric Island and the simple models.
-For the categorical model, the comparison between the asymmetric Island and the simple models is quite similar to that for the linear model except for the distinguishable attribution for poultry and ruminants in highly rural areas and the much wider credible intervals of $F$ for four sources.
-For the two fitted models, the narrower credible intervals noticed in very rural areas from the asymmetric Island model might indicate that the asymmetric Island model gives a little more accurate sampling distribution for poultry and ruminants genotypes especially typed in humans living in rural areas.
-Further, the relatively larger credible intervals observed in the categorical model might be the effect of allocating the number of cases to each qualitative scale (as a dummy variable), particularly in rural areas due to small observations.
+Overall, the attribution results are relatively stable irrespective of the genetic model or how rurality is represented in the attribution model.
+The majority of human cases are attributed to ruminants and poultry, with more cases attributed to ruminants in rural areas and more cases attributed to poultry in urban areas.
+For both models of the genotype distribution, the linear and categorical models of rurality show broadly the same trend, suggesting that the additional flexibility given by the categorical model is not required, and that the shift in attribution as rurality changes is adequately modelled by a linear trend on the logit scale.
+The linear model has the advantage of tighter credible intervals as it can share data across the 7 levels of rurality, resulting in a clearer separation of ruminant and poultry attribution, particularly in highly rural areas where the data are sparse.
 
-Figure~(\ref{fig2}) shows that poultry starts becoming the dominant source contributable to the infection between rural areas with moderate and with high urban influence.
-In contrast, ruminants are becoming the most attributable source when locations are getting farther away from rural areas with a moderate urban influence.
-As to water and other sources, they seem to be less influenced by the rurality levels because the probabilities remain less than 0.2 over the seven categories.
+There are some small differences between the genotype models, with the Dirichlet model showing a greater attribution to poultry (ranging from 40\% in highly rural areas to 70\% in main urban centres) than the asymmetric Island model (ranging from 30\% in rural areas to 65\% in urban centres).
+This occurs in both the linear and categorical models.
+Interestingly, the asymmetric Island model attributes approximately 7\% of human cases across all rurality levels to sources other than poultry, ruminants and water, and gives a small attribution to water in highly rural areas, whilst the Dirichlet model indicates that both these sources are unimportant.
 
-Moreover, with a view to examining the robustness of our results, the approach is also applied to data with the time period of post-intervention from 2008 to 2014.
-The source attribution at different rurality levels is then illustrated in Figure~(\ref{fig3}), displaying that the overall attribution over four sources is moderately consistent with the patterns we previously discovered.
-However, Figure~(\ref{fig3}) shows a marked decline of attribution probability for poultry and a small rise in probability for ruminants in main urban areas.
-Despite the fact that the intervention did not eliminate the infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.
+The small differences in attribution observed between the asymmetric Island and Dirichlet models may be due to the additional genetic information available to the asymmetric Island model.
+To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that that each source was equally likely.
+Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=7$ and $n=6$ respectively) while only being observed once each among the source isolates, so that the Dirichlet model has little information available to distinguish between sources.
+The asymmetric Island model, however, can exploit the genetic relationship between genotypes.
+ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables \ref{tab:mlst} and \ref{tab:counts}).
+Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources.
+
+As noted previously, a major public health initiative in 2007 led to a significant reduction in the number of cases of campylobacteriosis in New Zealand.
+In order to examine the effects of this change on attribution probabilities, we repeated the analysis using only data from 2008 to 2014.
+Figure~\ref{fig3} shows that the general trend in attribution by rurality is consistent with that seen with all the data, with ruminant associated campylobacteriosis in highly rural areas to poultry associated cases in urban centers linked by a linear trend.
+However, the proportion attributed to poultry across all rurality levels decreased (and thus attribution to ruminants increased) in this post-intervention dataset, being most marked in urban centers.
+Thus, although the intervention did not eliminate infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.
 

### 2018-03-18 paper/attr_figures.R
updates figures for better sizing and y axis labels

@@ -62,21 +62,22 @@ plot_df <- df %>%
 # plot
 library(ggplot2)
 
-cairo_pdf("attr_all.pdf", width=7, height=7)
+cairo_pdf("attr_all.pdf", width=7, height=5)
 ggplot(plot_df) +
   geom_ribbon(aes(x=UR_num, ymin=li, ymax=ui, fill=Source), alpha=0.3) +
   geom_line(aes(x=UR_num, y=m, col=Source), lwd=1) +
   facet_grid(GenotypeModel~AttributionModel) +
   scale_x_continuous(name=NULL, breaks=c(-3,3), labels=c("Highly rural", "Highly Urban"), expand=c(0,0)) +
   scale_y_continuous(name="Percentage of cases", labels=scales::percent_format(), limits=c(0,1), expand=c(0,0)) +
-  scale_color_manual(values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
-  scale_fill_manual(values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
+  scale_color_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
+  scale_fill_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
   theme_bw() +
   theme(text = element_text(family="Times"),
-        legend.position = c(0.98,0.90),
+        legend.position = c(0.99,0.89),
         legend.justification = "right",
         legend.margin=margin(0,0,0,0),
         legend.background = element_rect(fill = 'transparent'),
-        axis.text.x = element_text(hjust=c(-0.1,1.1)))
+        axis.text.x = element_text(hjust=c(-0.1,1.1)),
+        axis.text.y = element_text(vjust=c(-0.1,rep(0.5, 3), 1.1)))
 dev.off()
 

### 2018-03-18 paper/attr_figures_2009.R

@@ -63,21 +63,21 @@ plot_df <- df %>%
 # plot
 library(ggplot2)
 
-cairo_pdf("attr_2009.pdf", width=7, height=7)
+cairo_pdf("attr_2009.pdf", width=7, height=5)
 ggplot(plot_df) +
   geom_ribbon(aes(x=UR_num, ymin=li, ymax=ui, fill=Source), alpha=0.3) +
   geom_line(aes(x=UR_num, y=m, col=Source), lwd=1) +
   facet_grid(GenotypeModel~AttributionModel) +
   scale_x_continuous(name=NULL, breaks=c(-3,3), labels=c("Highly rural", "Highly Urban"), expand=c(0,0)) +
   scale_y_continuous(name="Percentage of cases", labels=scales::percent_format(), limits=c(0,1), expand=c(0,0)) +
-  scale_color_manual(values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
-  scale_fill_manual(values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
+  scale_color_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
+  scale_fill_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
   theme_bw() +
   theme(text = element_text(family="Times"),
-        legend.position = c(0.98,0.90),
+        legend.position = c(0.99,0.89),
         legend.justification = "right",
         legend.margin=margin(0,0,0,0),
         legend.background = element_rect(fill = 'transparent'),
-        axis.text.x = element_text(hjust=c(-0.1,1.1)))
+        axis.text.x = element_text(hjust=c(-0.1,1.1)),
+        axis.text.y = element_text(vjust=c(-0.1,rep(0.5, 3), 1.1)))
 dev.off()
-

### 2018-03-19 paper/finalish/article_1023.tex
adds some discussiony stuff

@@
 \section{Discussions}
-Models that determine the source of human infection, particularly with zoonotic pathogens that originate in animal populations, are of considerable value to public health policy-makers.
-However, the complexity of model building may increase because of the complex transmission pathways of pathogens, in particular when the model combines the epidemiological and molecular data.
-But, is it necessary to use more complex genetic models for the source attribution of zoonoses?
-There is limited research on whether using a simpler model could work as effective as complex genetic models.
+Models that determine the source of human infection, particularly with zoonotic pathogens that originate in animal populations, are of considerable value to public health policy makers.
+However, the complexity of models may increase due to the complex transmission pathways of pathogens, particularly when a model combines epidemiological and molecular data.
+An outstanding question is whether such complexity is required, or whether a simpler model may work as effectively.
 Here we developed a relatively simple model to estimate the attribution probability for each source of \textit{Campylobacter} infection.
-This model differs to the asymmetric Island model, it does not model the pathogenic evolution but uses a simple genetic classification such as a sequence type of each isolate to illustrate the sampling distribution of genotypes.
-With the aim of filling the research gap, our results showed that the simple model can do at least similar work, compared to the asymmetric Island model.
+This model differs to the asymmetric Island model, in that it does not model pathogenic evolution, opting instead to use infer the sampling distribution of genotypes directly from the observed count data.
+
+Our results showed that the simple Dirichlet model and the asymmetric Island model showed largely similar results for the final attribution, with both models demonstrating a clear effect of rurality on attribution: cases in rural areas are more likely to have originated from ruminants, while those in main urban centers are more likely to be of poultry origin.
+While the overall attribution was consistent, it would be expected that the conditional probabilities for a given genotype might differ markedly between models.
+For those genotypes observed infrequently (or not at all) among the sources, the Dirichlet model has little information, while the asymmeteric Island model can exploit additional information from cases with similar (but not identical) genetic profiles.
+In the case of MLST data, with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often.
+Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution well where the bulk of the data lie.
+It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute much to the overall attribution.
+In other circumstances, such as where we have many more than 7 loci, we would have many more rare genotypes, so that the Dirichlet model might provide little useful information.
+At the extreme example of whole genome MLST (wgMLST) where each isolate might be unique, it would provide essentially no information at all.
+In these circumstances, however, the asymmetric Island model would be expected to still perform well, assuming that information could be transferred between similar genotypes.
+
+As with any Bayesian analysis, it is of interest to examine the sensitivity of hte results to the choice of prior distributions.
+We originally used standard normal priors for regression coefficients on the logit scale.
+We also considered...
 

### 2018-03-19 paper/case_rates_by_year.R
adds case rates by year R script

@@ -1,0 +1,12 @@
+# script to produce barplot figure...
+
+library(meshblocknz)
+library(dplyr)
+# 1. Load meshblocknz and compute population by rurality by year
+#    Technically want this only for the meshblocks that we're actually sampling from
+mb2013 %>% filter(DHB_name == "Mid Central") %>% nrow
+
+# 2. Once done, compute number of cases by rurality by year
+
+# Now produce a ggplot barplot...
+

### 2018-03-19 paper/finalish/article_1023.tex
updates case rate figure and text

@@
 Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is quite stable.
-We then calculate the infection rate per 100,000 population and illustrate the prevalence in urban and rural areas from 2006 to 2014 in Figure~(\ref{fig1}).
-The Figure excludes the year of 2005 because the estimates population of both areas is unavailable and the data of typed human cases are incomplete for the entire year.
-As illustrated in the Figure, the infection rates in 2006 are about 60 and 85 in urban and rural areas, respectively.
-We notice that the burden of infection in both areas drop remarkably in 2008, resulting from an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
+The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
 It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.
@@
   \centering
-  \includegraphics[width=.5\linewidth]{Figures/Databar}
-  \caption{The rate per 100,000 population given the typed human cases and the population coming from urban or rural areas in the Manawatu region of New Zealand during 2006 and 2014.
-A poultry intervention conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years, particularly in urban areas.}
-  \label{fig1}
+  \includegraphics[width=.8\linewidth]{Figures/case_rates}
+  \caption{Case rates per 100,000 population in urban and rural areas of the Manawatu region of New Zealand from 2005 through 2014.
+An intervention in the poultry industry conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years in urban areas.}
+  \label{fig:case_rates}
 \end{figure}

### 2018-03-19 paper/case_rates_by_year.R

@@ -2,11 +2,67 @@
 
 library(meshblocknz)
 library(dplyr)
+library(tidyr)
+library(sf)
+library(lubridate)
+library(forcats)
+
 # 1. Load meshblocknz and compute population by rurality by year
 #    Technically want this only for the meshblocks that we're actually sampling from
-mb2013 %>% filter(DHB_name == "Mid Central") %>% nrow
+# load phu dataset
+phu <- read_sf("~/code/epiclustR/shape_files/midcentral_phu.shp")
+
+pop <- mb2013 %>% filter(DHB_name == "Mid Central") %>%
+  filter(MB2006 %in% phu$MB06) %>% select(MB2013, MB2006, Pop2001, Pop2006, Pop2013, UR2006_num, UR2013_num) %>%
+  gather(Year, Population, Pop2001:Pop2013) %>% extract(Year, into="Year", regex="([0-9]+)", convert=TRUE)
+
+# OK, now fit a linear model per meshblock to the population
+pops <- split(pop, pop$MB2013)
+
+predict_popn <- function(dat) {
+  m <- lm(Population ~ Year, data=dat)
+  data.frame(Year=2005:2015, Population=predict(m, data.frame(Year=2005:2015)))
+}
+
+pop_interp <- dplyr::bind_rows(lapply(pops, predict_popn), .id='MB2013') %>% mutate(MB2013 = as.numeric(MB2013)) %>%
+  left_join(mb2013 %>% select(MB2006, MB2013, UR2006_num) %>% unique)
+
+# now rurality by year popn...
 
 # 2. Once done, compute number of cases by rurality by year
+manawatu <- read.csv("extract_attribution.csv") %>%
+  mutate(SampledDate = as.Date(SampledDate), Year = year(SampledDate)) %>%
+  filter(Year >= 2005, Year <= 2014) %>%
+  mutate(Source4 = fct_collapse(Source, Poultry = c("Supplier A", "Supplier B", "Supplier Other", "Otherpoultry"),
+                                Ruminants = c("Cattle", "Sheep"),
+                                Water = "Environmental water",
+                                Other = c("Cat_dog_pet", "Duck_poultry", "Spent_hen", "Pig",
+                                          "Turkey", "Water_bird_wild", "Wild_bird_other"))) %>%
+  filter(Source4 == "Human", !is.na(UR_num))
 
[…]

### 2018-03-19 paper/attr_figures_intervention.R
adds some code to generate the attribution pre/post intervention

@@ -1,0 +1,84 @@
+library(islandR)
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(forcats)
+
+manawatu <- read.csv("~/data/sa_report/data/extract_attribution.csv") %>%
+  mutate(SampledDate = as.Date(SampledDate), Year = year(SampledDate)) %>%
+  filter(Year >= 2005, Year <= 2014) %>%
+  mutate(Source4 = fct_collapse(Source, Poultry = c("Supplier A", "Supplier B", "Supplier Other", "Otherpoultry"),
+                                Ruminants = c("Cattle", "Sheep"),
+                                Water = "Environmental water",
+                                Other = c("Cat_dog_pet", "Duck_poultry", "Spent_hen", "Pig",
+                                          "Turkey", "Water_bird_wild", "Wild_bird_other"))) %>%
+  filter(Source4 != "Human" | !is.na(UR_num) ) %>%
+  mutate(UR_fact = factor(UR_num), Intervention=factor(ifelse(Year < 2008, 'Before', 'After')))
+
+num_samples <- 1000
+
+# Fit the sequence type distribution using the island model
+st_i = st_fit(formula = Source4 ~ ST,
+         non_primary = "Human",
+         data = manawatu,
+         method="island",
+         sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
+         samples = num_samples)
+
+# and using the Dirichlet
+st_d = st_fit(formula = Source4 ~ ST,
+              non_primary = "Human",
+              data = manawatu,
+              method="dirichlet",
+              iters=num_samples)
+
+# do the attribution
+at_if <- attribution(ST ~ UR_num*Intervention, st_i, manawatu %>% filter(Source4 == "Human"))
+at_id <- attribution(ST ~ UR_num*Intervention, st_d, manawatu %>% filter(Source4 == "Human"))
+
+df_il <- predict(at_if, FUN=identity) %>% extract(X, into="UR_num", regex="UR_num=([-0-9]+)", convert=TRUE, remove=FALSE) %>%
+  extract(X, into="Intervention", regex="Intervention([A-Za-z]+)=1") %>%
+  replace_na(replace=list(Intervention="After")) %>%
+  mutate(GenotypeModel = 'Island', AttributionModel='Linear')
[…]

### 2018-03-19 paper/finalish/article_1023.tex
updates discussion

@@
 
-\section{Discussions}
+\section{Discussion}
 Models that determine the source of human infection, particularly with zoonotic pathogens that originate in animal populations, are of considerable value to public health policy makers.
@@
 Our results showed that the simple Dirichlet model and the asymmetric Island model showed largely similar results for the final attribution, with both models demonstrating a clear effect of rurality on attribution: cases in rural areas are more likely to have originated from ruminants, while those in main urban centers are more likely to be of poultry origin.
-While the overall attribution was consistent, it would be expected that the conditional probabilities for a given genotype might differ markedly between models.
-For those genotypes observed infrequently (or not at all) among the sources, the Dirichlet model has little information, while the asymmeteric Island model can exploit additional information from cases with similar (but not identical) genetic profiles.
-In the case of MLST data, with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often.
-Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution well where the bulk of the data lie.
-It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute much to the overall attribution.
-In other circumstances, such as where we have many more than 7 loci, we would have many more rare genotypes, so that the Dirichlet model might provide little useful information.
-At the extreme example of whole genome MLST (wgMLST) where each isolate might be unique, it would provide essentially no information at all.
-In these circumstances, however, the asymmetric Island model would be expected to still perform well, assuming that information could be transferred between similar genotypes.
+As most people in the Manawatu region live in urban centers, this highlights the importance of poultry as a reservoir for campylobacteriosis, which is well established in the literature \cite{MullA, Marsh, MullM, Leve}.
 
-As with any Bayesian analysis, it is of interest to examine the sensitivity of hte results to the choice of prior distributions.
+When we ran models allowing attribution probabilities to differ before and after the intervention in the poultry industry in 2007, we saw a clear difference, with much lower poultry (and higher ruminant) attribution, particularly in main urban centers.
+Considering that most people in the Manawatu region live in urban areas, and that case rates in urban areas decreased from 2008 onwards, it is clear that this intervention coincided with a dramatic reduction in poultry attributed illness as reported elsewhere \cite{AnnS}.
+Given that campylobacteriosis cases in rural areas are mostly attributed to ruminant sources, and that case rates in these areas have been higher than those in urban centers since 2008, there is a clear need for public health interventions to focus in this area.
+
+While the overall attribution was consistent between the Dirichlet and asymmetric Island models, it would be expected that the conditional probabilities for a given genotype might differ markedly.
+For those genotypes observed infrequently (or not at all) among the sources, the Dirichlet model has little information while the asymmeteric Island model can exploit information from cases with similar (but not identical) genetic profiles.
+In the case of MLST data with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often.
+Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution where the bulk of the data lie.
+It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute significantly to the overall attribution.
+In other circumstances, such as where we have many more than 7 loci, we would expect to have many more rare genotypes, so that the Dirichlet model might provide little useful information.
+At the extreme example of whole genome MLST (wgMLST) where each isolate would typically be unique, it would provide essentially no information at all.
+In such circumstances, however, the asymmetric Island model would be expected to still perform well, assuming that information could still be transferred between similar genotypes.
+
+As with any Bayesian analysis, it is of interest to examine the sensitivity of the results to the choice of prior distributions.
 We originally used standard normal priors for regression coefficients on the logit scale.
-We also considered...
+We also considered priors with $\sigma^2=4$, and while this meant that $f$ tended to drift further from 0, the resulting attribution probabilities $F$ did not, largely as the attribution is dominated by poultry and ruminant sources, with the water source in particular being close to zero.
+Thus, $f_\mathsf{poultry}$ and $f_\mathsf{ruminants}$ are positive, while $f_\mathsf{water}$ is negative and the magnitude of these can increase without making all that much difference to their corresponding $F$'s.
+The prior on $f$ thus tends to restrict this ill-behaviour rather than acting as a strong constraint on attribution probabilities.
+The prior $\mathbf{\alpha}_j$ in the Dirichlet also makes little difference if kept small, as it most strongly effects genotypes that are rare, which do not contribute significantly to the overall attribution.
+It can be thought of as augmenting with new data, so that $\mathbf{\alpha}_j=\mathbf{1}$ is equivalent to adding a single observation of each genotype to source $j$.
+Thus, large values of $\mathbf{\alpha}_j$ will cause the genotype distributions across sources to look more similar, and hence result in equal attribution to each source.
 
-In light of the different estimates of the sampling distribution of genotypes from the asymmetric Island and the simple models, two approaches of model fitting (linear and non-linear) were utilized to estimate the attribution probability via the Metropolis-Hastings algorithm.
-Under the assumption of a logit scale on the attribution probability, we examined how the values of $f$ affected the resulting attribution probability.
-We changed the standard deviation of the prior distribution from 1 to 2 in the algorithm.
-The associated posterior mean of attribution probability for each source was invariant despite the fact that the proposals of $f$ jumped even wider in the random walk.
-This is because the denominator of the equation~\eqref{capF} balances the impact made by the values of $f$.
-Further, we included a rurality covariate in the model fitting.
-Our findings not only coincided with the hypothesis that poultry is the major cause of human campylobacteriosis in urban areas \cite{MullA, Marsh, MullM, Leve} but also revealed the linear geographic effect on the infection.
-It means that poultry is less likely responsible for the infection when cases dwell away from urban areas.
-Furthermore, the result of decreasing poultry contribution in main urban areas from fitting the data with post-intervention period accorded with what Sears A.
-and other researchers investigated the effect of poultry intervention \cite{AnnS}.
-This could provide additional information that would help policy-makers to target interventions to populations residing in urban compared to rural areas.
-
-In conclusion, the simple model is comparable to the asymmetric Island model.
-The potential of the inclusion of covariates in the model could be applied to help make a decision on the control of the infection.
-In addition, the simple model (although based on the genetic data) does not capture the evolutionary processes that could be also applied to any type of classification such as phenotype in the research of interest.
-Nonetheless, there is some concern to apply our approach.
-One is the role of water that has been recently addressed in the literature \cite{Wagen} by treating it as a medium instead in the transmission network.
-The other is the larger pathogen datasets resulted from rapidly evolving molecular tools such as the extending MLST schemes typing from seven loci to whole-genome MLST.
-Hence, reconsidering the role of water as a source of infection and selecting the invaluable information from such large pathogenic data to fit the extended model will be the potential challenges in future research.
+Future research might focus on possible extensions to these models.
+One direction is to adapt the models further with additional covariates, which might include age, occupation, and other risk factors such as contact with animals.
+For example, there is evidence that children in rural areas are at higher risk of campylobacteriosis through contact with farm animals \cite{MullM, Spencer2012}.
+Another direction is in expanding the role of water in particular.
+In these models, we've assumed that water is a source of human campylobacteriosis infection, but water differs from the other food and environmental sources in that it is not an amplifying reservoir for \emph{Campylobacter} \cite{Wagen}.
+In contrast, the genotypes found in water might be expected to originate in the other sources present here, particularly ruminants and wild birds, but also potentially from humans as well via discharge of unprocessed human waste.
+Hence, water acts as a transmission pathway from sources to humans, being both an end point (reduced water quality from faecal contamination) and a source (human consumption of water, either recreationally or through untreated water supplies).
+While there is presently little evidence that water is an important source for human campylobacteriosis from the current models, these are fit on sporadic cases of campylobacteriosis.
+However, water is known as a key source of outbreaks of campylobacteriosis, such as the large outbreak in Havelock North, New Zealand in 2016 where an estimated 5,500 out of 14,000 residents became ill \cite{dia_campy_report1}.
+Thus, characterising both characterising the source of \emph{Campylobacter} found in water has important implications for both water quality and public health.
 

### 2018-03-20 paper/finalish/article_1023.tex
adds some acknowledgements
updates paper following review

@@
 \author{Sih-Jing Liao\affil{a},
-Jonathan Marshall\affil{a}\corrauth\ \footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
+Jonathan Marshall\affil{a}\corrauth\footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
 \address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
@@
 Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways.
-Modeling jointly the epidemiological data and genetic information on the microbial isolates derived from cases provides a methodology for tracing back the source of infection.
-The isolate genetic information and the epidemiological data may be modeled at a variety of levels.
+Modeling jointly the epidemiological data and genetic information of microbial isolates derived from cases provides a methodology for tracing back the source of infection.
 In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the extent to which each case resides in a rural compared to urban environment, is estimated.
-A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic data-free model.
-We show that the inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.
+A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic-free model.
+We show that inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.
 
@@
 In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}.
-In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the nature of disease transmission.
-For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease.
+In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, and our ability to intervene depends on how much we know about the nature of disease transmission.
+For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease.
 Quantifying the burden of illness attributable to each pathway and source provides valuable information to health professionals making decisions on public health policy for disease control.
 
-Over the last few decades there have been many models developed for the analysis of zoonotic diseases.
-Epidemiological models make use of the incidence of cases and geographic data to capture the main characteristics of disease spread.
-For example, the Susceptible-Infectious-Recovered model \cite{Kerm} classifies populations into a susceptible, infective and recovered status enabling an estimation of the number of individuals who are likely to be infected.
-Also, models with a combination of spatial and temporal data enable the prediction of the time frame and areas where a disease outbreak is likely to occur \cite{HeldL, Hoehl, Simo}.
-However zoonotic transmission from animals to humans may be complex, involving many sources and exposures linked by differing pathways.
+Zoonotic transmission from animals to humans may be complex, involving many sources and exposures linked by differing pathways.
 For instance, infected wild birds may contaminate environmental water and cause disease spread to water users, either humans or other animals \cite{Wagen}.
@@
   \end{center}
-  \caption{The allelic profile of genotypes ST-45 and ST-474 is composed of seven allele numbers at each of the seven loci.}
+  \caption{The allelic profiles of a selection of genotypes, composed of seven allele numbers at each of the seven loci.}
   \label{tab:mlst}
@@
 Such sequences have sufficient variation to distinguish differing pathogen lineages, while being relatively stable within lineages.
-Each unique nucleotide sequence (allele) at each housekeeping gene (locus) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, with is usually also assigned a number.
+Each unique nucleotide sequence (allele) at each housekeeping gene (locus) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, which is assigned a sequence type (ST) number.
 
@@
 
-In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration.
+In this paper we develop statistical models for source attribution, and demonstrate their use on a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh}.
 Human campylobacteriosis is caused mainly by \textit{C.
 jejuni} and \textit{C.
-coli} which are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}.
-The common symptoms of infection are diarrhea, abdominal pain and fever.
-However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}.
+coli} which are the dominant species associated with approximately 80\% and 15\% of illnesses, respectively \cite{Guert}.
+Common symptoms of infection are diarrhea, abdominal pain and fever, however, a severe complication named Guillain Barr\'{e} syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}.
 The pathogen can be spread between animals, or from animals and wild birds to humans.
-The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces.
+Transmission routes may be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces.
 In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilizes only the prevalence of each type to derive the sampling distribution.
-We then extend both models to incorporate covariates such as rurality indicating the area where a human case resides, exploring the effect of human case rurality on attribution results, modeling attribution by rurality using a linear trend on the logit scale or using separate categories in a Bayesian context, and performing model comparison.
+We then extend both models in a Bayesian context to incorporate covariates, exploring the effect of human case rurality on attribution results via a linear trend on the logit scale or with separate categories, and performing model comparison.
 
@@
 \subsection{Notation and assumptions}
-The ultimate goal of attribution models is to estimate the probability that the observed human cases arise from each putative source.
+The goal of attribution models is to estimate the probability that the observed human cases arise from each putative source.
 Our data comprise microbial genotype information from each observed human case obtained from analysis of stool samples, and also from a pool of non-human cases corresponding to potential zoonotic sources of disease
@@
 Given this genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to the set of genotypes observed among human cases.
-Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.
+Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between different genetic models.
 
 Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection.
-Let $I$ genotypes be the total number of unique sequence types (STs) detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$.
-Typically, the number of detected STs $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.
+Let $I$ genotypes be the total number of unique types detected from all isolates, and denote $n_j$ as the marginal frequency of types found in source $j$, where $\sum_j^J n_j = N$.
+Typically, the number of detected types $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.
 
@@
 \end{align}
-where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, and $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$.
+where $p(\text{ST}_h \vert \text{source }j)$ is the probability that genotype $\text{ST}_h$ arises from the sampling distribution of source $j$, and $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$.
 Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \eqref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.
@@
 With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution.
-Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$.
+Let $x_{ij}$ denote the count of genotype ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$.
 To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
@@
 \end{align}
-where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$.
+where $x_{ij}$ can be $0$, indicating genotypes are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of all types found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$.
 As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
@@
 Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models.
-For each posterior sample, we make inferences for $p(\text{source }j)$ using the likelihood~\eqref{Olikh}.
+For each posterior sample, we infer $p(\text{source }j)$ using the likelihood~\eqref{Olikh}.
 This has the effect of integrating over the uncertainty in $p(\text{ST}_h \vert \text{source }j)$ when estimating $p(\text{source }j)$.
@@
 
-The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014.
-The sampling was carried out over the same time period and from the same geographical location, providing $H+N=3,480$ isolates taken from human and non-human cases, $H=1,400$ of which are from humans.
-The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others.
-The total number of unique STs typed from all isolates is $I=355$, and 35\% of genotypes are found among human cases.
-Table~\ref{tab:counts} lists five common genotypes found in human and source isolates.
-The first four genotypes are frequently observed in human cases.
-For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-2026 are typically found in ruminants.
-Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-2026 are more likely attributable to ruminants \cite{Muell, Coll, Cart}.
-Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.
+The source attribution study described here uses human case data ($H=1,400$) collected in the Manawatu region of New Zealand from March 2005 to December 2014.
+In addition, over the same time period and from the same geographical location, isolates $N=2,080$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others.
+The total number of unique genotypes from all isolates is $I=355$, with 35\% of genotypes found among human cases.
+Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases.
+As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants\cite{Muell, Coll, Cart}.
+The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.
 
-The data also contain two location variables: one is binary, the other is categorical.
-The binary variable specifies whether a typed human case is from urban or rural areas, whereas the categorical variable further splits up the areas into seven levels.
-The levels are highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area.
-Table~(\ref{tab2}) lists the number of typed human cases in each classification of rurality.
-The degree of rurality is a set of integers with a range of [-3,3] corresponding to each defined geographical level.
-Overall, more than 70\% of typed cases lived in urban areas, and approximately 7.5\% of individuals have no information about the location.
-We will neglect the missing data in the following analysis by assuming that these cases are randomly missing.
-Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is quite stable.
+The data also contains location information, in the form of an ordinal classification of urban and rural areas with seven levels coded from -3 to 3: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area.
+Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random.
+Table~\ref{tab2} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census\cite{Stat1, Stat2}.
 The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
@@
       \end{tabular}
-      \caption{Each category of rurality is defined along with the respective number of human cases typed during 2005 and 2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
+      \caption{The number of human cases in each rurality class from 2005--2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
       \label{tab2}
@@
   \caption{Case rates per 100,000 population in urban and rural areas of the Manawatu region of New Zealand from 2005 through 2014.
-An intervention in the poultry industry conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years in urban areas.}
+An intervention in the poultry industry conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years, particularly in urban areas.}
   \label{fig:case_rates}
@@
 To extend this analysis so as to include individual level covariates, we need to calculate subject-specific attribution (conditional) probabilities.
-To that end, let $F_{jh}$ denote the attribution probability of source $j$ for the $h^\textrm{th}$ human case, with constraints $\sum_{j=1}^J F_j =1$ and $0 \leq F_j \leq 1$, where $j=1, \ldots, J$.
+To that end, let $F_{jh}$ denote the attribution probability of source $j$ for the $h^\textrm{th}$ human case, with constraints $\sum_{j=1}^J F_{jh} =1$ and $0 \leq F_{jh} \leq 1$, where $j=1, \ldots, J$.
 We model the probabilities $F_{jh}$ using a linear model on the logit scale, that is,
@@
 where $f_{4h}=0$ is treated as the baseline of $f_{jh}$.
-Consider the case where the ST data for each human case is supplemented by $p$ additional variables.
-We develop a general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$.
-The model for the $h^\textrm{th}$ individual is then of the form
+Consider the case where the genotype data for each human case is supplemented by $p$ additional variables.
+A general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$, then has the form
 \begin{align}
@@
 \end{align}
-so that more covariates can be considered in the model to further investigate how the attribution is associated with the factors of interest.
+for the $h$-th individual.
 Note that if there is a single categorical variable with $L$ levels, then $F_{jh}$ and $f_{jh}$ will take no more than $L$ distinct values.
@@
 
-To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the categorical variable representing the classified location of each human case.
+To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the variable ranging from -3 to 3 representing the classified location of each human case.
 Then two ways of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical.
@@
 \setcounter{enumi}{-1}
-\item Sample $T$ random values from $N(0,1)$ as the initial values points of the parameter set $\boldsymbol{\theta}$ for model~\eqref{modL} or model~\eqref{modC}.
+\item Sample $T$ random values from $N(0,1)$ as initial values of the parameter set $\boldsymbol{\theta}$ for model~\eqref{modL} or model~\eqref{modC}.
 \item Sample a permutation $P_T$ of $\{1, \ldots, T\}$.\label{algorithm_loop}
@@
 \begin{enumerate}
-\item Propose a candidate $\boldsymbol{\theta}^*$ with $\theta_{(t)}$ updated by a normal proposal distribution, $N(\theta_{(t)}, 1)$
+\item Propose a candidate $\boldsymbol{\theta}^*$ with $\theta_{(t)}$ updated by a normal proposal distribution, $Q(\theta_{(t)}^*, \theta_{(t)}) = N(\theta_{(t)}, 1)$.
 
-\item Use $\boldsymbol{\theta}^*$ to calculate a new set of $f^*$ for source $j$, $j=1, 2, 3$, via model~\eqref{modL} or model~\eqref{modC}, and find the associated $F^*$ for each case by putting the vector $(f^*, f_4=0)$ in equation~\eqref{capF}
+\item Use $\boldsymbol{\theta}^*$ to calculate a new set of $f^*$ for source $j$, $j=1, 2, 3$, via model~\eqref{modL} or model~\eqref{modC}, and find the associated $F^*$ for each case by putting the vector $(f^*, f_4=0)$ in equation~\eqref{capF}.
 
@@
 \begin{align*}
-\alpha=\frac{L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)}{L\Big(F; \text{ST}, \boldsymbol{Z}\Big)}\frac{Q\Big(\theta_{(t)} \vert \theta_{(t)}^*\Big)}{Q\Big(\theta_{(t)}^* \vert \theta_{(t)}\Big)}\frac{p\Big(\theta_{(t)}^*\Big)}{p\Big(\theta_{(t)}\Big)},
+\alpha=\frac{L\Big(F^*; \text{ST}\Big)}{L\Big(F; \text{ST}, \boldsymbol{Z}\Big)}\frac{Q\Big(\theta_{(t)} \vert \theta_{(t)}^*\Big)}{Q\Big(\theta_{(t)}^* \vert \theta_{(t)}\Big)}\frac{p\Big(\theta_{(t)}^*\Big)}{p\Big(\theta_{(t)}\Big)},
 \end{align*}
-where the likelihood $L\Big(F^*; \text{ST}, \boldsymbol{Z}\Big)$ involves the rurality covariate $\boldsymbol{Z}$ resulting from the attribution probability, $p(\text{source }j\vert \boldsymbol{Z}$) with an assumption that the genotype distribution on sources is independent of the covariate.
+where the likelihood $L\Big(F^*; \text{ST})$ is given by equation \eqref{Olikh}.
 \item Sample $u$ from $U(0,1)$.
-If $\alpha$ meets $u < \min \left\{1, \alpha \right\}$, in the next state the current proposals of $f^{(*)}$, $F^{(*)}$ and $L\Big(F^{(*)}; \text{ST}, \boldsymbol{Z}\Big)$ are updated; otherwise, the previous proposals are retained.
+If $u < \min \left\{1, \alpha \right\}$, the proposals $f^{(*)}$, $F^{(*)}$ and $L\Big(F^{(*)}; \text{ST}\Big)$ are utilised for the next iteration; otherwise, the previous values are retained.
 \end{enumerate}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=.8\linewidth]{Figures/genotype_figure.pdf}
-\caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
-\label{fig:genotype_dist}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=.8\linewidth]{Figures/attr_2009.pdf}
-\caption{Posterior mean attribution ($F$) of human cases from the post-intervention time period (2008-2014) with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area.
-The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
-\label{fig3}
-\end{figure}
-
-To investigate the association between source attribution and human infection, we apply the methods described in the preceding sections to the campylobacteriosis data from the Manawatu in New Zealand.
-The posterior mean of $F$ with 80\% credible intervals for each source are illustrated for each rurality grade in Figure~\ref{fig2}.
-The graphs are categorized by the type of genetic model (asymmetric Island or Dirichlet) and the manner in which the rurality variable is modelled (categorical or linear on the logit scale).
-
-Overall, the attribution results are relatively stable irrespective of the genetic model or how rurality is represented in the attribution model.
-The majority of human cases are attributed to ruminants and poultry, with more cases attributed to ruminants in rural areas and more cases attributed to poultry in urban areas.
-For both models of the genotype distribution, the linear and categorical models of rurality show broadly the same trend, suggesting that the additional flexibility given by the categorical model is not required, and that the shift in attribution as rurality changes is adequately modelled by a linear trend on the logit scale.
-The linear model has the advantage of tighter credible intervals as it can share data across the 7 levels of rurality, resulting in a clearer separation of ruminant and poultry attribution, particularly in highly rural areas where the data are sparse.
-
-There are some small differences between the genotype models, with the Dirichlet model showing a greater attribution to poultry (ranging from 40\% in highly rural areas to 70\% in main urban centres) than the asymmetric Island model (ranging from 30\% in rural areas to 65\% in urban centres).
-This occurs in both the linear and categorical models.
-Interestingly, the asymmetric Island model attributes approximately 7\% of human cases across all rurality levels to sources other than poultry, ruminants and water, and gives a small attribution to water in highly rural areas, whilst the Dirichlet model indicates that both these sources are unimportant.
-
-The small differences in attribution observed between the asymmetric Island and Dirichlet models may be due to the additional genetic information available to the asymmetric Island model.
-To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that that each source was equally likely.
-Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=7$ and $n=6$ respectively) while only being observed once each among the source isolates, so that the Dirichlet model has little information available to distinguish between sources.
-The asymmetric Island model, however, can exploit the genetic relationship between genotypes.
-ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables \ref{tab:mlst} and \ref{tab:counts}).
-Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources.
-
-As noted previously, a major public health initiative in 2007 led to a significant reduction in the number of cases of campylobacteriosis in New Zealand.
-In order to examine the effects of this change on attribution probabilities, we repeated the analysis using only data from 2008 to 2014.
-Figure~\ref{fig3} shows that the general trend in attribution by rurality is consistent with that seen with all the data, with ruminant associated campylobacteriosis in highly rural areas to poultry associated cases in urban centers linked by a linear trend.
-However, the proportion attributed to poultry across all rurality levels decreased (and thus attribution to ruminants increased) in this post-intervention dataset, being most marked in urban centers.
-Thus, although the intervention did not eliminate infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.
-
-Lastly, we use deviance information criterion (DIC) for model comparison.
-DIC values obtained from our MCMC runs are displayed in Table~\ref{tab3}.
-Overall there is a clear signal that a linear representation of rurality (on the logit scale) is adequate.
-Note that the asymmetric Island and Dirichlet models are not directly comparable by DIC as the likelihoods are on different scales: the Dirichlet model assumes all potential sequence types have been observed so that $\sum_j p(\boldsymbol{\pi}_j) = 1$, whereas the asymmetric Island model allows for unobserved sequence types so that $\sum_j p(\boldsymbol{\pi}_j) < 1$.
-
 \begin{table}
@@
   \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 (whole period) and to the data from 2008 to 2014 (post-intervention period) given the genotypic information is from the asymmetric Island model or the Dirichlet model.}
-  \label{tab3}
+  \label{tab:dic}
 \end{table}
 
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{Figures/genotype_figure.pdf}
+\caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
+\label{fig:genotype_dist}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{Figures/attr_intervention.pdf}
+\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area.
+The attribution is generated using the linear model given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
+\label{fig:intervention}
+\end{figure}
+
+Posterior attribution of human cases of campylobacteriosis ($F$) with 80\% credible intervals for each source are illustrated for each rurality grade in Figure~\ref{fig2}.
+The graphs are categorized by the type of genetic model (asymmetric Island or Dirichlet) and the manner in which the rurality variable is modelled (categorical or linear on the logit scale).
+
+Overall, the attribution results are relatively stable irrespective of the genetic model or how rurality is represented in the attribution model.
+The majority of human cases are attributed to ruminants and poultry, with more cases attributed to ruminants in rural areas and more cases attributed to poultry in urban areas.
+For the Dirichlet and asymmetric Island models, the linear and categorical models of rurality show broadly the same trend, suggesting that the additional flexibility given by the categorical model is not required, and that the shift in attribution as rurality changes is adequately modelled by a linear trend on the logit scale.
+The linear model has the advantage of tighter credible intervals as it can share data across the 7 levels of rurality, resulting in a clearer separation of ruminant and poultry attribution, particularly in highly rural areas where the data are sparse.
+
+There are some small differences between the genotype models, with the Dirichlet model showing a greater attribution to poultry (ranging from 40\% in highly rural areas to 70\% in main urban centres) than the asymmetric Island model (ranging from 30\% in rural areas to 65\% in urban centres).
+This occurs in both the linear and categorical models.
+Interestingly, the asymmetric Island model attributes approximately 7\% of human cases across all rurality levels to sources other than poultry, ruminants and water, and gives a small attribution to water in highly rural areas, whilst the Dirichlet model indicates that both these sources are unimportant.
+
+We use deviance information criterion (DIC) for model comparison.
+DIC values obtained from our MCMC runs are displayed in Table~\ref{tab:dic}.
+Overall there is a clear signal that a linear representation of rurality (on the logit scale) is adequate.
+Note that the asymmetric Island and Dirichlet models are not directly comparable by DIC as the likelihoods are on different scales: the Dirichlet model assumes all potential sequence types have been observed so that $\sum_j p(\boldsymbol{\pi}_j) = 1$, whereas the asymmetric Island model allows for unobserved sequence types so that $\sum_j p(\boldsymbol{\pi}_j) < 1$.
+
+The small differences in attribution observed between the asymmetric Island and Dirichlet models may be due to the additional genetic information available to the asymmetric Island model.
+To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that that each source was equally likely.
+Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=7$ and $n=6$ respectively) while only being observed once each among the source isolates, so that the Dirichlet model has little information available to distinguish between sources.
+The asymmetric Island model, however, can exploit the genetic relationship between genotypes.
+ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables \ref{tab:mlst} and \ref{tab:counts}).
+Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources.
+In contrast, both models provide similar probabilities for ST-2026 and ST474 which are both observed frequently.
+
+As noted previously, a major public health initiative in 2007 led to a significant reduction in the number of cases of campylobacteriosis in New Zealand.
+In order to examine the effects of this change on attribution probabilities, we repeated the analysis by including an interaction with time period (2005--2007 and 2008--2014).
+Figure~\ref{fig:intervention} shows that the general trend in attribution by rurality for each of the time periods utilising a linear trend on the logit scale to incorporate rurality.
+
+TODO: FILL THIS IN...
+
+is consistent with that seen with all the data, with ruminant associated campylobacteriosis in highly rural areas to poultry associated cases in urban centers linked by a linear trend.
+However, the proportion attributed to poultry across all rurality levels decreased (and thus attribution to ruminants increased) in this post-intervention dataset, being most marked in urban centers.
+Thus, although the intervention did not eliminate infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.
+
 \section{Discussion}
-Models that determine the source of human infection, particularly with zoonotic pathogens that originate in animal populations, are of considerable value to public health policy makers.
-However, the complexity of models may increase due to the complex transmission pathways of pathogens, particularly when a model combines epidemiological and molecular data.
+Models that determine the source of human infection, particularly for zoonotic pathogens that originate in animal populations, are of considerable value to public health policy makers.
+However, such models may be complex, particularly when utilising evolutionary models.
 An outstanding question is whether such complexity is required, or whether a simpler model may work as effectively.
@@
 
-Our results showed that the simple Dirichlet model and the asymmetric Island model showed largely similar results for the final attribution, with both models demonstrating a clear effect of rurality on attribution: cases in rural areas are more likely to have originated from ruminants, while those in main urban centers are more likely to be of poultry origin.
+Our results showed that the Dirichlet model and the asymmetric Island model showed largely similar results for final attribution, with both models demonstrating a clear effect of rurality on attribution: cases in rural areas are more likely to have originated from ruminants, while those in main urban centers are more likely to be of poultry origin.
 As most people in the Manawatu region live in urban centers, this highlights the importance of poultry as a reservoir for campylobacteriosis, which is well established in the literature \cite{MullA, Marsh, MullM, Leve}.
 
-When we ran models allowing attribution probabilities to differ before and after the intervention in the poultry industry in 2007, we saw a clear difference, with much lower poultry (and higher ruminant) attribution, particularly in main urban centers.
+When we ran models allowing attribution probabilities to differ before and after the intervention in the poultry industry in 2007, we saw a clear difference, with much lower poultry (and higher ruminant) attribution, particularly in main urban centers in 2008--2014.
 Considering that most people in the Manawatu region live in urban areas, and that case rates in urban areas decreased from 2008 onwards, it is clear that this intervention coincided with a dramatic reduction in poultry attributed illness as reported elsewhere \cite{AnnS}.
@@
 While the overall attribution was consistent between the Dirichlet and asymmetric Island models, it would be expected that the conditional probabilities for a given genotype might differ markedly.
-For those genotypes observed infrequently (or not at all) among the sources, the Dirichlet model has little information while the asymmeteric Island model can exploit information from cases with similar (but not identical) genetic profiles.
+For those genotypes observed infrequently (or not at all) among the sources the Dirichlet model has little information while the asymmeteric Island model can exploit information from cases with similar (but not identical) genetic profiles.
 In the case of MLST data with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often.
-Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution where the bulk of the data lie.
+Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution well where the bulk of the data lie.
 It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute significantly to the overall attribution.
@@
 We originally used standard normal priors for regression coefficients on the logit scale.
-We also considered priors with $\sigma^2=4$, and while this meant that $f$ tended to drift further from 0, the resulting attribution probabilities $F$ did not, largely as the attribution is dominated by poultry and ruminant sources, with the water source in particular being close to zero.
+We also considered priors with $\sigma^2=4$, and while this meant that $f$ tended to drift further from 0, the resulting attribution probabilities $F$ did not change, largely as the attribution is dominated by poultry and ruminant sources, with the water source in particular being close to zero.
 Thus, $f_\mathsf{poultry}$ and $f_\mathsf{ruminants}$ are positive, while $f_\mathsf{water}$ is negative and the magnitude of these can increase without making all that much difference to their corresponding $F$'s.
@@
 The prior $\mathbf{\alpha}_j$ in the Dirichlet also makes little difference if kept small, as it most strongly effects genotypes that are rare, which do not contribute significantly to the overall attribution.
-It can be thought of as augmenting with new data, so that $\mathbf{\alpha}_j=\mathbf{1}$ is equivalent to adding a single observation of each genotype to source $j$.
+The prior can be thought of as data augmentation such that $\mathbf{\alpha}_j=\mathbf{1}$ is equivalent to adding a single observation of each genotype to source $j$.
 Thus, large values of $\mathbf{\alpha}_j$ will cause the genotype distributions across sources to look more similar, and hence result in equal attribution to each source.
@@
 Future research might focus on possible extensions to these models.
-One direction is to adapt the models further with additional covariates, which might include age, occupation, and other risk factors such as contact with animals.
+One direction is to adapt the models with additional covariates, which might include age, occupation, and other risk factors such as contact with animals.
 For example, there is evidence that children in rural areas are at higher risk of campylobacteriosis through contact with farm animals \cite{MullM, Spencer2012}.
-Another direction is in expanding the role of water in particular.
+Another direction is in expanding the role of water.
 In these models, we've assumed that water is a source of human campylobacteriosis infection, but water differs from the other food and environmental sources in that it is not an amplifying reservoir for \emph{Campylobacter} \cite{Wagen}.
-In contrast, the genotypes found in water might be expected to originate in the other sources present here, particularly ruminants and wild birds, but also potentially from humans as well via discharge of unprocessed human waste.
+In contrast, genotypes found in water might be expected to originate in the other sources present here, particularly ruminants and wild birds, but also potentially from humans as well via discharge of unprocessed human waste.
 Hence, water acts as a transmission pathway from sources to humans, being both an end point (reduced water quality from faecal contamination) and a source (human consumption of water, either recreationally or through untreated water supplies).
@@
 However, water is known as a key source of outbreaks of campylobacteriosis, such as the large outbreak in Havelock North, New Zealand in 2016 where an estimated 5,500 out of 14,000 residents became ill \cite{dia_campy_report1}.
-Thus, characterising both characterising the source of \emph{Campylobacter} found in water has important implications for both water quality and public health.
+Thus, characterising the source of \emph{Campylobacter} found in water has important implications for both water quality and public health.
 
 \section*{Acknowledgements}
+
+We thank Mid Central Public Health Services, and the Molecular Epidemiology and Public Health Laboratory (mEpiLab, Massey University) for data collection, and the Ministry of Primary Industries for funding the Manawatu campylobacteriosis sentinel surveillance site.
 

### 2018-03-20 paper/finalish/article_1023.tex
updates to include attribution pre/post intervention

@@
 Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases.
-As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants\cite{Muell, Coll, Cart}.
+As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants \cite{Muell, Coll, Cart}.
 The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.
@@
 Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random.
-Table~\ref{tab2} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census\cite{Stat1, Stat2}.
+Table~\ref{tab2} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census \cite{Stat1, Stat2}.
 The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
@@
 \centering
-\includegraphics[width=.8\linewidth]{Figures/attr_all.pdf}
+\includegraphics[width=.9\linewidth]{Figures/attr_all.pdf}
 \caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas.
@@
 \centering
-\includegraphics[width=.8\linewidth]{Figures/genotype_figure.pdf}
+\includegraphics[width=.9\linewidth]{Figures/genotype_figure.pdf}
 \caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
@@
 \centering
-\includegraphics[width=.8\linewidth]{Figures/attr_intervention.pdf}
+\includegraphics[width=.9\linewidth]{Figures/attr_intervention.pdf}
 \caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area.
@@
 Figure~\ref{fig:intervention} shows that the general trend in attribution by rurality for each of the time periods utilising a linear trend on the logit scale to incorporate rurality.
-
-TODO: FILL THIS IN...
-
-is consistent with that seen with all the data, with ruminant associated campylobacteriosis in highly rural areas to poultry associated cases in urban centers linked by a linear trend.
-However, the proportion attributed to poultry across all rurality levels decreased (and thus attribution to ruminants increased) in this post-intervention dataset, being most marked in urban centers.
+There is a clear difference, with a significantly lower attribution to poultry (and correspondingly higher attribution to ruminants) in all but the most rural of areas, being strongest in highly urban areas.
 Thus, although the intervention did not eliminate infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.

### 2018-03-20 paper/finalish/article_1023.tex
update figure captions, and fix missing dash in ST number

@@
 \includegraphics[width=.9\linewidth]{Figures/genotype_figure.pdf}
-\caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
+\caption{Posterior probability for each source for four sequence types from the asymmetric Island (red) and Dirichlet (blue) models, assuming that each source is \emph{apriori} equally likely.}
 \label{fig:genotype_dist}
@@
 \includegraphics[width=.9\linewidth]{Figures/attr_intervention.pdf}
-\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area.
+\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for poultry (red), ruminants (blue), water (green), and other sources (purple) over the rurality scales from highly rural area to main urban area (Table~\ref{tab2}).
 The attribution is generated using the linear model given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
@@
 Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources.
-In contrast, both models provide similar probabilities for ST-2026 and ST474 which are both observed frequently.
+In contrast, both models provide similar probabilities for ST-2026 and ST-474 which are both observed frequently.
 

### 2018-03-20 paper/attr_figures.R
update code for figures

@@ -67,10 +67,10 @@ ggplot(plot_df) +
   geom_ribbon(aes(x=UR_num, ymin=li, ymax=ui, fill=Source), alpha=0.3) +
   geom_line(aes(x=UR_num, y=m, col=Source), lwd=1) +
   facet_grid(GenotypeModel~AttributionModel) +
-  scale_x_continuous(name=NULL, breaks=c(-3,3), labels=c("Highly rural", "Highly Urban"), expand=c(0,0)) +
+  scale_x_continuous(name=NULL, breaks=c(-3,3), labels=c("Rural", "Urban"), expand=c(0,0)) +
   scale_y_continuous(name="Percentage of cases", labels=scales::percent_format(), limits=c(0,1), expand=c(0,0)) +
-  scale_color_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
-  scale_fill_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
+  scale_color_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue2", Other="plum4", Water="green4")) +
+  scale_fill_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue2", Other="plum4", Water="green4")) +
   theme_bw() +
   theme(text = element_text(family="Times"),
         legend.position = c(0.99,0.89),

### 2018-03-20 paper/attr_figures_intervention.R

@@ -65,14 +65,15 @@ plot_df <- df %>%
 library(ggplot2)
 
 cairo_pdf("attr_intervention.pdf", width=7, height=5)
+#pdf("attr_intervention.pdf", width=7, height=5, colormodel = 'gray')
 ggplot(plot_df) +
   geom_ribbon(aes(x=UR_num, ymin=li, ymax=ui, fill=Source), alpha=0.3) +
   geom_line(aes(x=UR_num, y=m, col=Source), lwd=1) +
   facet_grid(GenotypeModel~Intervention) +
-  scale_x_continuous(name=NULL, breaks=c(-3,3), labels=c("Highly rural", "Highly Urban"), expand=c(0,0)) +
+  scale_x_continuous(name=NULL, breaks=c(-3,3), labels=c("Rural", "Urban"), expand=c(0,0)) +
   scale_y_continuous(name="Percentage of cases", labels=scales::percent_format(), limits=c(0,1), expand=c(0,0)) +
-  scale_color_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
-  scale_fill_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue", Other="plum4", Water="green4")) +
+  scale_color_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue2", Other="plum4", Water="green4")) +
+  scale_fill_manual(name=NULL, values = c(Poultry="brown", Ruminants="steelblue2", Other="plum4", Water="green4")) +
   theme_bw() +
   theme(text = element_text(family="Times"),
         legend.position = c(0.99,0.89),

### 2018-03-20 paper/genotype_figure.R

@@ -1,0 +1,67 @@
+library(islandR)
+library(dplyr)
+library(tidyr)
+library(ggjoy)
+
+# Fit the sequence type distribution using the island model
+st_i = replicate(10, {
+  s=st_fit(formula = Source ~ ST,
+            non_primary = "Human",
+            data = manawatu,
+            method="island",
+            sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
+            samples = 1000)
+  lapply(1:1000, function(i) { data.frame(ST=s$types, s$sampling_distribution[,,i], Iteration=i) })
+  }
+  )
+
+# Reformat to a data.frame
+st_il <- apply(st_i, 2, function(x) { do.call(rbind, x) })
+st_il2 <- lapply(seq_along(st_il), function(x) { cbind(st_il[[x]], Chain=x)})
+st_id <- do.call(rbind, st_il2) %>% mutate(Iteration = Iteration + (Chain-1)*1000) %>%
+  select(-Chain)
+
+# and using the Dirichlet
+st_d = st_fit(formula = Source ~ ST,
+            non_primary = "Human",
+            data = manawatu,
+            method="dirichlet",
+            iters=10000)
+
+st_dl <- lapply(1:10000, function(i) { data.frame(ST=as.numeric(st_d$types), st_d$sampling_distribution[,,i], Iteration=i) })
+st_dd <- do.call(rbind, st_dl)
+
+# Join them up
+st <- bind_rows(cbind(st_id, Model='Island'), cbind(st_dd, Model='Dirichlet')) %>%
+  gather(Source, P, -ST, -Iteration, -Model)
+
+final <- st %>% group_by(Iteration, Model, ST) %>%
+  mutate(P = P/sum(P)) %>% spread(Model, P)
+
+save(list='final', file='model_compare.Rdata')
+load('model_compare.Rdata')
+
+sts <- c(403, 2343, 2026, 474)
+plot_dat <- final %>% filter(ST %in% sts) %>%
+  gather(Model, P, Dirichlet, Island) %>% ungroup %>% mutate(ST = factor(ST, levels=sts, labels = paste0("ST-", sts)),
+                                                             Scale = ifelse(Model == "Dirichlet" & ST %in% c("ST-403","ST-2343"), 2, 1.2))
+
+cairo_pdf("genotype_figure.pdf", width=7, height=4)
[…]

### 2018-03-20 paper/finalish/article_1023.tex
updates latex document to new template from wiley

@@
-\documentclass[times, doublespace]{simauth}%For paper submission
-%\documentclass[times]{simauth}
-% Packages
-\usepackage{mdframed}
-\usepackage{subcaption}
-\usepackage{subfig}
-\usepackage{url}
-\usepackage{color}
-%\usepackage[T1]{fontenc}
-%\usepackage{times}
-%\usepackage{bm}
-%\usepackage{mathtime}
+\RequirePackage{etex}
+\documentclass[AMA,STIX1COL]{WileyNJD-v2}
 
-% Customised Commands
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{corollary}{Corollary}[theorem]
-\newtheorem{lemma}[theorem]{Lemma}
-\newenvironment{proof}[1][Proof]{\begin{trivlist}
-\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-\newcommand{\qed}{\nobreak \ifvmode \relax \else
-      \ifdim\lastskip<1.5em \hskip-\lastskip
-      \hskip1.5em plus0em minus0.5em \fi \nobreak
-      \vrule height0.75em width0.5em depth0.25em\fi}
+\articletype{Research Article}%
 
-% Bibliographystyle
-\bibliographystyle{wileyj}
+\received{26 April 2016}
+\revised{6 June 2016}
+\accepted{6 June 2016}
+
+\raggedbottom
 
 \begin{document}
-\runninghead{Extending statistical models for the source attribution of zoonotic diseases}
+
 \title{Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis}
-\author{Sih-Jing Liao\affil{a},
-Jonathan Marshall\affil{a}\corrauth\footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
-\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
-\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
-\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}
 
-\begin{abstract}
+\author[1]{Sih-Jing Liao}
 
-Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways.
+\author[1]{Jonathan Marshall*}
+
+\author[1]{Martin L. Hazelton}
+
+\author[2]{Nigel P. French}
+
+\authormark{Sih-Jing Liao \textsc{et al}}
+
+\address[1]{\orgdiv{Institute of Fundamental Sciences},\orgname{Massey University}, \orgaddress{\state{Palmerston North}, \country{New Zealand}}}
+
+\address[2]{\orgdiv{mEpiLab, Hopkirk Research Institute}, \orgname{Massey University}, \orgaddress{\state{Palmerston North}, \country{New Zealand}}}
+
+\address[3]{\orgdiv{Org Division}, \orgname{Org Name}, \orgaddress{\state{State name}, \country{Country name}}}
+
+\corres{*J Marshall, Institute of Fundamental Sciences, Massey University, Palmerston North, New Zealand. \email{J.C.marshall@massey.ac.nz}}
+
+%\presentaddress{This is sample for present address text this is sample for present address text}
+
+\abstract[Summary]{Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways.
 Modeling jointly the epidemiological data and genetic information of microbial isolates derived from cases provides a methodology for tracing back the source of infection.
@@
 A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic-free model.
-We show that inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.
-
-\end{abstract}
+We show that inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.}
 
 \keywords{source attribution; \textit{Campylobacter}; genetic model; Dirichlet prior; HPD interval; DIC}
+
+\jnlcitation{\cname{%
+\author{S.J. Liao}, 
+\author{J. Marshall}, 
+\author{M.L. Hazelton}, and 
+\author{N.P. French}
+\author{T. Woollings}} (\cyear{2018}), 
+\ctitle{Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis}, \cjournal{Statistics In Medicine}.} %, \cvol{2018;00:1--6}.}
+
 \maketitle
+
+%\footnotetext{\textbf{Abbreviations:} ANA, anti-nuclear antibodies; APC, antigen-presenting cells; IRF, interferon regulatory factor}
 

### 2018-03-20 paper/finalish/article_1023.tex
tidy up table/figure references

@@
 For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the commonly used 7-gene MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
-An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab:mlst}).
+An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~\ref{tab:mlst}.
 It shows that the genotypes ST-2026 and ST-474 have different allelic combinations across all seven loci, while ST-403 differs from ST-2026 only at \textit{gln}A, and ST-2343 differs from ST-474 at \textit{glt}A and \textit{pgm}.
@@
 Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random.
-Table~\ref{tab2} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census \cite{Stat1, Stat2}.
+Table~\ref{tab:rurality} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census \cite{Stat1, Stat2}.
 The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008.
@@
       \caption{The number of human cases in each rurality class from 2005--2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
-      \label{tab2}
+      \label{tab:rurality}
 \end{table}
@@
 \includegraphics[width=.9\linewidth]{Figures/attr_intervention.pdf}
-\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for poultry (red), ruminants (blue), water (green), and other sources (purple) over the rurality scales from highly rural area to main urban area (Table~\ref{tab2}).
+\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for poultry (red), ruminants (blue), water (green), and other sources (purple) over the rurality scales from highly rural area to main urban area (Table~\ref{tab:rurality}).
 The attribution is generated using the linear model given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
@@
 The asymmetric Island model, however, can exploit the genetic relationship between genotypes.
-ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables \ref{tab:mlst} and \ref{tab:counts}).
+ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables~\ref{tab:mlst} and \ref{tab:counts}).
 Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources.

### 2018-03-20 R/datasets.R
update the manawatu dataset to contain year and urban/rural categories

@@ -14,8 +14,8 @@
 #'   \item{TKT}{The TKT housekeeping gene}
 #'   \item{UNC}{The UNC housekeeping gene}
 #'   \item{Source}{The source (Human, Poultry, Ruminants, Water, Other)}
-#'   \item{Location}{A factor containing Urban or Rural for each Human case}
-#'   \item{Intervention}{A factor containing before or after for each Human case, specifying whether
-#'   it occurred before or after the 2007-2008 intervention in the Poultry industry}
+#'   \item{Year}{The year the sample was collected}
+#'   \item{UR2006_num}{A numeric code containing the level of Rurality for each Human case}
+#'   \item{UR2006_name}{The corresponding name of the numeric identifier `UR2006_num` for each Human case}
 #' }
 "manawatu"

### 2018-03-20 paper/attr_figures.R
updates attr_figures to use manawatu dataset from islandR

@@ -1,48 +1,43 @@
 library(islandR)
 library(dplyr)
 library(tidyr)
-library(lubridate)
-library(forcats)
+library(ggplot2)
 
-manawatu <- read.csv("~/data/sa_report/data/extract_attribution.csv") %>%
-  mutate(SampledDate = as.Date(SampledDate), Year = year(SampledDate)) %>%
-  filter(Year >= 2005, Year <= 2014) %>%
-  mutate(Source4 = fct_collapse(Source, Poultry = c("Supplier A", "Supplier B", "Supplier Other", "Otherpoultry"),
-                                Ruminants = c("Cattle", "Sheep"),
-                                Water = "Environmental water",
-                                Other = c("Cat_dog_pet", "Duck_poultry", "Spent_hen", "Pig",
-                                          "Turkey", "Water_bird_wild", "Wild_bird_other"))) %>%
-  filter(Source4 != "Human" | !is.na(UR_num)) %>%
-  mutate(UR_fact = factor(UR_num))
+# filter out those where we don't have a location, and have a factor version
+# of UR2006_num
+dataset <- manawatu %>% filter(Source != "Human" | !is.na(UR2006_num)) %>%
+  mutate(UR2006_fact = factor(UR2006_num))
+
+num_samples = 1
 
 # Fit the sequence type distribution using the island model
-st_i = st_fit(formula = Source4 ~ ST,
+st_i = st_fit(formula = Source ~ ST,
          non_primary = "Human",
-         data = manawatu,
-         method="island",
+         data = dataset,
+         method = "island",
          sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
-         samples = 1000)
+         samples = num_samples)
 
 # and using the Dirichlet
-st_d = st_fit(formula = Source4 ~ ST,
+st_d = st_fit(formula = Source ~ ST,
               non_primary = "Human",
-              data = manawatu,
-              method="dirichlet",
-              iters=1000)
+              data = dataset,
+              method = "dirichlet",
+              iters = num_samples)
 
 # do the attribution
[…]

### 2018-03-20 data-raw/process_data.R
reorder the UR2006_name factor correctly in the manawatu dataset

@@ -4,5 +4,10 @@ library(devtools)
 base_path <- "data-raw"
 manawatu <- read.csv(file.path(base_path, "Manawatu.csv"))
 
+# reorder the UR2006_name factor
+library(dplyr)
+order <- manawatu %>% select(UR2006_num, UR2006_name) %>% unique %>% arrange(UR2006_num) %>% pull(UR2006_name)
+manawatu$UR2006_name = factor(manawatu$UR2006_name, levels=order)
+
 # save data within package
 devtools::use_data(manawatu, overwrite=TRUE)

### 2018-03-20 paper/attr_figures.R
updates attr_figures some more

@@ -8,7 +8,7 @@ library(ggplot2)
 dataset <- manawatu %>% filter(Source != "Human" | !is.na(UR2006_num)) %>%
   mutate(UR2006_fact = factor(UR2006_num))
 
-num_samples = 1
+num_samples = 10
 
 # Fit the sequence type distribution using the island model
 st_i = st_fit(formula = Source ~ ST,
@@ -23,20 +23,21 @@ st_d = st_fit(formula = Source ~ ST,
               non_primary = "Human",
               data = dataset,
               method = "dirichlet",
-              iters = num_samples)
+              samples = num_samples)
 
 # do the attribution
-at_if <- attribution(ST ~ UR2006_fact, st_i, manawatu %>% filter(Source == "Human"))
-at_il <- attribution(ST ~ UR2006_num, st_i, manawatu %>% filter(Source == "Human"))
-at_df <- attribution(ST ~ UR2006_fact, st_d, manawatu %>% filter(Source == "Human"))
-at_dl <- attribution(ST ~ UR2006_num, st_d, manawatu %>% filter(Source == "Human"))
+humans <- dataset %>% filter(Source == "Human")
+at_if <- attribution(ST ~ UR2006_fact, st_i, humans)
+at_il <- attribution(ST ~ UR2006_num, st_i, humans)
+at_df <- attribution(ST ~ UR2006_fact, st_d, humans)
+at_dl <- attribution(ST ~ UR2006_num, st_d, humans)
 
 df_if <- predict(at_if, FUN=identity) %>% extract(X, into="UR2006_num", regex="UR2006_fact([-0-9]+)=1", convert=TRUE) %>%
-  replace_na(replace=list(UR_num=-3)) %>% mutate(GenotypeModel = 'Island', AttributionModel='Categorical')
+  replace_na(replace=list(UR2006_num=-3)) %>% mutate(GenotypeModel = 'Island', AttributionModel='Categorical')
 df_il <- predict(at_il, FUN=identity) %>% extract(X, into="UR2006_num", regex="UR2006_num=([-0-9]+)", convert=TRUE) %>%
   mutate(GenotypeModel = 'Island', AttributionModel='Linear')
 df_df <- predict(at_df, FUN=identity) %>% extract(X, into="UR2006_num", regex="UR2006_fact([-0-9]+)=1", convert=TRUE) %>%
[…]

### 2018-03-20 R/sample_dist_dirichlet.R
update st_fit_dirichlet so it is consistent with st_fit_island

@@ -20,11 +20,11 @@ rdirichlet<-function(n,a)
 #' @param prior   The Dirichlet prior. Currently only a constant. Defaults to 1.
 #' @param non_primary one or more sources that should be considered 'output' sources. No genotype distribution is computed for these,
 #'        but P(ST | source) will be computed for these STs in addition to those observed on other sources.
-#' @param iters the number of iterations to sample.
+#' @param samples the number of iterations to sample. Defaults to 100.
 #' @param data optional data frame from which to take variables in \code{formula} and \code{sequence}.
 #' @return an object of class dirichlet which derives from sampling_dist.
 #' @seealso \code{\link{st_fit}}, \code{\link{print.sample_dist}}, \code{\link{plot.sample_dist}}, \code{\link{summary.sample_dist}}
-st_fit_dirichlet <- function(formula, prior = 1, non_primary = "Human", iters = 100, data) {
+st_fit_dirichlet <- function(formula, prior = 1, non_primary = "Human", samples = 100, data) {
   mod.terms = terms(formula, data=data)
   mod.frame = model.frame(formula, data=data)
 
@@ -59,7 +59,7 @@ st_fit_dirichlet <- function(formula, prior = 1, non_primary = "Human", iters = 
     a
   }
 
-  out <- replicate(iters, dirichlet_matrix(alpha))
+  out <- replicate(samples, dirichlet_matrix(alpha))
   # righto, now construct a useful object...
   x = list(types = rownames(counts), sources = source_names,
            sampling_distribution = out, model = "dirichlet")

### 2018-03-21 paper/finalish/article.tex
use sensible identifiers for the figures

--- 2018-03-20 paper/finalish/article_1023.tex
@@
 The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
-\label{fig2}
+\label{fig:attribution}
 \end{figure}
@@
 
-Posterior attribution of human cases of campylobacteriosis ($F$) with 80\% credible intervals for each source are illustrated for each rurality grade in Figure~\ref{fig2}.
+Posterior attribution of human cases of campylobacteriosis ($F$) with 80\% credible intervals for each source are illustrated for each rurality grade in Figure~\ref{fig:attribution}.
 The graphs are categorized by the type of genetic model (asymmetric Island or Dirichlet) and the manner in which the rurality variable is modelled (categorical or linear on the logit scale).

### 2018-03-21 paper/finalish/article.tex
move the attribution figure to fig_attribution.pdf
fixes from Jing

@@
 For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the commonly used 7-gene MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit).
-An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~\ref{tab:mlst}.
+An illustrative example of MLST data for \emph{Campylobacter} is presented in Table~\ref{tab:mlst}.
 It shows that the genotypes ST-2026 and ST-474 have different allelic combinations across all seven loci, while ST-403 differs from ST-2026 only at \textit{gln}A, and ST-2343 differs from ST-474 at \textit{glt}A and \textit{pgm}.
@@
 The goal of attribution models is to estimate the probability that the observed human cases arise from each putative source.
-Our data comprise microbial genotype information from each observed human case obtained from analysis of stool samples, and also from a pool of non-human cases corresponding to potential zoonotic sources of disease
+Our data comprise microbial genotype information from each observed human case obtained from analysis of stool samples, and also from a pool of non-human cases corresponding to potential zoonotic sources of disease.
 
@@
 To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$.
-Therefore, $p(\text{ST}_h \vert \text{source }j)$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.
+Therefore, $p(\text{ST}_h \vert \text{source }j)$ can be obtained by simulating from the Dirichlet posterior.
 
-Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models.
+Lastly, to estimate the attribution probability, 500 samples of $p(\text{ST}_h \vert \text{source }j)$ are generated using the asymmetric Island or simple models.
 For each posterior sample, we infer $p(\text{source }j)$ using the likelihood~\eqref{Olikh}.
@@
 \begin{align*}
-\alpha=\frac{L\Big(F^*; \text{ST}\Big)}{L\Big(F; \text{ST}, \boldsymbol{Z}\Big)}\frac{Q\Big(\theta_{(t)} \vert \theta_{(t)}^*\Big)}{Q\Big(\theta_{(t)}^* \vert \theta_{(t)}\Big)}\frac{p\Big(\theta_{(t)}^*\Big)}{p\Big(\theta_{(t)}\Big)},
+\alpha=\frac{L\Big(F^*; \text{ST}\Big)}{L\Big(F; \text{ST}\Big)}\frac{Q\Big(\theta_{(t)} \vert \theta_{(t)}^*\Big)}{Q\Big(\theta_{(t)}^* \vert \theta_{(t)}\Big)}\frac{p\Big(\theta_{(t)}^*\Big)}{p\Big(\theta_{(t)}\Big)},
 \end{align*}
-where the likelihood $L\Big(F^*; \text{ST})$ is given by equation \eqref{Olikh}.
+where the likelihood $L\Big(F^*; \text{ST}\Big)$ is given by equation \eqref{Olikh}.
 \item Sample $u$ from $U(0,1)$.
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/attr_all.pdf}
+\includegraphics[width=.9\linewidth]{Figures/fig_attribution.pdf}
 \caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas.
@@
 
-There are some small differences between the genotype models, with the Dirichlet model showing a greater attribution to poultry (ranging from 40\% in highly rural areas to 70\% in main urban centres) than the asymmetric Island model (ranging from 30\% in rural areas to 65\% in urban centres).
+There are some small differences between the genotype models, with the Dirichlet model showing a greater attribution to poultry (ranging from 40\% in highly rural areas to 75\% in main urban centres) than the asymmetric Island model (ranging from 30\% in rural areas to 65\% in urban centres).
 This occurs in both the linear and categorical models.
@@
 The small differences in attribution observed between the asymmetric Island and Dirichlet models may be due to the additional genetic information available to the asymmetric Island model.
-To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that that each source was equally likely.
+To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that each source was equally likely.
 Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=7$ and $n=6$ respectively) while only being observed once each among the source isolates, so that the Dirichlet model has little information available to distinguish between sources.
@@
 The prior on $f$ thus tends to restrict this ill-behaviour rather than acting as a strong constraint on attribution probabilities.
-The prior $\mathbf{\alpha}_j$ in the Dirichlet also makes little difference if kept small, as it most strongly effects genotypes that are rare, which do not contribute significantly to the overall attribution.
-The prior can be thought of as data augmentation such that $\mathbf{\alpha}_j=\mathbf{1}$ is equivalent to adding a single observation of each genotype to source $j$.
-Thus, large values of $\mathbf{\alpha}_j$ will cause the genotype distributions across sources to look more similar, and hence result in equal attribution to each source.
+The prior $\boldsymbol{\alpha}_j$ in the Dirichlet also makes little difference if kept small, as it most strongly effects genotypes that are rare, which do not contribute significantly to the overall attribution.
+The prior can be thought of as data augmentation such that $\boldsymbol{\alpha}_j=\boldsymbol{1}$ is equivalent to adding a single observation of each genotype to source $j$.
+Thus, large values of $\boldsymbol{\alpha}_j$ will cause the genotype distributions across sources to look more similar, and hence result in equal attribution to each source.
 

### 2018-03-21 paper/finalish/article.tex
switch Jing to corresponding author
Add link to github islandR package

@@
 
-\author[1]{Sih-Jing Liao}
+\author[1]{Sih-Jing Liao*}
 
-\author[1]{Jonathan Marshall*}
+\author[1]{Jonathan Marshall}
 
@@
 
-\corres{*J Marshall, Institute of Fundamental Sciences, Massey University, Palmerston North, New Zealand. \email{J.C.marshall@massey.ac.nz}}
+\corres{*Sih-Jing Liao, Institute of Fundamental Sciences, Massey University, Palmerston North, New Zealand. \email{S.J.Liao@massey.ac.nz}}
 
@@
 
+All code for fitting the genotype distributions, estimating attribution probabilities and producing figures, along with the data are available in the R package `islandR` from \url{https://github.com/jmarshallnz/islandR}.
+
 \section{Results}

### 2018-03-21 paper/attr_figures_intervention.R
update R code for intervention figure using islandR

@@ -1,85 +1,0 @@
-library(islandR)
-library(dplyr)
-library(tidyr)
-library(lubridate)
-library(forcats)
-
-manawatu <- read.csv("~/data/sa_report/data/extract_attribution.csv") %>%
-  mutate(SampledDate = as.Date(SampledDate), Year = year(SampledDate)) %>%
-  filter(Year >= 2005, Year <= 2014) %>%
-  mutate(Source4 = fct_collapse(Source, Poultry = c("Supplier A", "Supplier B", "Supplier Other", "Otherpoultry"),
-                                Ruminants = c("Cattle", "Sheep"),
-                                Water = "Environmental water",
-                                Other = c("Cat_dog_pet", "Duck_poultry", "Spent_hen", "Pig",
-                                          "Turkey", "Water_bird_wild", "Wild_bird_other"))) %>%
-  filter(Source4 != "Human" | !is.na(UR_num) ) %>%
-  mutate(UR_fact = factor(UR_num), Intervention=factor(ifelse(Year < 2008, 'Before', 'After')))
-
-num_samples <- 1000
-
-# Fit the sequence type distribution using the island model
-st_i = st_fit(formula = Source4 ~ ST,
-         non_primary = "Human",
-         data = manawatu,
-         method="island",
-         sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
-         samples = num_samples)
-
-# and using the Dirichlet
-st_d = st_fit(formula = Source4 ~ ST,
-              non_primary = "Human",
-              data = manawatu,
-              method="dirichlet",
-              iters=num_samples)
-
-# do the attribution
-at_if <- attribution(ST ~ UR_num*Intervention, st_i, manawatu %>% filter(Source4 == "Human"))
-at_id <- attribution(ST ~ UR_num*Intervention, st_d, manawatu %>% filter(Source4 == "Human"))
-
-df_il <- predict(at_if, FUN=identity) %>% extract(X, into="UR_num", regex="UR_num=([-0-9]+)", convert=TRUE, remove=FALSE) %>%
-  extract(X, into="Intervention", regex="Intervention([A-Za-z]+)=1") %>%
-  replace_na(replace=list(Intervention="After")) %>%
-  mutate(GenotypeModel = 'Island', AttributionModel='Linear')
[…]

### 2018-03-21 paper/fig_intervention.R

@@ -1,0 +1,73 @@
+library(islandR)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+# filter out those where we don't have a location
+dataset <- manawatu %>% filter(Source != "Human" | !is.na(UR2006_num)) %>%
+  mutate(Intervention = ifelse(Year < 2008, "Before", "After"))
+
+num_samples = 10
+
+# Fit the sequence type distribution using the island model
+st_i = st_fit(formula = Source ~ ST,
+              non_primary = "Human",
+              data = dataset,
+              method = "island",
+              sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
+              samples = num_samples)
+
+# and using the Dirichlet
+st_d = st_fit(formula = Source ~ ST,
+              non_primary = "Human",
+              data = dataset,
+              method = "dirichlet",
+              samples = num_samples)
+
+# do the attribution
+humans <- dataset %>% filter(Source == "Human")
+at_if <- attribution(ST ~ UR2006_num*Intervention, st_i, humans)
+at_id <- attribution(ST ~ UR2006_num*Intervention, st_d, humans)
+
+df_il <- predict(at_if, FUN=identity) %>% extract(X, into="UR2006_num", regex="UR2006_num=([-0-9]+)", convert=TRUE, remove=FALSE) %>%
+  extract(X, into="Intervention", regex="Intervention([A-Za-z]+)=1") %>%
+  replace_na(replace=list(Intervention="After")) %>%
+  mutate(GenotypeModel = 'Island', AttributionModel='Linear')
+df_dl <- predict(at_id, FUN=identity) %>% extract(X, into="UR2006_num", regex="UR2006_num=([-0-9]+)", convert=TRUE, remove=FALSE) %>%
+  extract(X, into="Intervention", regex="Intervention([A-Za-z]+)=1") %>%
+  replace_na(replace=list(Intervention="After")) %>%
+  mutate(GenotypeModel = 'Dirichlet', AttributionModel='Linear')
+
+df <- bind_rows(df_il, df_dl)
+
+save(list='df', file='attribution_fits_intervention.Rdata')
+load('attribution_fits_intervention.Rdata')
+
+plot_df <- df %>%
+  group_by(GenotypeModel, AttributionModel, Source, UR2006_num, Intervention) %>% summarize(
+    m = mean(p),
[…]

### 2018-03-21 paper/finalish/article.tex
move intervention figure to fig_intervention.pdf
update paper + scripts for correct number of iterations for model fitting

@@
 
-Lastly, to estimate the attribution probability, 500 samples of $p(\text{ST}_h \vert \text{source }j)$ are generated using the asymmetric Island or simple models.
+Lastly, to estimate the attribution probability, 100 samples of $p(\text{ST}_h \vert \text{source }j)$ are generated using the asymmetric Island or simple models.
 For each posterior sample, we infer $p(\text{source }j)$ using the likelihood~\eqref{Olikh}.
@@
 For example, $\boldsymbol{\theta}=\{\alpha_1, \alpha_2, \alpha_3, \beta_1, \beta_2, \beta_3\}$ in model~\eqref{modL} or $\boldsymbol{\theta}=\{\beta_{11}, \beta_{12}, \ldots, \beta_{dj}, \ldots, \beta_{73}\}$ in model~\eqref{modC}.
-We use the Metropolis-Hastings algorithm to update $\theta_{(t)}$ and hence $f_{jh}$ and $ F_{jh}$, in a Markov chain with a length of 30,000 iterations.
-The first 5,000 samples are removed as the burn-in period (during which time the chain converges) and the sequence is thinned every 10$^{th}$ sample to reduce computer storage.
+We use the Metropolis-Hastings algorithm to update $\theta_{(t)}$ and hence $f_{jh}$ and $ F_{jh}$, in a Markov chain with a length of 11,000 iterations.
+The first 1,000 samples are removed as the burn-in period (during which time the chain converges) and the sequence is thinned every 100$^{th}$ sample to reduce computer storage.
 Here are the steps in detail:
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/attr_intervention.pdf}
+\includegraphics[width=.9\linewidth]{Figures/fig_intervention.pdf}
 \caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for poultry (red), ruminants (blue), water (green), and other sources (purple) over the rurality scales from highly rural area to main urban area (Table~\ref{tab:rurality}).

### 2018-03-21 paper/fig_attribution.R

@@ -8,7 +8,7 @@ library(ggplot2)
 dataset <- manawatu %>% filter(Source != "Human" | !is.na(UR2006_num)) %>%
   mutate(UR2006_fact = factor(UR2006_num))
 
-num_samples = 10
+num_samples = 100
 
 # Fit the sequence type distribution using the island model
 st_i = st_fit(formula = Source ~ ST,

### 2018-03-21 paper/fig_intervention.R

@@ -7,7 +7,7 @@ library(ggplot2)
 dataset <- manawatu %>% filter(Source != "Human" | !is.na(UR2006_num)) %>%
   mutate(Intervention = ifelse(Year < 2008, "Before", "After"))
 
-num_samples = 10
+num_samples = 100
 
 # Fit the sequence type distribution using the island model
 st_i = st_fit(formula = Source ~ ST,

### 2018-03-21 paper/case_rates_by_year.R
adds R script and update figure for case rates

@@ -1,68 +1,0 @@
-# script to produce barplot figure...
-
-library(meshblocknz)
-library(dplyr)
-library(tidyr)
-library(sf)
-library(lubridate)
-library(forcats)
-
-# 1. Load meshblocknz and compute population by rurality by year
-#    Technically want this only for the meshblocks that we're actually sampling from
-# load phu dataset
-phu <- read_sf("~/code/epiclustR/shape_files/midcentral_phu.shp")
-
-pop <- mb2013 %>% filter(DHB_name == "Mid Central") %>%
-  filter(MB2006 %in% phu$MB06) %>% select(MB2013, MB2006, Pop2001, Pop2006, Pop2013, UR2006_num, UR2013_num) %>%
-  gather(Year, Population, Pop2001:Pop2013) %>% extract(Year, into="Year", regex="([0-9]+)", convert=TRUE)
-
-# OK, now fit a linear model per meshblock to the population
-pops <- split(pop, pop$MB2013)
-
-predict_popn <- function(dat) {
-  m <- lm(Population ~ Year, data=dat)
-  data.frame(Year=2005:2015, Population=predict(m, data.frame(Year=2005:2015)))
-}
-
-pop_interp <- dplyr::bind_rows(lapply(pops, predict_popn), .id='MB2013') %>% mutate(MB2013 = as.numeric(MB2013)) %>%
-  left_join(mb2013 %>% select(MB2006, MB2013, UR2006_num) %>% unique)
-
-# now rurality by year popn...
-
-# 2. Once done, compute number of cases by rurality by year
-manawatu <- read.csv("extract_attribution.csv") %>%
-  mutate(SampledDate = as.Date(SampledDate), Year = year(SampledDate)) %>%
-  filter(Year >= 2005, Year <= 2014) %>%
-  mutate(Source4 = fct_collapse(Source, Poultry = c("Supplier A", "Supplier B", "Supplier Other", "Otherpoultry"),
-                                Ruminants = c("Cattle", "Sheep"),
-                                Water = "Environmental water",
-                                Other = c("Cat_dog_pet", "Duck_poultry", "Spent_hen", "Pig",
-                                          "Turkey", "Water_bird_wild", "Wild_bird_other"))) %>%
-  filter(Source4 == "Human", !is.na(UR_num))
-
[…]

### 2018-03-21 paper/fig_case_rates.R

@@ -1,0 +1,52 @@
+library(meshblocknz)
+library(dplyr)
+library(tidyr)
+library(islandR)
+library(ggplot2)
+
+# Our region is that covered by Mid Central Public Health Unit, which is the intersection
+# of the Mid Central District Health Board and the Manawatu-Wanganui region:
+pop <- mb2013 %>% filter(DHB_name == "Mid Central", RC2013_name == "Manawatu-Wanganui Region") %>%
+  select(MB2013, MB2006, Pop2001, Pop2006, Pop2013, UR2006_num, UR2013_num) %>%
+  gather(Year, Population, Pop2001:Pop2013) %>%
+  extract(Year, into="Year", regex="([0-9]+)", convert=TRUE)
+
+# OK, now fit a linear model per meshblock to the population
+pops <- split(pop, pop$MB2013)
+
+predict_popn <- function(dat) {
+  m <- lm(Population ~ Year, data=dat)
+  data.frame(Year=2005:2014, Population=predict(m, data.frame(Year=2005:2014)))
+}
+
+pop_interp <- dplyr::bind_rows(lapply(pops, predict_popn), .id='MB2013') %>%
+  mutate(MB2013 = as.numeric(MB2013)) %>%
+  left_join(mb2013 %>% select(MB2006, MB2013, UR2006_num) %>% unique)
+
+# now rurality by year popn...
+
+# 2. Once done, compute number of cases by rurality by year
+manawatu
+
+case_rates <- pop_interp %>% group_by(UR2006_num, Year) %>% summarize(Population=sum(Population)) %>%
+  left_join(manawatu %>% group_by(UR2006_num, Year) %>% summarize(Cases = n()))
+
+rates <- case_rates %>% ungroup %>% mutate(Group = ifelse(UR2006_num > 0, "Urban", "Rural")) %>%
+  group_by(Group, Year) %>% summarize(Population=sum(Population, na.rm=TRUE), Cases=sum(Cases, na.rm=TRUE)) %>%
+  mutate(CaseRate = Cases/Population*100000) %>%
+  mutate(CaseRate = ifelse(Year == 2005, CaseRate/0.75, CaseRate)) %>% # In 2005 we have cases from 75% of the year
+  ungroup() %>%
+  mutate(Year = factor(Year))
+
+
+pdf('fig_case_rates.pdf', width=6, height=3.5)
+ggplot(rates, aes(x=Year, y=CaseRate, fill=Group)) +
+  geom_col(position='dodge') +
[…]

### 2018-03-21 paper/fig_genotype_dist.R
adds R script for genotype figure

@@ -4,31 +4,26 @@ library(tidyr)
 library(ggjoy)
 
 # Fit the sequence type distribution using the island model
-st_i = replicate(10, {
-  s=st_fit(formula = Source ~ ST,
-            non_primary = "Human",
-            data = manawatu,
-            method="island",
-            sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
-            samples = 1000)
-  lapply(1:1000, function(i) { data.frame(ST=s$types, s$sampling_distribution[,,i], Iteration=i) })
-  }
-  )
+num_samples <- 10000
 
-# Reformat to a data.frame
-st_il <- apply(st_i, 2, function(x) { do.call(rbind, x) })
-st_il2 <- lapply(seq_along(st_il), function(x) { cbind(st_il[[x]], Chain=x)})
-st_id <- do.call(rbind, st_il2) %>% mutate(Iteration = Iteration + (Chain-1)*1000) %>%
-  select(-Chain)
+st_i = st_fit(formula = Source ~ ST,
+              non_primary = "Human",
+              data = manawatu,
+              method = "island",
+              sequences = ~ ASP + GLN + GLT + GLY + PGM + TKT + UNC,
+              samples=num_samples)
+
+st_il <- lapply(1:num_samples, function(i) { data.frame(ST=as.numeric(st_i$types), st_i$sampling_distribution[,,i], Iteration=i) })
+st_id <- do.call(rbind, st_il)
 
 # and using the Dirichlet
 st_d = st_fit(formula = Source ~ ST,
-            non_primary = "Human",
-            data = manawatu,
-            method="dirichlet",
-            iters=10000)
+              non_primary = "Human",
+              data = manawatu,
+              method = "dirichlet",
+              samples = num_samples)
 
-st_dl <- lapply(1:10000, function(i) { data.frame(ST=as.numeric(st_d$types), st_d$sampling_distribution[,,i], Iteration=i) })
+st_dl <- lapply(1:num_samples, function(i) { data.frame(ST=as.numeric(st_d$types), st_d$sampling_distribution[,,i], Iteration=i) })
 st_dd <- do.call(rbind, st_dl)
 
 # Join them up
@@ -38,15 +33,17 @@ st <- bind_rows(cbind(st_id, Model='Island'), cbind(st_dd, Model='Dirichlet')) %
[…]

### 2018-03-21 paper/finalish/article.tex
move figures to appropriate filenames
use lowercase figures for consistency
adds R script for generating the tables, and update the text accordingly. Just DIC to go then we're done

@@
   \begin{center}
-    \begin{tabular}{crrrrrrr}
+	\begin{tabular}{rrrrrrrr}
 	\toprule
-      ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm}& \textit{tkt}& \textit{unc}A\\ \midrule
+	ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm} & \textit{tkt} & \textit{unc}A\\
+	\midrule
+	2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
+	474 & 2 & 4 & 1 & 2 & 2 & 1 & 5\\
+	2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
 	403 & 10 & 27 & 16 & 19 & 10 & 5 & 7\\
-      2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
-      474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
-      2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
 	\bottomrule
@@
 
-The source attribution study described here uses human case data ($H=1,400$) collected in the Manawatu region of New Zealand from March 2005 to December 2014.
-In addition, over the same time period and from the same geographical location, isolates $N=2,080$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others.
-The total number of unique genotypes from all isolates is $I=355$, with 35\% of genotypes found among human cases.
+The source attribution study described here uses human case data ($H=1,460$) collected in the Manawatu region of New Zealand from March 2005 to December 2014.
+In addition, over the same time period and from the same geographical location, isolates $N=2,128$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others.
+The total number of unique genotypes from all isolates is $I=348$, with 36\% of genotypes found among human cases.
 Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases.
@@
 The data also contains location information, in the form of an ordinal classification of urban and rural areas with seven levels coded from -3 to 3: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area.
-Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random.
+Approximately 8\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random.
 Table~\ref{tab:rurality} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census \cite{Stat1, Stat2}.
@@
 	\toprule
-      ST & Human & Poultry & Ruminants & Water & Others\\ \midrule
-      42 & 54 & 7 & 52 & 10 & 2\\
-      45  & 141 & 154 & 10 & 21 & 54\\
-      474 & 239 & 60 & 15 & 5 & 7\\
-      2026 & 37 & 0 & 46 & 2 & 5\\
-      2381 & 0 & 0 & 0 & 21 & 3\\
+	ST & Human & Poultry & Ruminants & Water & Others\\
+	\midrule
+	42 & 59 & 7 & 53 & 10 & 2\\
+	45 & 149 & 155 & 10 & 21 & 54\\
+	474 & 247 & 60 & 15 & 5 & 9\\
+	2026 & 28 & 0 & 40 & 5 & 2\\
+	2381 & 0 & 0 & 0 & 60 & 3\\
 	\bottomrule
@@
   \centering
-      \begin{tabular}{rlrrrr}
+	\begin{tabular}{rlrrr}
 	\toprule
-        Rurality scale & Description & Human cases & 2006 & 2013 \\ \midrule
-        -3 & Highly rural/remote area & 19 & 1,572 & 1,527\\
-        -2 & Rural area with low urban influence & 92 & 8,373 & 8,304\\
-        -1 & Rural area with moderate urban influence & 127 & 10,392 & 10,734\\
-        0 & Rural area with high urban influence & 70 & 6,579 & 7,155\\
-        1 & Independent urban area & 230 & 28,611 & 28,188\\
-        2 & Satellite urban area & 178 & 19,725 & 20,525\\
-        3 & Main urban area & 579 & 76,047 & 78,108\\
+	Rurality scale & Description & Human cases & 2006 & 2013\\
+	\midrule
+	-3 & Highly rural/remote area & 16 & 1,572 & 1,527\\
+	-2 & Rural area with low urban influence & 103 & 8,382 & 8,316\\
+	-1 & Rural area with moderate urban influence & 124 & 10,392 & 10,734\\
+	0 & Rural area with high urban influence & 78 & 6,579 & 7,155\\
+	1 & Independent urban area & 240 & 28,611 & 28,188\\
+	2 & Satellite urban area & 187 & 19,725 & 20,526\\
+	3 & Main urban area & 596 & 76,047 & 78,108\\
 	\bottomrule
@@
   \centering
-  \includegraphics[width=.8\linewidth]{Figures/case_rates}
+  \includegraphics[width=.8\linewidth]{figures/fig_case_rates.pdf}
   \caption{Case rates per 100,000 population in urban and rural areas of the Manawatu region of New Zealand from 2005 through 2014.
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/fig_attribution.pdf}
+\includegraphics[width=.9\linewidth]{figures/fig_attribution.pdf}
 \caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas.
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/genotype_figure.pdf}
+\includegraphics[width=.9\linewidth]{figures/fig_genotype_dist.pdf}
 \caption{Posterior probability for each source for four sequence types from the asymmetric Island (red) and Dirichlet (blue) models, assuming that each source is \emph{apriori} equally likely.}
@@
 \centering
-\includegraphics[width=.9\linewidth]{Figures/fig_intervention.pdf}
+\includegraphics[width=.9\linewidth]{figures/fig_intervention.pdf}
 \caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for poultry (red), ruminants (blue), water (green), and other sources (purple) over the rurality scales from highly rural area to main urban area (Table~\ref{tab:rurality}).
@@
 To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that each source was equally likely.
-Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=7$ and $n=6$ respectively) while only being observed once each among the source isolates, so that the Dirichlet model has little information available to distinguish between sources.
+Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=6$ cases each) with ST-403 not being observed among the sources, and ST-2343 being observed once in poultry, so that the Dirichlet model has little information available to distinguish between sources.
 The asymmetric Island model, however, can exploit the genetic relationship between genotypes.

### 2018-03-21 paper/fig_case_rates.R

@@ -1,4 +1,4 @@
-library(meshblocknz)
+library(meshblocknz) # from https://github.com/jmarshallnz/meshblocknz
 library(dplyr)
 library(tidyr)
 library(islandR)

### 2018-03-21 paper/tables.R

@@ -1,0 +1,77 @@
+library(islandR)
+library(meshblocknz) # from https://github.com/jmarshallnz/meshblocknz
+library(dplyr)
+library(tidyr)
+library(knitr)
+
+# Table 1: Allelic profiles
+
+st1 <- c(403, 2026, 474, 2343)
+manawatu %>%
+  filter(ST %in% st1) %>%
+  select(ST, ASP:UNC) %>%
+  unique %>%
+  knitr::kable(format='latex', booktabs=TRUE, linesep='', row.names=FALSE)
+
+# Table 2: Frequency of STs
+st2 <- c(42, 45, 474, 2026, 2381)
+manawatu %>%
+  filter(ST %in% st2) %>%
+  group_by(ST, Source) %>%
+  summarize(Count=n()) %>%
+  ungroup %>%
+  spread(Source, Count, fill=0) %>%
+  select(ST, Human, Poultry, Ruminants, Water, Others=Other) %>%
+  knitr::kable(format='latex', booktabs=TRUE, linesep='')
+
+# Table 3: Rurality information
+cases <- manawatu %>%
+  filter(Source == "Human", !is.na(UR2006_num)) %>%
+  group_by(UR2006_num, UR2006_name) %>%
+  summarize(`Human cases` = n())
+
+# Our region is that covered by Mid Central Public Health Unit, which is the intersection
+# of the Mid Central District Health Board and the Manawatu-Wanganui region:
+popn <- mb2013 %>% filter(DHB_name == "Mid Central", RC2013_name == "Manawatu-Wanganui Region") %>%
+  select(UR2006_num, Pop2006, Pop2013) %>%
+  group_by(UR2006_num) %>% summarize_all(.funs=sum)
+
+cases %>% left_join(popn) %>%
+  rename(`Rurality scale`=UR2006_num, Description=UR2006_name,
+         `2006` = Pop2006, `2013` = Pop2013) %>%
+  knitr::kable('latex', booktabs=TRUE, linesep='', format.args=list(big.mark=','))
+
+# Table 4: DIC values
+
+# TODO: Need to grab this out of the (saved) model objects from fig_attribution.R
+
+# Additional information for the text:
+
+# Total number of cases by human/sources
+manawatu %>%
+  group_by(Type=ifelse(Source == "Human", "Human", "Source")) %>%
+  summarize(Count=n())
+
+# Total number of genotypes
[…]

### 2018-03-21 paper/fig_attribution.R
update figures based on actual code runs

@@ -55,7 +55,7 @@ plot_df <- df %>%
   mutate(Source = factor(Source, levels=rev(c("Other", "Water", "Ruminants", "Poultry"))))
 
 # plot
-pdf("fit_attribution.pdf", width=7, height=5)
+pdf("fig_attribution.pdf", width=7, height=5)
 ggplot(plot_df) +
   geom_ribbon(aes(x=UR2006_num, ymin=li, ymax=ui, fill=Source), alpha=0.3) +
   geom_line(aes(x=UR2006_num, y=m, col=Source), lwd=1) +

### 2018-03-21 R/mcmc.R
adds loglik as another parameter to the log-likelihood

@@ -415,8 +415,9 @@ mcmc_no_ar1 = function(humans, X, phi, iterations = 10000, burnin = 1000, thinni
     }
     if (i > burnin && i %% thinning == 0) {
       post_i = post_i + 1;
-      posterior[[post_i]] = list(p     = curr$p,
-                                 theta = curr$theta)
+      posterior[[post_i]] = list(p      = curr$p,
+                                 theta  = curr$theta,
+                                 loglik = sum(curr$log_likelihood))
     }
   }
 
@@ -492,10 +493,11 @@ mcmc = function(humans, t, X, formula, phi, iterations = 10000, burnin = 1000, t
     }
     if (i > burnin && i %% thinning == 0) {
       post_i = post_i + 1;
-      posterior[[post_i]] = list(p     = curr$p,
-                                 theta = curr$theta,
-                                 tau   = curr$tau,
-                                 rho   = curr$rho)
+      posterior[[post_i]] = list(p      = curr$p,
+                                 theta  = curr$theta,
+                                 tau    = curr$tau,
+                                 rho    = curr$rho,
+                                 loglik = sum(curr$log_likelihood))
     }
   }
 

### 2018-03-21 R/attribution.R
adds computation of DIC as DIC(attribution)

@@ -356,3 +356,34 @@ predict.attribution <- function(object, newdata=NULL, FUN=median, ...) {
 iterations.attribution <- function(x) {
   length(x$posterior)
 }
+
+#' Generic for retrieving the DIC from a model fit
+#' @param x an object
+#' @return the DIC from the model fit
+#' @export
+DIC <- function(x) {
+  UseMethod("DIC", x)
+}
+
+#' Retrieve the Deviance Information Critieria (DIC) of an attribution object
+#' @export
+#' @param x An object of class `attribution`
+#' @return The DIC of the model
+DIC.attribution <- function(x) {
+  # Posterior mean of theta, phi, and log likelihood
+  post_theta  = simplify2array(lapply(seq_along(x$posterior), function(i) { x$posterior[[i]]$theta }))
+  post_phi    = x$genotype_data$sampling_distribution
+  post_loglik = simplify2array(lapply(seq_along(x$posterior), function(i) { x$posterior[[i]]$loglik }))
+  mean_theta  = apply(post_theta, 1:2, mean)
+  mean_phi    = apply(post_phi, 1:2, mean)
+  mean_loglik = mean(post_loglik)
+  # log likelihood at posterior mean
+  mean_p      = x$model_matrix %*% mean_theta
+  loglik_mean = sum(log_lik_full(x$cases, mean_phi, mean_p))
+  # compute DIC
+  d_theta_bar = -2*loglik_mean
+  d_bar_theta = -2*mean_loglik
+  p_d         = d_bar_theta - d_theta_bar
+  DIC = d_theta_bar + 2*p_d
+  DIC
+}

### 2018-03-22 paper/finalish/article.tex
update tables.r to compute DICs for each model, and update the corresponding table in the text

@@
   \begin{center}
-    \begin{tabular}{crrrrr}
+	\begin{tabular}{lrr}
 	\toprule
-      & \multicolumn{2}{c}{Island model}  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Dirichlet model} \\
-      Data & Linear & Categorical & & Linear & Categorical\\ \midrule
-      whole period  & 7913.6 & 7943.3  & & 10328.4  & 10389.8 \\
-      post-intervention period  & 5413.4  & 5439.5  & & 6615.1   & 6668.9 \\
+	Model & Dirichlet & Island\\
+	\midrule
+	Linear & 10,968.3 & 12,276.4\\
+	Categorical & 10,976.4 & 12,287.2\\
 	\bottomrule
@@
   \end{center}
-  \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 (whole period) and to the data from 2008 to 2014 (post-intervention period) given the genotypic information is from the asymmetric Island model or the Dirichlet model.}
+  \caption{DIC values for the linear model and for the categorical model applied to the data from 2005 to 2014 for the the asymmetric Island and Dirichlet models.}
   \label{tab:dic}

### 2018-03-22 paper/fig_attribution.R

@@ -43,7 +43,15 @@ df_dl <- predict(at_dl, FUN=identity) %>% extract(X, into="UR2006_num", regex="U
 
 df <- bind_rows(df_if, df_il, df_df, df_dl)
 
-save(list='df', file='attribution_fits.Rdata')
+# compute DIC's
+dic_data <-
+  bind_rows(data.frame(DIC = DIC(at_il), GenotypeModel='Island', AttributionModel='Linear'),
+            data.frame(DIC = DIC(at_if), GenotypeModel='Island', AttributionModel='Categorical'),
+            data.frame(DIC = DIC(at_dl), GenotypeModel='Dirichlet', AttributionModel='Linear'),
+            data.frame(DIC = DIC(at_df), GenotypeModel='Dirichlet', AttributionModel='Categorical'))
+write.csv(dic_data, 'dic_attribution.csv', row.names=FALSE)
+
+save(list=c('df', 'at_if', 'at_il', 'at_df', 'at_dl'), file='attribution_fits.Rdata')
 load('attribution_fits.Rdata')
 
 plot_df <- df %>%

### 2018-03-22 paper/fig_intervention.R

@@ -40,7 +40,7 @@ df_dl <- predict(at_id, FUN=identity) %>% extract(X, into="UR2006_num", regex="U
 
 df <- bind_rows(df_il, df_dl)
 
-save(list='df', file='attribution_fits_intervention.Rdata')
+save(list=c('df', 'at_if', 'at_id'), file='attribution_fits_intervention.Rdata')
 load('attribution_fits_intervention.Rdata')
 
 plot_df <- df %>%

### 2018-03-22 paper/tables.R

@@ -43,7 +43,12 @@ cases %>% left_join(popn) %>%
 
 # Table 4: DIC values
 
-# TODO: Need to grab this out of the (saved) model objects from fig_attribution.R
+dic_attr <- read.csv('dic_attribution.csv')
+dic_attr %>% mutate(DIC=round(DIC, 1),
+                    Model = factor(AttributionModel, levels=c("Linear", "Categorical"))) %>%
+  select(Model, GenotypeModel, DIC) %>%
+  spread(GenotypeModel, DIC) %>%
+  knitr::kable('latex', booktabs=TRUE, linesep='', format.args=list(big.mark=','))
 
 # Additional information for the text:
 

### 2018-03-22 paper/finalish/article.tex
update link to code repository

@@
 
-All code for fitting the genotype distributions, estimating attribution probabilities and producing figures, along with the data are available in the R package `islandR` from \url{https://github.com/jmarshallnz/islandR}.
+All code for fitting the genotype distributions, estimating attribution probabilities and producing figures, along with the data are available in the R package `islandR` from \url{https://github.com/jmarshallnz/dirichlet_island}.
 

### 2018-03-22 paper/finalish/article.tex
get rid of unnecessary constractions 'don't' and 'we've'

@@
 Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution well where the bulk of the data lie.
-It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute significantly to the overall attribution.
+It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they do not contribute significantly to the overall attribution.
 In other circumstances, such as where we have many more than 7 loci, we would expect to have many more rare genotypes, so that the Dirichlet model might provide little useful information.
@@
 Another direction is in expanding the role of water.
-In these models, we've assumed that water is a source of human campylobacteriosis infection, but water differs from the other food and environmental sources in that it is not an amplifying reservoir for \emph{Campylobacter} \cite{Wagen}.
+In these models, we have assumed that water is a source of human campylobacteriosis infection, but water differs from the other food and environmental sources in that it is not an amplifying reservoir for \emph{Campylobacter} \cite{Wagen}.
 In contrast, genotypes found in water might be expected to originate in the other sources present here, particularly ruminants and wild birds, but also potentially from humans as well via discharge of unprocessed human waste.

### 2018-03-27 paper/finalish/article.tex
adds acknowledgement to NZFSSRC
update date + address

@@
 
-\received{26 April 2016}
-\revised{6 June 2016}
-\accepted{6 June 2016}
+\received{27 March 2018}
+\revised{27 March 2018}
+\accepted{27 March 2018}
 
@@
 
-\address[1]{\orgdiv{Institute of Fundamental Sciences},\orgname{Massey University}, \orgaddress{\state{Palmerston North}, \country{New Zealand}}}
+\address[1]{\orgdiv{Institute of Fundamental Sciences}, \orgname{Massey University}, \orgaddress{\state{Palmerston North}, \country{New Zealand}}}
 
@@
 We thank Mid Central Public Health Services, and the Molecular Epidemiology and Public Health Laboratory (mEpiLab, Massey University) for data collection, and the Ministry of Primary Industries for funding the Manawatu campylobacteriosis sentinel surveillance site.
+Nigel French is supported by the New Zealand Food Safety Science and Research Centre.
 

### 2018-03-27 paper/finalish/article.tex
fix citations for ST-2026

@@
 Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases.
-As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants \cite{Muell, Coll, Cart}.
-The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.
+As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants \cite{Muell, MullA, Coll}.
+The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{Cart, french2014evolution}.
 

### 2018-03-27 paper/finalish/article.tex
utilised -> utilized

@@
 \item Sample $u$ from $U(0,1)$.
-If $u < \min \left\{1, \alpha \right\}$, the proposals $f^{(*)}$, $F^{(*)}$ and $L\Big(F^{(*)}; \text{ST}\Big)$ are utilised for the next iteration; otherwise, the previous values are retained.
+If $u < \min \left\{1, \alpha \right\}$, the proposals $f^{(*)}$, $F^{(*)}$ and $L\Big(F^{(*)}; \text{ST}\Big)$ are utilized for the next iteration; otherwise, the previous values are retained.
 \end{enumerate}
@@
 In order to examine the effects of this change on attribution probabilities, we repeated the analysis by including an interaction with time period (2005--2007 and 2008--2014).
-Figure~\ref{fig:intervention} shows that the general trend in attribution by rurality for each of the time periods utilising a linear trend on the logit scale to incorporate rurality.
+Figure~\ref{fig:intervention} shows that the general trend in attribution by rurality for each of the time periods utilizing a linear trend on the logit scale to incorporate rurality.
 There is a clear difference, with a significantly lower attribution to poultry (and correspondingly higher attribution to ruminants) in all but the most rural of areas, being strongest in highly urban areas.
@@
 Models that determine the source of human infection, particularly for zoonotic pathogens that originate in animal populations, are of considerable value to public health policy makers.
-However, such models may be complex, particularly when utilising evolutionary models.
+However, such models may be complex, particularly when utilizing evolutionary models.
 An outstanding question is whether such complexity is required, or whether a simpler model may work as effectively.

### 2018-03-27 paper/finalish/article.tex
drop islandR package reference

@@
 
-All code for fitting the genotype distributions, estimating attribution probabilities and producing figures, along with the data are available in the R package `islandR` from \url{https://github.com/jmarshallnz/dirichlet_island}.
+All code for fitting the genotype distributions, estimating attribution probabilities and producing figures, along with the data are available from \url{https://github.com/jmarshallnz/dirichlet_island}.
 
